### 2017-12-21 papers/cua.tex
add first slide for CUA presentation

\documentclass{book}

\usepackage{nopageno}

\usepackage{vmargin}
\usepackage{times}
\usepackage{graphics}
\usepackage{amsmath, amscd, amsxtra, amsthm}
\usepackage{amssymb}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage{framed}
\usepackage{mdframed}
\usepackage{float}
\usepackage{color}

\usepackage{xstring}

\usepackage{ragged2e}

\usepackage{enumitem}

%% Suggested by http://tex.stackexchange.com/questions/349580
\usepackage{array}

% This package makes cut-and-paste of carets work right, so
% we can cut-and-paste from the PDF document into Maxima.

\usepackage[T1]{fontenc}

\usepackage{tikz}
\usetikzlibrary{positioning, fit}

\begin{document}

\tikzstyle{task} = [draw, fill=blue!20, text width=8em, text depth=8em, text centered,
  minimum width=10em, minimum height=10em, rounded corners]

\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=2em, text centered,
  minimum width=5em, minimum height=2em, rounded corners]

\newcommand\task[3]{
\node (#1) [task] at (#2) {#1};

\IfSubStr{#3}{l}{
\foreach \i in {-8,...,8}%
{
  \draw[xshift=-5em, ultra thick] (#2) +(0,0.2 * \i) -- +(0.2,0.2 * \i) coordinate (#1 Port \i);

}
}{}

\IfSubStr{#3}{r}{
\foreach \i in {-8,...,8}%
{
  \draw[xshift=5em, ultra thick] (#2) +(-0.2,0.2 * \i) -- +(0,0.2 * \i) coordinate (#1 Port \i);

}
}{}
}

\begin{center}
\begin{tikzpicture}

\task{Task 1}{0,0}{r}
\task{Task 2}{15 em,0}{l}
\task{Task 3}{0,-15 em}{r}

\node (message) [message] at (0,-1 em) {Message 1};
\node (message2) [message] at (15 em, -1 em) {Message 1};
\node (message2a) [message] at (16 em, -2.5 em) {Message 2};
\node (message3) [message] at (0, -15 em) {Message 2};

\draw [color=violet, thick, ->] (node cs:name=message,angle=20) -- +(.3,0) |- (Task 1 Port 3) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 3)
    -- +(.3,0) node [near end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);

\draw [color=violet, thick, ->] (node cs:name=message3,angle=0) -- +(.3,0) |- (Task 3 Port 0) node [midway, anchor=south, color=black] {18} -- +(1,0) |- (Task 2 Port 3)
    -- +(.3,0) |- (node cs:name=message2a, angle=160);

\end{tikzpicture}
\end{center}

% https://events.ccc.de/congress/2007/Fahrplan/attachments/986_inside_the_mac_osx_kernel.pdf

% Mac OS X Internals: A Systems Approach

\end{document}

### 2017-12-22 papers/cua.tex
cua: use matrix mode to draw tasks

@@
 
-\newcommand\task[3]{
-\node (#1) [task] at (#2) {#1};
+\newcommand\task[4]{
+\node (#1) [matrix, task] at (#2) {\node {#1}; #4
 
@@
 {
-  \draw[xshift=-5em, ultra thick] (#2) +(0,0.2 * \i) -- +(0.2,0.2 * \i) coordinate (#1 Port \i);
+  \draw[xshift=-5em, ultra thick] (0,0.2 * \i) -- (0.2,0.2 * \i) coordinate (#1 Port \i);
 
@@
 {
-  \draw[xshift=5em, ultra thick] (#2) +(-0.2,0.2 * \i) -- +(0,0.2 * \i) coordinate (#1 Port \i);
+  \draw[xshift=5em, ultra thick] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate (#1 Port \i);
 
@@
 }{}
+
+\\};
 }
@@
 
-\task{Task 1}{0,0}{r}
-\task{Task 2}{15 em,0}{l}
-\task{Task 3}{0,-15 em}{r}
+\task{Task 1}{0,0}{r}{
+\node (message1) [message] {Message 1};
+}
+\task{Task 2}{15 em,0}{l}{
+\node (message2) [message] {Message 1};
+\node (message2a) [message] at (1 em, -2 em) {Message 2};
+}
+\task{Task 3}{0,-15 em}{r}{
+\node (message3) [message] {Message 2};
+}
 
-\node (message) [message] at (0,-1 em) {Message 1};
-\node (message2) [message] at (15 em, -1 em) {Message 1};
-\node (message2a) [message] at (16 em, -2.5 em) {Message 2};
-\node (message3) [message] at (0, -15 em) {Message 2};
-
-\draw [color=violet, thick, ->] (node cs:name=message,angle=20) -- +(.3,0) |- (Task 1 Port 3) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 3)
+\draw [color=violet, thick, ->] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 3) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 3)
     -- +(.3,0) node [near end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);

### 2017-12-24 papers/cua.tex
cua: add send rights in messages and line up tick marks correctly again

@@
 \usepackage{tikz}
-\usetikzlibrary{positioning, fit}
+\usetikzlibrary{positioning, fit, backgrounds}
 
@@
 
-\tikzstyle{task} = [draw, fill=blue!20, text width=8em, text depth=8em, text centered,
-  minimum width=10em, minimum height=10em, rounded corners]
+\tikzstyle{task} = [draw, fill=blue!20, rounded corners]
 
-\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=2em, text centered,
+\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=3em, text centered,
   minimum width=5em, minimum height=2em, rounded corners]
@@
 \newcommand\task[4]{
-\node (#1) [matrix, task] at (#2) {\node {#1}; #4
+\matrix (#1) [task, inner xsep=0] at (#2) { \node {#1}; \\ #4
 
+% print port tick marks in fill color if they are disabled, in order to make
+% them expand out the dimensions of the task box
+
+\foreach \i in {-8,...,8}
+{
 \IfSubStr{#3}{l}{
-\foreach \i in {-8,...,8}%
-{
   \draw[xshift=-5em, ultra thick] (0,0.2 * \i) -- (0.2,0.2 * \i) coordinate (#1 Port \i);
-
+}{
+  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i) coordinate;
 }
-}{}
-
 \IfSubStr{#3}{r}{
-\foreach \i in {-8,...,8}%
-{
   \draw[xshift=5em, ultra thick] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate (#1 Port \i);
-
+}{
+  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate;
 }
-}{}
+}
 
@@
 \begin{center}
+
 \begin{tikzpicture}
@@
 \task{Task 1}{0,0}{r}{
-\node (message1) [message] {Message 1};
+% print two sendright nodes, one to fill the message1 box to the correct size,
+% and another one to actually draw it, since message1 will overwrite the
+% first sendright
+\node (sendright) [draw, fill=orange] {Send 41};
+\node (message1) [fit=(sendright), message] {Message 1};
+\node (sendright) [draw, fill=orange] {Send 41};
 }

### 2018-01-14 papers/cua.tex
working on CUA lecture slides

@@
-\documentclass{book}
+\documentclass{beamer}
+\usetheme{Madrid}
 
-\usepackage{nopageno}
+\title{Building a GNU/Hurd Cluster}
+\author{Brent Baccala}
+\institute{\tt cosine@freesoft.org}
+\date{January 17, 2018}
 
-\usepackage{vmargin}
-\usepackage{times}
-\usepackage{graphics}
-\usepackage{amsmath, amscd, amsxtra, amsthm}
-\usepackage{amssymb}
-\usepackage[retainorgcmds]{IEEEtrantools}
-\usepackage{framed}
-\usepackage{mdframed}
-\usepackage{float}
+\setbeamertemplate{footline}{}
+\beamertemplatenavigationsymbolsempty
+
 \usepackage{color}
+\usepackage{comment}
+\usepackage{graphicx}
+
+\usepackage{tabularx}
 
@@
 
-\usepackage{ragged2e}
-
-\usepackage{enumitem}
-
-%% Suggested by http://tex.stackexchange.com/questions/349580
-\usepackage{array}
-
-% This package makes cut-and-paste of carets work right, so
-% we can cut-and-paste from the PDF document into Maxima.
-
-\usepackage[T1]{fontenc}
-
 \usepackage{tikz}
-\usetikzlibrary{positioning, fit, backgrounds}
+\usetikzlibrary{positioning, fit, backgrounds, arrows, patterns}
 
 \begin{document}
+
+\begin{frame}
+\titlepage
+\begin{block}{Abstract}
+The GNU Hurd is a UNIX-like operating system based on Mach, an
+operating system kernel developed at Carnegie Mellon University and
+designed to support multi-node clusters, which have become ubiquitous
+in high performance computing.  This talk will highlight current
+trends in cluster computing, outline the design of Mach and Hurd, and
+summarize the speaker's work to realize Hurd's potential as a
+POSIX-compatible operating system that can leverage the power of an
+entire cluster.
+\end{block}
+\end{frame}
+
+\begin{frame}
+\frametitle{Cisco 5108 Blade Server Chassis with eight B200 M2s}
+\includegraphics[width=\textwidth]{ourcisco.jpg}
+\end{frame}
+
+\begin{frame}
+\frametitle{Processor families in TOP500}
+\includegraphics[width=\textwidth]{Processor_families_in_TOP500_supercomputers.png}
+\end{frame}
+
+\newcommand\topfh[7]{
+#1 & #2 & #3 & #4 & #6 & #7 \\
+& \multicolumn{4}{l}{\hskip 1cm #5} \\
+\\
+}
+
+\begin{frame}
+\frametitle{Top 10 positions of the TOP500 (November 2017)}
+{\tiny
+\begin{tabularx}{\textwidth}{lllllll}
+\bf Rank
+& \bf Name
+& \bf Processor
+& \bf Vendor
+& \bf Cores
+& \bf Operating system
+\\
+\\
+\topfh{1}
+{Sunway TaihuLight}
+{SW26010}
+{NRCPC}
+{National Supercomputing Center in Wuxi, CHN}
+{10,649,600}
+{Linux (Raise)}
+
+\topfh{2}
+{Tianhe-2}
+{Xeon E5–2692, Xeon Phi 31S1P}
+{NUDT}
+{National Supercomputing Center in Guangzhou, CHN}
+{3,120,000}
+{Linux (Kylin)}
+
+\topfh{3}
+{Piz Daint}
+{Xeon E5-2690v3, Nvidia Tesla P100}
+{Cray}
+{Swiss National Supercomputing Centre, SWI}
+{361,760}
+{Linux (CLE)}
+
+\topfh{4}
+{Gyoukou}
+{Xeon D-1571, PEZY-SC2}
+{ExaScaler}
+{Japan Agency for Marine-Earth Science and Technology, JAP}
+{19,860,000}
+{Linux (CentOS)}
+
+\topfh{5}
+{Titan}
+{Opteron 6274, Nvidia Tesla K20X}
+{Cray}
+{Oak Ridge National Laboratory, USA}
+{560,640}
+{Linux (CLE, SLES based)}
+
+\topfh{6}
+{Sequoia}
+{IBM A2}
+{IBM}
+{Lawrence Livermore National Laboratory, USA}
+{1,572,864}
+{Linux (RHEL and CNK)}
+
+\topfh{7}
+{Trinity}
+{Xeon E5–2698v3, Xeon Phi}
+{Cray}
+{Los Alamos National Laboratory, USA}
+{979,968}
+{Linux (CLE)}
+
+\topfh{8}
+{Cori}
+{Xeon Phi 7250}
+{Cray}
+{National Energy Research Scientific Computing Center, USA}
+{622,336}
+{Linux (CLE)}
+
+\topfh{9}
+{Oakforest-PACS}
+{Xeon Phi 7250}
+{Fujitsu}
+{Kashiwa, Joint Center for Advanced High Performance Computing, JPN}
+{556,104}
+{Linux}
+
+\topfh{10}
+{K computer}
+{SPARC64 VIIIfx}
+{Fujitsu}
+{Riken, Advanced Institute for Computational Science (AICS), JPN}
+{705,024}
+{Linux}
+
+\end{tabularx}
+Source: {\tt https://www.top500.org/}
+}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Single System Image}
+\begin{definition}
+A single system image (SSI) cluster is a cluster of machines that appears to be one single system.
+
+\hfill Wikipedia
+\end{definition}
+\begin{center}
+\begin{tikzpicture}
+\node (a) [draw] at (0:1) {};
+\node (b) [draw] at (60:1) {} edge (a);
+\node (c) [draw] at (120:1) {} edge (b);
+\node (d) [draw] at (180:1) {} edge (c);
+\node (e) [draw] at (240:1) {} edge (d);
+\node (f) [draw] at (300:1) {} edge (e) edge (a);
+\end{tikzpicture}
+\end{center}
+\begin{semiverbatim}
+\$ nproc
+96
+\$ free -h
+              total        used        free
+Mem:           768G         45G        723G
+Swap:            0B          0B          0B
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}
+\frametitle{Single System Image}
+
+\begin{itemize}
+\item Key Features
+\begin{itemize}
+\item Cluster-wide login
+\item Single Process Space
+\item Process Migration
+\item Single Root Filesystem
+\item Redundant Filesystem
+\item Single I/O Space - block devices
+\item Single IPC Space
+standard Linux inter-process communication mechanisms, shared memory, semaphores, SYSV message queues, pipes and Unix domain sockets.
+\item Cluster-wide IP Addresses
+\end{itemize}
+\item Advantages
+\begin{itemize}
+\item familiar programming and management tools
+\item ease of leveraging additional computing power
+\end{itemize}
+\item Disadvantages
+\begin{itemize}
+\item less fault isolation
+\item node failure more difficult
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Cluster Computing - History}
+\begin{itemize}
+\item Beowulf
+\item Hadoop
+\item Cluster Linux
+Kerrighed - Latest version is 3.0.0 which was released on 14 June 2010 and based on Linux kernel 2.6.30.
+OpenSSI - last development release 2010; last stable release 2007
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Single System Image - History}
+\begin{itemize}
+\item VMS
+\item SCALEMP
+\item Cluster Linux
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{GNU Hurd}
+\begin{itemize}
+\item History
+\item Microkernels in general
+\item CMU Mach
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{CMU Mach}
+\begin{itemize}
+\item Tasks / Threads
+\item Interprocess Communication
+\item Memory Management
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Techniques to exploit parallelism}
+\begin{itemize}
+\item Instruction level parallelism
+\begin{itemize}
+\item Superscalar vector operations
+\item Speculative execution
+\end{itemize}
+\item Thread level parallelism
+\item Node level parallelism
+\begin{itemize}
+\item Clusters
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Adding 512 64-bit integers - conventional x86-64 instrs}
+\begin{semiverbatim}
+\tiny
+int a[512];
+int b[512];
+int c[512];
+
+int main () \{
+  for (int i=0; i<512; i++) \{
+    a[i] = b[i] + c[i];
+  \}
+\}
+
+gcc -O2 -o ex2 ex2.c
+
+00000000004003e0 <main>:
+4003e0:31 c0                xor    %eax,%eax
+4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
+4003e8:8b 90 60 18 60 00    mov    0x601860(%rax),%edx
+4003ee:03 90 60 10 60 00    add    0x601060(%rax),%edx
+4003f4:48 83 c0 04          add    $0x4,%rax
+4003f8:89 90 5c 20 60 00    mov    %edx,0x60205c(%rax)
+4003fe:48 3d 00 08 00 00    cmp    $0x800,%rax
+400404:75 e2                jne    4003e8 <main+0x8>
+400406:31 c0                xor    %eax,%eax
+400408:c3                   retq
+
+\end{semiverbatim}
+\vskip 0pt plus 100fill
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Adding 512 64-bit integers - SSE superscalar instructions}
+\begin{semiverbatim}
+\tiny
+int a[512] __attribute__((aligned(0x10)));
+int b[512] __attribute__((aligned(0x10)));
+int c[512] __attribute__((aligned(0x10)));
+
+int main () \{
+  for (int i=0; i<512; i++) \{
+    a[i] = b[i] + c[i];
+  \}
+\}
+
+gcc -O3 -o ex2 ex2.c
+
+00000000004003e0 <main>:
+4003e0:31 c0                xor    %eax,%eax
+4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
+4003e8:66 0f 6f 80 40 10 60 movdqa 0x601040(%rax),%xmm0
+4003ef:00 
+4003f0:48 83 c0 10          add    $0x10,%rax
+4003f4:66 0f fe 80 30 18 60 paddd  0x601830(%rax),%xmm0
+4003fb:00 
+4003fc:0f 29 80 30 20 60 00 movaps %xmm0,0x602030(%rax)
+400403:48 3d 00 08 00 00    cmp    $0x800,%rax
+400409:75 dd                jne    4003e8 <main+0x8>
+40040b:31 c0                xor    %eax,%eax
+40040d:c3                   retq
+
+\end{semiverbatim}
+\vskip 0pt plus 100fill
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Adding 512 64-bit integers - AVX2 superscalar instructions}
+\begin{semiverbatim}
+\tiny
+int a[512] __attribute__((aligned(0x20)));
+int b[512] __attribute__((aligned(0x20)));
+int c[512] __attribute__((aligned(0x20)));
+
+int main () \{
+  for (int i=0; i<512; i++) \{
+    a[i] = b[i] + c[i];
+  \}
+\}
+
+gcc -march=native -O3 -o ex2 ex2.c
+
+00000000004003e0 <main>:
+4003e0:31 c0                xor    %eax,%eax
+4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
+4003e8:c5 fd 6f 80 60 10 60 vmovdqa 0x601060(%rax),%ymm0
+4003ef:00 
+4003f0:c5 fd fe 80 60 18 60 vpaddd 0x601860(%rax),%ymm0,%ymm0
+4003f7:00 
+4003f8:48 83 c0 20          add    $0x20,%rax
+4003fc:c5 fd 7f 80 40 20 60 vmovdqa %ymm0,0x602040(%rax)
+400403:00 
+400404:48 3d 00 08 00 00    cmp    $0x800,%rax
+40040a:75 dc                jne    4003e8 <main+0x8>
+40040c:c5 f8 77             vzeroupper 
+40040f:31 c0                xor    %eax,%eax
+400411:c3                   retq
+
+\end{semiverbatim}
+\vskip 0pt plus 100fill
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Adding 512 64-bit integers - AVX512 superscalar instrs}
+\begin{semiverbatim}
+\tiny
+int a[512] __attribute__((aligned(0x40)));
+int b[512] __attribute__((aligned(0x40)));
+int c[512] __attribute__((aligned(0x40)));
+
+int main () \{
+  for (int i=0; i<512; i++) \{
+    a[i] = b[i] + c[i];
+  \}
+\}
+
+gcc -mavx512vl -O3 -o ex2 ex2.c
+
+00000000004003e0 <main>:
+4003e0:31 c0                xor    %eax,%eax
+4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
+4003e8:62 f1 fd 48 6f 80 80 vmovdqa64 0x601080(%rax),%zmm0
+4003ef:10 60 00 
+4003f2:48 83 c0 40          add    $0x40,%rax
+4003f6:62 f1 7d 48 fe 80 40 vpaddd 0x601440(%rax),%zmm0,%zmm0
+4003fd:14 60 00 
+400400:62 f1 fd 48 7f 80 40 vmovdqa64 %zmm0,0x601840(%rax)
+400407:18 60 00 
+40040a:48 3d 00 08 00 00    cmp    $0x800,%rax
+400410:75 d6                jne    4003e8 <main+0x8>
+400412:31 c0                xor    %eax,%eax
+400414:c3                   retq
+
+baccala@ideapad-S510p:~$ ./ex2
+Illegal instruction (core dumped)
+baccala@ideapad-S510p:~$ ./ex2
+
+\end{semiverbatim}
+\vskip 0pt plus 100fill
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Pointer Aliasing}
+\begin{semiverbatim}
+\tiny
+void operation (int * a, int * b, int * c)
+\{
+  for (int i=0; i<16; i++) \{
+    c[i] = a[i] + b[i];
+  \}
+\}
+\end{semiverbatim}
+\pause
+\begin{semiverbatim}
+\tiny
+int v[18] = \{1, 1\};
+
+int main () \{
+  operation(\&v[0], \&v[1], \&v[2]);
+
+  for (int i=0; i<18; i++) \{
+    printf("%d ", v[i]);
+  \}
+  printf("\\n");
+\}
+\end{semiverbatim}
+What does this code do?
+\pause
+\begin{semiverbatim}
+\tiny
+baccala@blade6:~\$ ./ex4
+1 1 2 3 5 8 13 21 34 55 89 144 233 377 610 987 1597 2584 
+baccala@blade6:~\$
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Pointer Aliasing - The {\tt restrict} keyword}
+\begin{semiverbatim}
+\tiny
+void operation (int * restrict a, int * restrict b, int * restrict c)
+\{
+  for (int i=0; i<16; i++) \{
+    a[i] = b[i] + c[i];
+  \}
+\}
+
+gcc -mavx512vl -O3 -c ex3.c
+
+0000000000000000 <operation>:
+   0:   62 f1 7e 48 6f 06       vmovdqu32 (%rsi),%zmm0
+   6:   62 f1 7d 48 fe 02       vpaddd (%rdx),%zmm0,%zmm0
+   c:   62 f1 7e 48 7f 07       vmovdqu32 %zmm0,(%rdi)
+  12:   c3                      retq
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}
+\frametitle{Intel Ivy Bridge Processor}
+\includegraphics[width=\textwidth]{ivybridge.jpg}
+\begin{center}
+\huge
+Notice the size of the graphics co-processor!
+\end{center}
+\end{frame}
+
+\begin{comment}
+\begin{frame}[fragile]
+\begin{semiverbatim}
+\tiny
+import java.io.IOException;
+import java.util.StringTokenizer;
+
+import org.apache.hadoop.conf.Configuration;
+import org.apache.hadoop.fs.Path;
+import org.apache.hadoop.io.IntWritable;
+import org.apache.hadoop.io.Text;
+import org.apache.hadoop.mapreduce.Job;
+import org.apache.hadoop.mapreduce.Mapper;
+import org.apache.hadoop.mapreduce.Reducer;
+import org.apache.hadoop.mapreduce.lib.input.FileInputFormat;
+import org.apache.hadoop.mapreduce.lib.output.FileOutputFormat;
+
+public class WordCount \{
+
+  public static class TokenizerMapper
+       extends Mapper<Object, Text, Text, IntWritable>\{
+
+    private final static IntWritable one = new IntWritable(1);
+    private Text word = new Text();
+
+    public void map(Object key, Text value, Context context
+                    ) throws IOException, InterruptedException \{
+      StringTokenizer itr = new StringTokenizer(value.toString());
+      while (itr.hasMoreTokens()) \{
+        word.set(itr.nextToken());
+        context.write(word, one);
+      \}
+    \}
+  \}
+
+  public static class IntSumReducer
+       extends Reducer<Text,IntWritable,Text,IntWritable> \{
+    private IntWritable result = new IntWritable();
+
+    public void reduce(Text key, Iterable<IntWritable> values,
+                       Context context
+                       ) throws IOException, InterruptedException \{
+      int sum = 0;
+      for (IntWritable val : values) \{
+        sum += val.get();
+      \}
+      result.set(sum);
+      context.write(key, result);
+    \}
+  \}
+
+  public static void main(String[] args) throws Exception \{
+    Configuration conf = new Configuration();
+    Job job = Job.getInstance(conf, ``word count'');
+    job.setJarByClass(WordCount.class);
+    job.setMapperClass(TokenizerMapper.class);
+    job.setCombinerClass(IntSumReducer.class);
+    job.setReducerClass(IntSumReducer.class);
+    job.setOutputKeyClass(Text.class);
+    job.setOutputValueClass(IntWritable.class);
+    FileInputFormat.addInputPath(job, new Path(args[0]));
+    FileOutputFormat.setOutputPath(job, new Path(args[1]));
+    System.exit(job.waitForCompletion(true) ? 0 : 1);
+  \}
+\}
+\end{semiverbatim}
+\end{frame}
+\end{comment}
+
+\begin{frame}[fragile]
+\frametitle{Hadoop MapReduce Tutorial - WordCount}
+\begin{semiverbatim}
+\tiny public class WordCount \{
+
+  public static class TokenizerMapper
+       extends Mapper<Object, Text, Text, IntWritable>\{
+
+    private final static IntWritable one = new IntWritable(1);
+    private Text word = new Text();
+
+    public void map(Object key, Text value, Context context)
+                     throws IOException, InterruptedException \{
+      StringTokenizer itr = new StringTokenizer(value.toString());
+      while (itr.hasMoreTokens()) \{
+        word.set(itr.nextToken());
+        context.write(word, one);
+      \}
+    \}
+  \}
+
+  public static class IntSumReducer
+       extends Reducer<Text,IntWritable,Text,IntWritable> \{
+    private IntWritable result = new IntWritable();
+
+    public void reduce(Text key, Iterable<IntWritable> values, Context context)
+                        throws IOException, InterruptedException \{
+      int sum = 0;
+      for (IntWritable val : values) \{
+        sum += val.get();
+      \}
+      result.set(sum);
+      context.write(key, result);
+    \}
+  \}
+\}
+\end{semiverbatim}
+\end{frame}
+
+% tikz commands to draw task boxes to illustrate Mach IPC concepts
 
@@
 
+% \task - takes four arguments
+% 1. name of task
+% 2. location of task box's center point
+% 3. "r" or "l", indicating if a set of tick marks representing
+%    ports should drawn on the left or right side of the task box
+%    "u" to make the task unlabled
+% 4. contents of the task box - other nodes
+%
+% The tick marks will be named "#1 Port #" for future reference
+
 \newcommand\task[4]{
-\matrix (#1) [task, inner xsep=0] at (#2) { \node {#1}; \\ #4
+\matrix (#1) [task, inner xsep=0] at (#2) { \node {\IfSubStr{#3}{u}{}{#1}}; \\ #4
 
 % print port tick marks in fill color if they are disabled, in order to make
-% them expand out the dimensions of the task box
+% them expand out the dimensions of the task box yet remain invisible
 
-\foreach \i in {-8,...,8}
+\foreach \i in {-7,...,7}
 {
 \IfSubStr{#3}{l}{
-  \draw[xshift=-5em, ultra thick] (0,0.2 * \i) -- (0.2,0.2 * \i) coordinate (#1 Port \i);
+  \draw[xshift=-5em, ultra thick] (0,0.2 * \i)  coordinate (#1 Port \i) -- (0.2,0.2 * \i);
 }{
-  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i) coordinate;
+  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i);
 }
@@
 }{
-  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate;
+  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i);
 }
@@
 }
+
+\begin{frame}
+\frametitle{Ports - Mach's Interprocess Communication (IPC)}
+
+\begin{center}
+
+\begin{tikzpicture}[transform canvas={scale=0.4, yshift=10em, xshift=-15em}]
+
+\task{Task 1 left}{0,0}{ru}{}
+\task{Task 2 left}{0,-11 em}{ru}{}
+\task{Task 3 left}{0,-22 em}{ru}{}
+\task{Task 4 left}{0,-33 em}{ru}{}
+\task{Task 1 right}{30 em,0}{lu}{}
+\task{Task 2 right}{30 em,-11 em}{lu}{}
+\task{Task 3 right}{30 em,-22 em}{lu}{}
+\task{Task 4 right}{30 em,-33 em}{lu}{}
+
+\draw [color=violet, thick, -triangle 90] (Task 1 left Port 5) -- (Task 1 right Port 5);
+\draw [color=violet, thick] (Task 2 left Port 0) -- +(1,0) |- (Task 1 right Port 5);
+\draw [color=violet, thick, -triangle 90] (Task 2 left Port 5) -- (Task 2 right Port 5);
+\draw [color=violet, thick, triangle 90-] (Task 2 left Port -5) -- (Task 2 right Port -5);
+
+\filldraw[fill=white] (Task 1 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
+\filldraw[fill=white, pattern=vertical lines] (Task 1 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
+
+\filldraw[fill=white] (Task 2 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
+\filldraw[fill=white, pattern=vertical lines] (Task 2 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
+
+\filldraw[fill=white] (Task 2 right Port -5) ++(-4,-0.5) rectangle +(1.8,1);
+\filldraw[fill=white, pattern=vertical lines] (Task 2 right Port -5) ++(-4,-0.5) rectangle +(1.8,1);
+
+\end{tikzpicture}
+\end{center}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Mach IPC}
+
+\begin{center}
+
+\begin{tikzpicture}
+
+\task{Task 1}{0,0}{r}{
+\node (message1) [message] {Message 1};
+}
+\task{Task 2}{15 em,0}{l}{
+\node (message2) [message] at (1 em, 1 em) {Message 1};
+\node (message2a) [message] at (1.9 em, -1 em) {Message 2};
+}
+\task{Task 3}{0,-11 em}{r}{
+\node (message3) [message] {Message 2};
+}
+
+\draw [color=violet, thick, ->] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 5)
+    -- +(.5,0) node [at end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);
+
+\draw [color=violet, thick, ->] (node cs:name=message3,angle=0) -- +(.3,0) |- (Task 3 Port 0) node [midway, anchor=south, color=black] {18} -- +(1,0) |- (Task 2 Port 5)
+    -- +(.5,0) |- (node cs:name=message2a, angle=160);
+
+\end{tikzpicture}
+\end{center}
+
+\end{frame}
+
+\begin{frame}
+
+\frametitle{Send Rights}
+
+\begin{center}
+
+\begin{tikzpicture}
+
+\task{Task 1}{0,0}{r}{
+% print two sendright nodes, one to fill the message1 box to the correct size,
+% and another one to actually draw it, since message1 will overwrite the
+% first sendright
+\node (sendright) [draw, fill=orange] {Send 41};
+\node (message1) [fit=(sendright), message] {Message 1};
+\node (sendright) [draw, fill=orange] {Send 41};
+
+\node (message3) at (0, -7em) [message] {Message 2};
+}
+
+\task{Task 2}{15 em,0}{l}{
+\node (sendright2) [draw, fill=orange] {Send 9};
+\node (message2) [fit=(sendright2), message] {Message 1};
+\node (sendright2) [draw, fill=orange] {Send 9};
+
+\node (message4) at (0, -7em) [message] {Message 2};
+}
+
+\draw [color=violet, thick, ->] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 5)
+    -- +(.5,0) node [at end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);
+
+\draw [color=violet, thick, <-] (node cs:name=message3,angle=20) -- +(.3,0) |- (Task 1 Port -5) node [midway, anchor=south, color=black] {41} -- (Task 2 Port -5)
+    -- +(.5,0) node [near end, anchor=south, color=black] {9} |- (node cs:name=message4, angle=160);
+
+\end{tikzpicture}
+\end{center}
+
+\end{frame}
+
+\begin{frame}
+
+\frametitle{Send Rights}
 
@@
 \task{Task 2}{15 em,0}{l}{
-\node (message2) [message] {Message 1};
-\node (message2a) [message] at (1 em, -2 em) {Message 2};
+\node (sendright2) [draw, fill=orange] {Send 9};
+\node (message2) [fit=(sendright2), message] {Message 1};
+\node (sendright2) [draw, fill=orange] {Send 9};
+%\node (message2) [message] {Message 1};
+% \node (message2a) [message] at (1 em, -2 em) {Message 2};
 }
-\task{Task 3}{0,-15 em}{r}{
+\task{Task 1 }{0,-11 em}{r}{
 \node (message3) [message] {Message 2};
 }
+\task{Task 2 }{15 em,-11 em}{l}{
+\node (message4) [message] {Message 2};
+}
 
-\draw [color=violet, thick, ->] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 3) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 3)
-    -- +(.3,0) node [near end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);
+\draw [color=violet, thick, ->] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 5)
+    -- +(.5,0) node [at end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);
 
-\draw [color=violet, thick, ->] (node cs:name=message3,angle=0) -- +(.3,0) |- (Task 3 Port 0) node [midway, anchor=south, color=black] {18} -- +(1,0) |- (Task 2 Port 3)
-    -- +(.3,0) |- (node cs:name=message2a, angle=160);
+%\draw [color=violet, thick, ->] (node cs:name=message3,angle=0) -- +(.3,0) |- (Task 3 Port 0) node [midway, anchor=south, color=black] {18} -- +(1,0) |- (Task 2 Port 3)
+%    -- +(.3,0) |- (node cs:name=message2a, angle=160);
 
@@
 \end{center}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Right Transfers}
+\end{frame}
+
+\begin{frame}
+\frametitle{Memory Transfers}
+\end{frame}
+
+\begin{frame}
+\frametitle{Hurd Filesystem}
+\end{frame}
+
+\begin{frame}
+\frametitle{Example: Basic File I/O}
+\end{frame}
+
+\begin{frame}
+\frametitle{Mach Memory Management}
+\end{frame}
+
+\begin{frame}
+\frametitle{Example: Shared Libraries}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Pseudocode: {\tt memory\_object\_data\_request}}
+\begin{semiverbatim}
+\tiny
+memory_object_data_request: (kernel requesting a single page)
+
+  if TERMINATING flag set, return
+  if this client is on NEXTERROR list and WRITE ACCESS WAS REQUESTED, send a m_o_data_error message,
+        move NEXTERROR to ERROR, remove the NEXTERROR list entry, and return
+  if the page is PAGINGOUT:
+    if WAITLIST is empty and ACCESSLIST is not, send lock request (internal flush) to ACCESSLIST client
+       ASSERT: there should only be one WRITE client on ACCESSLIST
+               (the client that returned the data we're paging out)
+    add requesting client to WAITLIST and return
+  if this client is already on ACCESSLIST, it flushed and is trying to re-aquire access,
+     so check to see if WAITLIST is empty
+     if so, remove client from ACCESSLIST and proceed (kernel flushed without being asked)
+     if not, we're flushing, so add client to WAITLIST, run internal_lock_completed on this page, and return
+  if WAITLIST is not empty, add requesting client to WAITLIST and return
+  if another client has WRITE access (WAITLIST is empty), add requesting client to WAITLIST,
+     send lock request (internal flush variant) and return
+  if WRITE access is requested and ACCESSLIST is not empty (other clients have access),
+     add requesting client to WAITLIST (it's empty),
+     send lock requests (internal flush variant) and return
+     (we could give out the page with READ access and wait for a unlock request, but I think not)
+  else
+     (ACCESSLIST is empty or READ access is requested and only READ clients are on ACCESSLIST)
+     if page is flagged INVALID, send m_o_data_error to this client and return
+     add this client to WAITLIST (it's empty)
+     unlock pager
+     read the page with pager_read_page()
+     relock pager
+     if pager_read_page() returned an error, mark page with ERROR and call send_error_to_WAITLIST
+     if no error, service_waitlist(deallocate=true) (read/write status supplied by filesystem), clear ERROR
+\end{semiverbatim}
+\end{frame}
 

### 2018-01-15 papers/cua.tex
cua lecture - working on memory management, and about to reorganize Tex IfSubStr

@@
 
-\usepackage{color}
+\usepackage{xcolor}
 \usepackage{comment}
@@
 \end{block}
-\end{frame}
-
-\begin{frame}
-\frametitle{Cisco 5108 Blade Server Chassis with eight B200 M2s}
-\includegraphics[width=\textwidth]{ourcisco.jpg}
-\end{frame}
-
-\begin{frame}
-\frametitle{Processor families in TOP500}
-\includegraphics[width=\textwidth]{Processor_families_in_TOP500_supercomputers.png}
 \end{frame}
@@
 
-\begin{frame}[fragile]
-\frametitle{Single System Image}
-\begin{definition}
-A single system image (SSI) cluster is a cluster of machines that appears to be one single system.
-
-\hfill Wikipedia
-\end{definition}
-\begin{center}
-\begin{tikzpicture}
-\node (a) [draw] at (0:1) {};
-\node (b) [draw] at (60:1) {} edge (a);
-\node (c) [draw] at (120:1) {} edge (b);
-\node (d) [draw] at (180:1) {} edge (c);
-\node (e) [draw] at (240:1) {} edge (d);
-\node (f) [draw] at (300:1) {} edge (e) edge (a);
-\end{tikzpicture}
-\end{center}
-\begin{semiverbatim}
-\$ nproc
-96
-\$ free -h
-              total        used        free
-Mem:           768G         45G        723G
-Swap:            0B          0B          0B
-
-\end{semiverbatim}
+\begin{frame}
+\frametitle{Cisco 5108 Blade Server Chassis with eight B200 M2s}
+\includegraphics[width=\textwidth]{ourcisco.jpg}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Single System Image}
-
-\begin{itemize}
-\item Key Features
-\begin{itemize}
-\item Cluster-wide login
-\item Single Process Space
-\item Process Migration
-\item Single Root Filesystem
-\item Redundant Filesystem
-\item Single I/O Space - block devices
-\item Single IPC Space
-standard Linux inter-process communication mechanisms, shared memory, semaphores, SYSV message queues, pipes and Unix domain sockets.
-\item Cluster-wide IP Addresses
-\end{itemize}
-\item Advantages
-\begin{itemize}
-\item familiar programming and management tools
-\item ease of leveraging additional computing power
-\end{itemize}
-\item Disadvantages
-\begin{itemize}
-\item less fault isolation
-\item node failure more difficult
-\end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Cluster Computing - History}
-\begin{itemize}
-\item Beowulf
-\item Hadoop
-\item Cluster Linux
-Kerrighed - Latest version is 3.0.0 which was released on 14 June 2010 and based on Linux kernel 2.6.30.
-OpenSSI - last development release 2010; last stable release 2007
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Single System Image - History}
-\begin{itemize}
-\item VMS
-\item SCALEMP
-\item Cluster Linux
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{GNU Hurd}
-\begin{itemize}
-\item History
-\item Microkernels in general
-\item CMU Mach
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{CMU Mach}
-\begin{itemize}
-\item Tasks / Threads
-\item Interprocess Communication
-\item Memory Management
-\end{itemize}
+\frametitle{Processor families in TOP500}
+\includegraphics[width=\textwidth]{Processor_families_in_TOP500_supercomputers.png}
 \end{frame}
@@
 \item Speculative execution
+\item Co-processors
 \end{itemize}
 \item Thread level parallelism
+\begin{itemize}
+\item Shared memory
+\item Caching
+\item Synchronization
+\end{itemize}
 \item Node level parallelism
 \begin{itemize}
+\item No shared memory
+\item High speed LAN
 \item Clusters
@@
   for (int i=0; i<512; i++) \{
-    a[i] = b[i] + c[i];
+    c[i] = a[i] + b[i];
   \}
@@
 
-gcc -O2 -o ex2 ex2.c
+baccala@ideapad-S510p:~\$ gcc -O2 -fno-align-loops -o ex1 ex1.c
 
@@
 4003e0:31 c0                xor    %eax,%eax
-4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
-4003e8:8b 90 60 18 60 00    mov    0x601860(%rax),%edx
-4003ee:03 90 60 10 60 00    add    0x601060(%rax),%edx
-4003f4:48 83 c0 04          add    $0x4,%rax
-4003f8:89 90 5c 20 60 00    mov    %edx,0x60205c(%rax)
-4003fe:48 3d 00 08 00 00    cmp    $0x800,%rax
-400404:75 e2                jne    4003e8 <main+0x8>
-400406:31 c0                xor    %eax,%eax
-400408:c3                   retq
+4003e2:8b 90 60 18 60 00    mov    0x601860(%rax),%edx
+4003e8:03 90 60 10 60 00    add    0x601060(%rax),%edx
+4003ee:48 83 c0 04          add    $0x4,%rax
+4003f2:89 90 5c 20 60 00    mov    %edx,0x60205c(%rax)
+4003f8:48 3d 00 08 00 00    cmp    $0x800,%rax
+4003fe:75 e2                jne    4003e2 <main+0x2>
+400400:31 c0                xor    %eax,%eax
+400402:c3                   retq
 
@@
 \tiny
-int a[512] __attribute__((aligned(0x10)));
-int b[512] __attribute__((aligned(0x10)));
-int c[512] __attribute__((aligned(0x10)));
+int a[512];
+int b[512];
+int c[512];
 
@@
 
-gcc -O3 -o ex2 ex2.c
+baccala@ideapad-S510p:~$ gcc -O3 -fno-align-loops -o ex1 ex1.c
 
@@
 4003e0:31 c0                         xor    %eax,%eax
-4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
-4003e8:66 0f 6f 80 40 10 60 movdqa 0x601040(%rax),%xmm0
-4003ef:00 
-4003f0:48 83 c0 10          add    $0x10,%rax
-4003f4:66 0f fe 80 30 18 60 paddd  0x601830(%rax),%xmm0
-4003fb:00 
-4003fc:0f 29 80 30 20 60 00 movaps %xmm0,0x602030(%rax)
-400403:48 3d 00 08 00 00    cmp    $0x800,%rax
-400409:75 dd                jne    4003e8 <main+0x8>
-40040b:31 c0                xor    %eax,%eax
-40040d:c3                   retq
+4003e2:66 0f 6f 80 60 10 60 00       movdqa 0x601060(%rax),%xmm0
+4003ea:48 83 c0 10                   add    $0x10,%rax
+4003ee:66 0f fe 80 50 18 60 00       paddd  0x601850(%rax),%xmm0
+4003f6:0f 29 80 50 20 60 00          movaps %xmm0,0x602050(%rax)
+4003fd:48 3d 00 08 00 00             cmp    $0x800,%rax
+400403:75 dd                         jne    4003e2 <main+0x2>
+400405:31 c0                         xor    %eax,%eax
+400407:c3                            retq
 
@@
 \tiny
-int a[512] __attribute__((aligned(0x20)));
-int b[512] __attribute__((aligned(0x20)));
-int c[512] __attribute__((aligned(0x20)));
+int a[512] __attribute__((aligned(32)));
+int b[512] __attribute__((aligned(32)));
+int c[512] __attribute__((aligned(32)));
 
@@
 
-gcc -march=native -O3 -o ex2 ex2.c
+baccala@ideapad-S510p:~$ gcc -march=native -O3 -o ex2 ex2.c
 
@@
 4003e0:31 c0                         xor    %eax,%eax
-4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
-4003e8:c5 fd 6f 80 60 10 60 vmovdqa 0x601060(%rax),%ymm0
-4003ef:00 
-4003f0:c5 fd fe 80 60 18 60 vpaddd 0x601860(%rax),%ymm0,%ymm0
-4003f7:00 
-4003f8:48 83 c0 20          add    $0x20,%rax
-4003fc:c5 fd 7f 80 40 20 60 vmovdqa %ymm0,0x602040(%rax)
-400403:00 
-400404:48 3d 00 08 00 00    cmp    $0x800,%rax
-40040a:75 dc                jne    4003e8 <main+0x8>
-40040c:c5 f8 77             vzeroupper 
-40040f:31 c0                xor    %eax,%eax
-400411:c3                   retq
+4003e2:c5 fd 6f 80 60 10 60 00       vmovdqa 0x601060(%rax),%ymm0
+4003ea:c5 fd fe 80 60 18 60 00       vpaddd 0x601860(%rax),%ymm0,%ymm0
+4003f2:48 83 c0 20                   add    $0x20,%rax
+4003f6:c5 fd 7f 80 40 20 60 00       vmovdqa %ymm0,0x602040(%rax)
+4003fe:48 3d 00 08 00 00             cmp    $0x800,%rax
+400404:75 dc                         jne    4003e2 <main+0x2>
+400406:c5 f8 77                      vzeroupper 
+400409:31 c0                         xor    %eax,%eax
+40040b:c3                            retq
 
@@
 \tiny
-int a[512] __attribute__((aligned(0x40)));
-int b[512] __attribute__((aligned(0x40)));
-int c[512] __attribute__((aligned(0x40)));
+int a[512] __attribute__((aligned(64)));
+int b[512] __attribute__((aligned(64)));
+int c[512] __attribute__((aligned(64)));
 
@@
 
-gcc -mavx512vl -O3 -o ex2 ex2.c
+baccala@ideapad-S510p:~$ gcc -mavx512vl -O3 -o ex3 ex3.c
 
@@
 4003e0:31 c0                         xor    %eax,%eax
-4003e2:66 0f 1f 44 00 00    nopw   0x0(%rax,%rax,1)
-4003e8:62 f1 fd 48 6f 80 80 vmovdqa64 0x601080(%rax),%zmm0
-4003ef:10 60 00 
-4003f2:48 83 c0 40          add    $0x40,%rax
-4003f6:62 f1 7d 48 fe 80 40 vpaddd 0x601440(%rax),%zmm0,%zmm0
-4003fd:14 60 00 
-400400:62 f1 fd 48 7f 80 40 vmovdqa64 %zmm0,0x601840(%rax)
-400407:18 60 00 
-40040a:48 3d 00 08 00 00    cmp    $0x800,%rax
-400410:75 d6                jne    4003e8 <main+0x8>
-400412:31 c0                xor    %eax,%eax
-400414:c3                   retq
+4003e2:62 f1 fd 48 6f 80 80 10 60 00 vmovdqa64 0x601080(%rax),%zmm0
+4003ec:48 83 c0 40                   add    $0x40,%rax
+4003f0:62 f1 7d 48 fe 80 40 18 60 00 vpaddd 0x601840(%rax),%zmm0,%zmm0
+4003fa:62 f1 fd 48 7f 80 40 20 60 00 vmovdqa64 %zmm0,0x602040(%rax)
+400404:48 3d 00 08 00 00             cmp    $0x800,%rax
+40040a:75 d6                         jne    4003e2 <main+0x2>
+40040c:31 c0                         xor    %eax,%eax
+40040e:c3                            retq
 
-baccala@ideapad-S510p:~$ ./ex2
+baccala@ideapad-S510p:~$ ./ex3
 Illegal instruction (core dumped)
-baccala@ideapad-S510p:~$ ./ex2
+baccala@ideapad-S510p:~$ ./ex3
 
@@
 \tiny
-void operation (int * a, int * b, int * c)
+void operation (int * ap, int * bp, int * cp)
 \{
   for (int i=0; i<16; i++) \{
-    c[i] = a[i] + b[i];
+    cp[i] = ap[i] + bp[i];
   \}
@@
 \end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt clDeviceQuery} on an Intel\textsuperscript{\textregistered} Core\texttrademark\ i5-4200U CPU}
+\begin{semiverbatim}
+\tiny
+  CL_DEVICE_NAME:                       Intel(R) HD Graphics Haswell Ultrabook GT2 Mobile
+  CL_DEVICE_VENDOR:                     Intel
+  CL_DRIVER_VERSION:                    1.1.1
+  CL_DEVICE_TYPE:                       CL_DEVICE_TYPE_GPU
+  CL_DEVICE_MAX_COMPUTE_UNITS:          20
+  CL_DEVICE_MAX_WORK_ITEM_DIMENSIONS:   3
+  CL_DEVICE_MAX_WORK_ITEM_SIZES:        512 / 512 / 512 
+  CL_DEVICE_MAX_WORK_GROUP_SIZE:        512
+  CL_DEVICE_MAX_CLOCK_FREQUENCY:        1000 MHz
+  CL_DEVICE_ADDRESS_BITS:               32
+  CL_DEVICE_MAX_MEM_ALLOC_SIZE:         1024 MByte
+  CL_DEVICE_GLOBAL_MEM_SIZE:            2048 MByte
+  CL_DEVICE_ERROR_CORRECTION_SUPPORT:   no
+  CL_DEVICE_LOCAL_MEM_TYPE:             global
+  CL_DEVICE_LOCAL_MEM_SIZE:             64 KByte
+  CL_DEVICE_MAX_CONSTANT_BUFFER_SIZE:   131072 KByte
+  CL_DEVICE_QUEUE_PROPERTIES:           CL_QUEUE_PROFILING_ENABLE
+  CL_DEVICE_IMAGE_SUPPORT:              1
+  CL_DEVICE_MAX_READ_IMAGE_ARGS:        128
+  CL_DEVICE_MAX_WRITE_IMAGE_ARGS:       8
+
+  CL_DEVICE_IMAGE <dim>                 2D_MAX_WIDTH     8192
+                                        2D_MAX_HEIGHT    8192
+                                        3D_MAX_WIDTH     8192
+                                        3D_MAX_HEIGHT    8192
+                                        3D_MAX_DEPTH     2048
+
+  CL_DEVICE_PREFERRED_VECTOR_WIDTH_<t>  CHAR 16, SHORT 8, INT 4, LONG 2, FLOAT 4, DOUBLE 0
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}
+\frametitle{Cluster Computing - History}
+\begin{itemize}
+\item Beowulf
+\item Hadoop
+\item Cluster Linux
+\begin{itemize}
+\item Kerrighed - Latest version is 3.0.0 which was released on 14 June 2010 and based on Linux kernel 2.6.30.
+\item OpenSSI - last development release 2010; last stable release 2007
+\end{itemize}
+\end{itemize}
 \end{frame}
@@
 
+\begin{frame}[fragile]
+\frametitle{Single System Image}
+\begin{definition}
+A single system image (SSI) cluster is a cluster of machines that appears to be one single system.
+
+\hfill Wikipedia
+\end{definition}
+\begin{center}
+\begin{tikzpicture}
+\node (a) [draw] at (0:1) {};
+\node (b) [draw] at (60:1) {} edge (a);
+\node (c) [draw] at (120:1) {} edge (b);
+\node (d) [draw] at (180:1) {} edge (c);
+\node (e) [draw] at (240:1) {} edge (d);
+\node (f) [draw] at (300:1) {} edge (e) edge (a);
+\end{tikzpicture}
+\end{center}
+\begin{semiverbatim}
+\$ nproc
+96
+\$ free -h
+              total        used        free
+Mem:           768G         45G        723G
+Swap:            0B          0B          0B
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}
+\frametitle{Single System Image}
+
+\begin{itemize}
+\item Key Features
+\begin{itemize}
+\item Cluster-wide login
+\item Single Process Space
+\item Process Migration
+\item Single Root Filesystem
+\item Redundant Filesystem
+\item Single I/O Space - block devices
+\item Single IPC Space
+standard Linux inter-process communication mechanisms, shared memory, semaphores, SYSV message queues, pipes and Unix domain sockets.
+\item Cluster-wide IP Addresses
+\end{itemize}
+\item Advantages
+\begin{itemize}
+\item familiar programming and management tools
+\item ease of leveraging additional computing power
+\end{itemize}
+\item Disadvantages
+\begin{itemize}
+\item less fault isolation
+\item node failure more difficult
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Single System Image - History}
+\begin{itemize}
+\item VMS
+\item SCALEMP
+\item Cluster Linux
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{GNU Hurd}
+\begin{itemize}
+\item History
+\item Microkernels in general
+\item CMU Mach
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{CMU Mach}
+\begin{itemize}
+\item Tasks / Threads
+\item Interprocess Communication
+\item Memory Management
+\end{itemize}
+\end{frame}
+
 % tikz commands to draw task boxes to illustrate Mach IPC concepts
@@
 \tikzstyle{task} = [draw, fill=blue!20, rounded corners]
+
+\tikzstyle{kernel} = [draw, fill=blue!20, rounded corners, minimum width=30em, minimum height=8em]
 
@@
 % \task - takes four arguments
-% 1. name of task
+% 1. name of task (used to refer to it from tikz; maybe printed as a label as well)
 % 2. location of task box's center point
@@
 %    ports should drawn on the left or right side of the task box
-%    "u" to make the task unlabled
+%    "u" to label the task "Task" instead of labeling it with its name
+%    "b" to make a "big" node (I wish I knew how to size the tick marks automatically)
 % 4. contents of the task box - other nodes
@@
 \newcommand\task[4]{
-\matrix (#1) [task, inner xsep=0] at (#2) { \node {\IfSubStr{#3}{u}{}{#1}}; \\ #4
+\matrix (#1) [task, inner xsep=0] at (#2) { \node {\IfSubStr{#3}{u}{Task}{#1}}; \\ #4
 
@@
 
+\newcommand\bigtask[4]{
+\matrix (#1) [task, inner xsep=0] at (#2) { \node {\IfSubStr{#3}{u}{Task}{#1}}; \\ #4
+
+% print port tick marks in fill color if they are disabled, in order to make
+% them expand out the dimensions of the task box yet remain invisible
+
+\foreach \i in {-20,...,7}
+{
+\IfSubStr{#3}{l}{
+  \draw[xshift=-5em, ultra thick] (0,0.2 * \i)  coordinate (#1 Port \i) -- (0.2,0.2 * \i);
+}{
+  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i);
+}
+\IfSubStr{#3}{r}{
+  \draw[xshift=5em, ultra thick] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate (#1 Port \i);
+}{
+  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i);
+}
+}
+
+\\};
+}
+
+% draw a box with vertical lines to represent a message queue
+% arg 1 is the name of a port to position it to the left of
+% arg 2 is name to be given the box
+
+\newcommand\queue[2]{
+\filldraw (#1) +(-3.1,0) node[rectangle, name=#2, minimum width=1.8cm, minimum height=1cm, draw, pattern=vertical lines] {};
+}
+
 \begin{frame}
@@
 
-\begin{tikzpicture}[transform canvas={scale=0.4, yshift=10em, xshift=-15em}]
+\begin{tikzpicture}[transform canvas={scale=0.4, yshift=12em, xshift=-15em}]
 
@@
 
-\draw [color=violet, thick, -triangle 90] (Task 1 left Port 5) -- (Task 1 right Port 5);
-\draw [color=violet, thick] (Task 2 left Port 0) -- +(1,0) |- (Task 1 right Port 5);
-\draw [color=violet, thick, -triangle 90] (Task 2 left Port 5) -- (Task 2 right Port 5);
-\draw [color=violet, thick, triangle 90-] (Task 2 left Port -5) -- (Task 2 right Port -5);
+\queue{Task 1 right Port 5}{queue 1}
+\queue{Task 2 right Port -5}{queue 2a}
+\queue{Task 2 right Port 5}{queue 2b}
+\queue{Task 3 right Port 0}{queue 3}
+\queue{Task 4 right Port 0}{queue 4}
 
-\filldraw[fill=white] (Task 1 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
-\filldraw[fill=white, pattern=vertical lines] (Task 1 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
+\draw [color=violet, thick, -triangle 90] (Task 1 left Port 5) -- (queue 1.west);
+\draw [color=violet, thick, -triangle 90] (queue 1.east) -- (Task 1 right Port 5);
+\draw [color=violet, thick] (Task 2 left Port 0) -- +(1,0) |- (queue 1.west);
+\draw [color=violet, thick, -triangle 90] (Task 2 left Port -3) -- +(1,0) |- (queue 4.west);
 
-\filldraw[fill=white] (Task 2 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
-\filldraw[fill=white, pattern=vertical lines] (Task 2 right Port 5) ++(-4,-0.5) rectangle +(1.8,1);
+\draw [color=violet, thick, -triangle 90] (Task 2 left Port 5) -- (queue 2b.west);
+\draw [color=violet, thick, triangle 90-] (Task 2 left Port -5) -- (queue 2a.west);
+\draw [color=violet, thick, -triangle 90] (queue 2b.east) -- (Task 2 right Port 5);
+\draw [color=violet, thick, triangle 90-] (queue 2a.east) -- (Task 2 right Port -5);
 
-\filldraw[fill=white] (Task 2 right Port -5) ++(-4,-0.5) rectangle +(1.8,1);
-\filldraw[fill=white, pattern=vertical lines] (Task 2 right Port -5) ++(-4,-0.5) rectangle +(1.8,1);
+\draw [color=violet, thick, -triangle 90] (Task 3 left Port 0) -- (queue 3.west);
+\draw [color=violet, thick, -triangle 90] (queue 3.east) -- (Task 3 right Port 0);
+
+\draw [color=violet, thick, -triangle 90] (Task 3 right Port 0) ++(-1,0) |- ++(-4,1) -- ++(0,-1);
+
+\draw [color=violet, thick, -triangle 90] (queue 4.east) -- (Task 4 right Port 0);
+
+\draw [color=violet, thick, -triangle 90] (Task 4 left Port 3) -- +(2,0) |- +(6,5.5) |- (queue 2a.east);
 
@@
 \begin{frame}
+\frametitle{Transfering a Send Right}
+\begin{center}
+\begin{tikzpicture}[node distance = .5cm]
 
-\frametitle{Send Rights}
-
-\begin{center}
-
-\begin{tikzpicture}
-
-\task{Task 1}{0,0}{r}{
+\bigtask{Task 1}{0,0}{r}{
 % print two sendright nodes, one to fill the message1 box to the correct size,
@@
 % first sendright
-\node (sendright) [draw, fill=orange] {Send 41};
-\node (message1) [fit=(sendright), message] {Message 1};
-\node (sendright) [draw, fill=orange] {Send 41};
-
-\node (message3) at (0, -7em) [message] {Message 2};
-}
-
-\task{Task 2}{15 em,0}{l}{
-\node (sendright2) [draw, fill=orange] {Send 9};
-\node (message2) [fit=(sendright2), message] {Message 1};
-\node (sendright2) [draw, fill=orange] {Send 9};
-
-\node (message4) at (0, -7em) [message] {Message 2};
-}
-
-\draw [color=violet, thick, ->] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 5)
-    -- +(.5,0) node [at end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);
-
-\draw [color=violet, thick, <-] (node cs:name=message3,angle=20) -- +(.3,0) |- (Task 1 Port -5) node [midway, anchor=south, color=black] {41} -- (Task 2 Port -5)
-    -- +(.5,0) node [near end, anchor=south, color=black] {9} |- (node cs:name=message4, angle=160);
-
-\end{tikzpicture}
-\end{center}
-
-\end{frame}
-
-\begin{frame}
-
-\frametitle{Send Rights}
-
-\begin{center}
-
-\begin{tikzpicture}
-
-\task{Task 1}{0,0}{r}{
-% print two sendright nodes, one to fill the message1 box to the correct size,
-% and another one to actually draw it, since message1 will overwrite the
-% first sendright
+\only<2->{
 \node (sendright) [draw, fill=orange] {Send 41};
@@
 }
-\task{Task 2}{15 em,0}{l}{
+
+\only<4->{\node (message3) at (0, -7em) [message] {Message 2};}
+}
+
+\bigtask{Task 2}{15 em,0}{l}{
+\only<3->{
 \node (sendright2) [draw, fill=orange] {Send 9};
@@
 \node (sendright2) [draw, fill=orange] {Send 9};
-%\node (message2) [message] {Message 1};
-% \node (message2a) [message] at (1 em, -2 em) {Message 2};
-}
-\task{Task 1 }{0,-11 em}{r}{
-\node (message3) [message] {Message 2};
-}
-\task{Task 2 }{15 em,-11 em}{l}{
-\node (message4) [message] {Message 2};
 }
 
-\draw [color=violet, thick, ->] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) node [midway, anchor=south, color=black] {27} -- (Task 2 Port 5)
-    -- +(.5,0) node [at end, anchor=south, color=black] {52} |- (node cs:name=message2, angle=160);
+\only<4->{\node (message4) at (0, -7em) [message] {Message 2};}
+}
 
-%\draw [color=violet, thick, ->] (node cs:name=message3,angle=0) -- +(.3,0) |- (Task 3 Port 0) node [midway, anchor=south, color=black] {18} -- +(1,0) |- (Task 2 Port 3)
-%    -- +(.3,0) |- (node cs:name=message2a, angle=160);
+\only<2->{\node (port 41) [left of=Task 1 Port -10, anchor=south] {41};}
+
+\only<2->{\draw [thick, color=red, dotted, ->] (node cs:name=sendright, angle=-20) -- (port 41);}
+
+\node [left of=Task 1 Port 5, anchor=south] {27};
+\node [right of=Task 2 Port 5, anchor=south] {52};
+\only<1,2>{\draw [thick, -triangle 90] (Task 1 Port 5) -- (Task 2 Port 5);}
+
+\only<3->{
+\draw [color=violet, thick, -triangle 90] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) -- (Task 2 Port 5)
+    -- +(.5,0) |- (node cs:name=message2, angle=160);
+}
+
+\only<3->{\node [right of=Task 2 Port -10, anchor=south] {9};}
+
+\only<3>{\draw [thick, triangle 90-] (Task 1 Port -10) -- (Task 2 Port -10);}
+
+\only<4->{
+\draw [color=violet, thick, triangle 90-] (node cs:name=message3,angle=20) -- +(.3,0) |- (Task 1 Port -10) -- (Task 2 Port -10)
+    -- +(.5,0) |- (node cs:name=message4, angle=160);
+}
 
@@
 \end{center}
-
 \end{frame}
@@
 \begin{frame}
-\frametitle{Right Transfers}
+\frametitle{Transfering a Receive Right}
+\begin{center}
+\begin{tikzpicture}[node distance = .5cm]
+
+\bigtask{Task 1}{0,0}{r}{
+% print two sendright nodes, one to fill the message1 box to the correct size,
+% and another one to actually draw it, since message1 will overwrite the
+% first sendright
+\only<2->{
+\node (sendright) [draw, fill=green] {Recv 41};
+\node (message1) [fit=(sendright), message] {Message 1};
+\node (sendright) [draw, fill=green] {Recv 41};
+}
+
+\only<4->{\node (message3) at (0, -7em) [message] {Message 2};}
+}
+
+\bigtask{Task 2}{15 em,0}{l}{
+\only<3->{
+\node (sendright2) [draw, fill=green] {Recv 9};
+\node (message2) [fit=(sendright2), message] {Message 1};
+\node (sendright2) [draw, fill=green] {Recv 9};
+}
+
+\only<4->{\node (message4) at (0, -7em) [message] {Message 2};}
+}
+
+\only<2->{\node (port 41) [left of=Task 1 Port -10, anchor=south] {41};}
+
+\only<2->{\draw [thick, color=red, dotted, ->] (node cs:name=sendright, angle=-20) -- (port 41);}
+
+\node [left of=Task 1 Port 5, anchor=south] {27};
+\node [right of=Task 2 Port 5, anchor=south] {52};
+\only<1,2>{\draw [thick, -triangle 90] (Task 1 Port 5) -- (Task 2 Port 5);}
+
+\only<3->{
+\draw [color=violet, thick, -triangle 90] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) -- (Task 2 Port 5)
+    -- +(.5,0) |- (node cs:name=message2, angle=160);
+}
+
+\only<3->{\node [right of=Task 2 Port -10, anchor=south] {9};}
+
+\only<3>{\draw [thick, -triangle 90] (Task 1 Port -10) -- (Task 2 Port -10);}
+
+\only<4->{
+\draw [color=violet, thick, -triangle 90] (node cs:name=message3,angle=20) -- +(.3,0) |- (Task 1 Port -10) -- (Task 2 Port -10)
+    -- +(.5,0) |- (node cs:name=message4, angle=160);
+}
+
+\end{tikzpicture}
+\end{center}
 \end{frame}
@@
 \frametitle{Memory Transfers}
+\begin{center}
+
+\begin{tikzpicture}[node distance = .5cm]
+
+\bigtask{Task 1}{0,0}{r}{
+% print two sendright nodes, one to fill the message1 box to the correct size,
+% and another one to actually draw it, since message1 will overwrite the
+% first sendright
+\node (message1) [message] {Message};
+\only<1>{
+\node (memory1) [below=1em, fill=orange!40, minimum size=2cm, dashed, thick, draw] at (message1.south) {Memory};
+\draw [thick, color=red, dotted, -] (message1) -- (memory1);
+}
+}
+
+\bigtask{Task 2}{15 em,0}{l}{
+\only<2->{
+\node (message2) [message] {Message};
+\node (memory2) [below=1em, fill=orange!40, minimum size=2cm, dashed, thick, draw] at (message2.south) {Memory};
+\draw [thick, color=red, dotted, -] (message2) -- (memory2);
+}
+}
+
+%\node [left of=Task 1 Port 5, anchor=south] {27};
+%\node [right of=Task 2 Port 5, anchor=south] {52};
+
+%\only<1>{\draw [thick, -triangle 90] (Task 1 Port 5) -- (Task 2 Port 5);}
+
+\only<2>{
+\draw [color=violet, thick, -triangle 90] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) -- (Task 2 Port 5)
+    -- +(.5,0) |- (node cs:name=message2, angle=160);
+}
+\end{tikzpicture}
+\end{center}
 \end{frame}
@@
 \frametitle{Hurd Filesystem}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt rpctrace} of {\tt ls} (9 of 144 RPCs)}
+\begin{semiverbatim}
+\tiny
+task135(pid9403)->task_get_special_port (4) = 0    143(pid9403)
+
+  143(pid9403)->exec_startup_get_info () = 0 134526367 134512692 288 188416 16777216 0 "ls\\0"
+        "LS_COLORS=rs=0:di=01;34:ln=01;36:mh=00:pi=40;33:so=01;35:do=01;35:bd=40;33;01:cd"
+        \{  140(pid9403)   145(pid9403)   146(pid9403)   147(pid9403)\}
+        \{  148(pid9403)   149(pid9403)   150(pid9403)   151(pid9403)   152(pid9403) (null)\}
+        \{18 0 0 0 0\}
+
+/* Standard port assignments for file_exec_file_name and exec_* */
+enum
+  \{
+    INIT_PORT_CWDIR,
+    INIT_PORT_CRDIR,
+    INIT_PORT_AUTH,
+    INIT_PORT_PROC,
+    INIT_PORT_CTTYID,
+    INIT_PORT_BOOTSTRAP,
+  \};
+
+  145(pid9403)->term_open_ctty (9403 9402) = 0    138(pid9403)
+
+  148(pid9403)->dir_lookup ("./" 2097161 0) = 0 1 "    166(pid9403)
+
+  166(pid9403)->dir_readdir (0 -1 0) = 0 "\\x01\\x9c\\x01\\0\\0\\0\\0\\0\\x10\\0\\0\\x01.\\0\\0\\0P\\xd7\\0\\0\\0\\0\\0\\0\\x10\\0\\0\\x02..\\0\\0\\x03\\x9c\\x01\\0\\0\\0\\0\\0\\x14\\0\\0\\x04.git\\0\\0\\0\\0\\x90_\\a\\0\\0\\0\\0\\0\\x18\\0\\0\\blibpager\\0\\0\\0\\0c\\x9a\\x01\\0" 10
+
+  166(pid9403)->dir_readdir (10 -1 0) = 0 " 0
+
+  145(pid9403)->io_stat () = 0 {14 969 0 0 0 1458877724 0 8397184 1 0 5 0 0 1515530623 0 1515530623 0 1515530623 0 512 8 0 0 0 0 0 1103856 1038 1163584 1039 0 0}
+
+  138(pid9403)->io_write ("junk  libpager machMessage.h  netmsg  papers  patches README.md\\n" -1) = 0 64
+
+task135(pid9403)->task_terminate () = 0 
+\end{semiverbatim}
 \end{frame}
@@
 \frametitle{Mach Memory Management}
+\begin{center}
+\begin{tikzpicture}
+
+\task{Filesystem Server}{0,0}{}{
+\node (memory1) [above left=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
+}
+\task{ls}{15 em,0}{}{
+%\node (memory1) [above right=-0.5, fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
+\node (memory2) [above right=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
+\node [below] at (memory2.north) {\tiny addr};
+}
+
+\node (kernel) [kernel] at (7.5 em, -10 em) {};
+\node (kmemory) [fill=white, minimum size=1.5cm, dashed, thick, draw] at (7.5 em, -10 em) {};
+\node [below] at (kernel.north) {Kernel};
+
+\draw [color=red, <-, thick] (2 em,1) -- node [midway, above] {io\_map} (13 em,1);
+\draw [color=red, ->, thick, dotted] (2 em,.9) -- node [midway, below] {\tiny port 153} (13 em,.9);
+
+\draw [color=red, ->, thick] (15 em,-3 em) -- node [midway, left] {vm\_map} (15 em,- 8 em);
+\draw [color=red, <-, thick, dotted] (15.2 em,-3 em) -- node [midway, right] {\tiny addr} (15.2 em,- 8 em);
+
+\draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 8 em);
+\draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 8 em);
+
+\end{tikzpicture}
+\end{center}
 \end{frame}
@@
 \frametitle{Example: Shared Libraries}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt rpctrace} of {\tt ls} (mapping the C library)}
+\begin{semiverbatim}
+\tiny
+  149(pid9403)->dir_lookup ("lib/i386-gnu/libc.so.0.3" 1 0) = 0 1 ""    155(pid9403)
+
+  155(pid9403)->io_read (-1 512) = 0 "\\x7fELF\\x01\\x01\\x01\\0\\0\\0\\0\\0\\0\\0\\0\\0\\x03\\0\\x03\\0\\x01\\0\\0\\0pE\\x05\\04\\0\\0\\0\\xfc\\x19''\\0\\0\\0\\0\\04\\0 \\0\\n\\0(\\0x\\0w\\0\\x06\\0\\0\\04\\0\\0\\04\\0\\0\\04\\0\\0\\0@\\x01\\0\\0@\\x01\\0\\0\\x05\\0\\0\\0"
+
+  155(pid9403)->io_stat () = 0 {23 5 0 158020 0 1513754748 0 33261 1 0 0 2239676 0 1515530623 0 1513435063 0 1513747658 0 8192 4384 0 0 0 0 0 0 0 0 0 0 0}
+
+  155(pid9403)->io_map () = 0    153(pid9403)  (null)
+
+task135(pid9403)->vm_map (0 2245448 0 1    153(pid9403) 0 32 5 7 1) = 0 17039360
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}
+\frametitle{The {\tt netmsg} Server}
 \end{frame}
@@
 
+\begin{frame}
+\frametitle{What kind of bugs have I found?}
+\begin{itemize}
+\item failure to update pagemap entry
+\item FLUSHING flag
+\item DEAD NAME notifications
+\item condition variable
+\item weak refs
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Work Needed / Help Wanted}
+\begin{itemize}
+\item 64-bit user space
+\item Multiprocessor support
+\item Rewrite of {\tt netmsg}
+\end{itemize}
+\end{frame}
+
 % https://events.ccc.de/congress/2007/Fahrplan/attachments/986_inside_the_mac_osx_kernel.pdf

### 2018-01-17 papers/cua.tex
cua lecture: version of the slides for the presentation

@@
 \usepackage{tikz}
-\usetikzlibrary{positioning, fit, backgrounds, arrows, patterns}
+\usetikzlibrary{positioning, fit, backgrounds, arrows, patterns, trees, shapes}
 
@@
 \begin{frame}
-\frametitle{Cisco 5108 Blade Server Chassis with eight B200 M2s}
-\includegraphics[width=\textwidth]{ourcisco.jpg}
-\end{frame}
-
-\begin{frame}
 \frametitle{Processor families in TOP500}
@@
 \begin{frame}
-\frametitle{Techniques to exploit parallelism}
+\frametitle{Techniques to exploit parallelism (partial list)}
 \begin{itemize}
@@
 \item Superscalar vector operations
-\item Speculative execution
+\end{itemize}
+\item Thread level parallelism
+\begin{itemize}
+\item Multiple cores with shared memory
+\end{itemize}
 \item Co-processors
-\end{itemize}
-\item Thread level parallelism
-\begin{itemize}
-\item Shared memory
-\item Caching
-\item Synchronization
+\begin{itemize}
+\item More cores; simpler cores
 \end{itemize}
@@
 \begin{itemize}
-\item No shared memory
+\item Multiple cores without shared memory
 \item High speed LAN
@@
 
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Adding 512 64-bit integers -- Multi-threaded Code}
+\begin{semiverbatim}
+\tiny
+#include <thread>
+
+int a[512], b[512], c[512];
+
+int operation(int * ap, int * bp, int * cp, int count)
+\{
+  for (int i = 0; i < count; i++) \{
+    cp[i] = ap[i] + bp[i];
+  \}
+\}
+
+int main () \{
+  const int num_threads = 3;
+  int start_index[num_threads+1];
+  std::thread thread[num_threads];
+
+  start_index[0] = 0;
+  for (int i=1; i < num_threads; i++) \{
+    start_index[i] = i*(512/num_threads) & ~15;
+  \}
+  start_index[num_threads] = 512;
+
+  for (int i=0; i < num_threads; i++) \{
+    thread[i] = std::thread(operation,
+                            &a[start_index[i]], &b[start_index[i]], &c[start_index[i]],
+                            start_index[i+1] - start_index[i]);
+  \}
+
+  for (int i=0; i < num_threads; i++) \{
+    thread[i].join();
+  \}
+\}
 \end{semiverbatim}
@@
 
-\begin{frame}
-\frametitle{Cluster Computing - History}
-\begin{itemize}
+\begin{frame}[fragile]
+\frametitle{Adding 512 64-bit integers -- OpenCL kernel}
+\begin{semiverbatim}
+\tiny
+void kernel simple_add(global const int* A, global const int* B, global int* C,
+                       global const int* N) \{
+    int ID, Nthreads, n, ratio, start, stop;
+
+    ID = get_global_id(0);
+    Nthreads = get_global_size(0);
+    n = N[0];
+
+    ratio = (n / Nthreads);  // number of elements for each thread
+    start = ratio * ID;
+    stop  = ratio * (ID + 1);
+
+    for (int i=start; i<stop; i++)
+        C[i] = A[i] + B[i];
+\}
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Adding 512 64-bit integers -- OpenCL program}
+\begin{semiverbatim}
+\tiny
+const int n = 512;
+int A[n], B[n], C[n];
+
+cl::Program program(context, sources);
+program.build({default_device});
+
+int N[1] = {n};
+
+// create buffers on device (allocate space on GPU)
+cl::Buffer buffer_A(context, CL_MEM_READ_WRITE, sizeof(int) * n);
+cl::Buffer buffer_B(context, CL_MEM_READ_WRITE, sizeof(int) * n);
+cl::Buffer buffer_C(context, CL_MEM_READ_WRITE, sizeof(int) * n);
+cl::Buffer buffer_N(context, CL_MEM_READ_ONLY,  sizeof(int));
+
+// create a queue (a queue of commands that the GPU will execute)
+cl::CommandQueue queue(context, default_device);
+
+// push write commands to queue
+queue.enqueueWriteBuffer(buffer_A, CL_TRUE, 0, sizeof(int)*n, A);
+queue.enqueueWriteBuffer(buffer_B, CL_TRUE, 0, sizeof(int)*n, B);
+queue.enqueueWriteBuffer(buffer_N, CL_TRUE, 0, sizeof(int),   N);
+
+cl::Kernel simple_add(program, "simple_add");
+simple_add.setArg(0, buffer_A);
+simple_add.setArg(1, buffer_B);
+simple_add.setArg(2, buffer_C);
+queue.enqueueNDRangeKernel(simple_add, cl::NullRange, cl::NDRange(20), cl::NullRange);
+
+// read result from GPU to here
+queue.enqueueReadBuffer(buffer_C, CL_TRUE, 0, sizeof(int)*n, C);
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}
+\frametitle{Techniques to exploit parallelism (partial list)}
+\begin{itemize}
+\item Instruction level parallelism
+\begin{itemize}
+\item Superscalar vector operations
+\item \alert{Short {\tt for} loops}
+\end{itemize}
+\item Thread level parallelism
+\begin{itemize}
+\item Multiple cores with shared memory
+\item \alert{Synchronization}
+\end{itemize}
+\item Co-processors
+\begin{itemize}
+\item More cores; simpler cores
+\item \alert{OpenCL}
+\end{itemize}
+\item Node level parallelism
+\begin{itemize}
+\item Multiple cores without shared memory
+\item High speed LAN
+\item Clusters
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Cisco 5108 Blade Server Chassis with eight B200 M2s}
+\includegraphics[width=\textwidth]{ourcisco.jpg}
+\end{frame}
+
+\begin{frame}
+\frametitle{Cluster Computing - Partial History}
+\begin{itemize}
+\item VMS
 \item Beowulf
+\begin{itemize}
+\item Pioneered in 1994 at NASA
+\item Message Passing Interface (MPI)
+\item Parallel Virtual Machine (PVM)
+\end{itemize}
 \item Hadoop
-\item Cluster Linux
-\begin{itemize}
-\item Kerrighed - Latest version is 3.0.0 which was released on 14 June 2010 and based on Linux kernel 2.6.30.
-\item OpenSSI - last development release 2010; last stable release 2007
-\end{itemize}
+\item Single System Image
 \end{itemize}
@@
 \item less fault isolation
-\item node failure more difficult
-\end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Single System Image - History}
-\begin{itemize}
-\item VMS
-\item SCALEMP
-\item Cluster Linux
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{GNU Hurd}
-\begin{itemize}
-\item History
-\item Microkernels in general
-\item CMU Mach
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{CMU Mach}
-\begin{itemize}
-\item Tasks / Threads
-\item Interprocess Communication
-\item Memory Management
-\end{itemize}
+\item node failure more difficult to handle
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{POSIX System System Image -- Partial History}
+\begin{itemize}
+\item Single System Image Linux
+\begin{itemize}
+\item MOSIX
+\begin{itemize}
+\item Developed at Hebrew University at Jerusalem since 1977
+\item Closed source; free download
+\item Latest version works with Linux 4.12.13; released in 2017
+\end{itemize}
+\item OpenMosix
+\begin{itemize}
+\item Forked from last free version of MOSIX in 2001
+\item Last development release 2006 (based on Linux 2.6.15)
+\item Last stable release 2004 (based on Linux 2.4.26; 38,794 line diff)
+\item project terminated in 2008
+\end{itemize}
+\item Kerrighed
+\begin{itemize}
+\item {\it Institut national de recherche en informatique et en automatique}
+\item Latest version (3.0.0) released on 14 June 2010
+\item Based on Linux 2.6.30
+\end{itemize}
+\item OpenSSI
+\begin{itemize}
+\item Funded by SCO; Compaq
+\item Last development release 2010 (based on Linux 2.6.12)
+\item Last stable release 2007 (based on Linux 2.6.11)
+\end{itemize}
+\end{itemize}
+\item ScaleMP
+% \item CMU Mach / GNU Hurd
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{CMU Mach / GNU Hurd}
+\begin{itemize}
+\item{CMU Mach}
+\begin{itemize}
+\item Developed at Carnegie Mellon University from 1985 to 1994
+\item Pioneered the Microkernel concept
+\item Kernel does Tasks / Threads / IPC / Memory Management
+\item Everything else in user space (device drivers; filesystem; networking)
+\item Designed for multi-node clusters
+\end{itemize}
+\item{GNU Hurd}
+\begin{itemize}
+\item Under development since 1990
+\item Runs on top of Mach and implements a full POSIX system
+\item Averages about two releases per year
+\item Available as part of Debian
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{How many Debian packages build on Hurd?}
+\includegraphics[width=\textwidth]{graph-big.png}
 \end{frame}
@@
 %    "b" to make a "big" node (I wish I knew how to size the tick marks automatically)
+%    "s" to make a "small" node
 % 4. contents of the task box - other nodes
@@
 % The tick marks will be named "#1 Port #" for future reference
+%
+% It's implemented as a tikz matrix with two cells - the task name in the top cell
+% and everything else in the bottom cell
 
@@
 
-\foreach \i in {-7,...,7}
+\IfSubStr{#3}{b}{\edef\starti{-20}}{\IfSubStr{#3}{s}{\edef\starti{-1}}{\edef\starti{-7}}}
+\IfSubStr{#3}{s}{\edef\endi{1}}{\edef\endi{7}}
+
+\foreach \i in {\starti,...,\endi}
 {
-\IfSubStr{#3}{l}{
-  \draw[xshift=-5em, ultra thick] (0,0.2 * \i)  coordinate (#1 Port \i) -- (0.2,0.2 * \i);
-}{
-  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i);
-}
-\IfSubStr{#3}{r}{
-  \draw[xshift=5em, ultra thick] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate (#1 Port \i);
-}{
-  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i);
-}
-}
-
-\\};
-}
-
-\newcommand\bigtask[4]{
-\matrix (#1) [task, inner xsep=0] at (#2) { \node {\IfSubStr{#3}{u}{Task}{#1}}; \\ #4
-
-% print port tick marks in fill color if they are disabled, in order to make
-% them expand out the dimensions of the task box yet remain invisible
-
-\foreach \i in {-20,...,7}
-{
-\IfSubStr{#3}{l}{
-  \draw[xshift=-5em, ultra thick] (0,0.2 * \i)  coordinate (#1 Port \i) -- (0.2,0.2 * \i);
-}{
-  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i);
-}
-\IfSubStr{#3}{r}{
-  \draw[xshift=5em, ultra thick] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate (#1 Port \i);
-}{
-  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i);
-}
+  \IfSubStr{#3}{l}{\edef\leftcolor{black}\edef\leftcoord{coordinate (#1 Port \i)}}
+                  {\edef\leftcolor{blue!20}\edef\leftcoord{}}
+  \IfSubStr{#3}{r}{\edef\rightcolor{black}\edef\rightcoord{coordinate (#1 Port \i)}}
+                  {\edef\rightcolor{blue!20}\edef\rightcoord{}}
+  \draw[xshift=-5em, ultra thick, color=\leftcolor] (0,0.2 * \i) \leftcoord -- (0.2,0.2 * \i);
+  \draw[xshift=5em, ultra thick, color=\rightcolor] (-0.2,0.2 * \i) -- (0,0.2 * \i) \rightcoord;
 }
@@
 
-\bigtask{Task 1}{0,0}{r}{
+\task{Task 1}{0,0}{br}{
 % print two sendright nodes, one to fill the message1 box to the correct size,
@@
 
-\bigtask{Task 2}{15 em,0}{l}{
+\task{Task 2}{15 em,0}{bl}{
 \only<3->{
@@
 
-\bigtask{Task 2}{15 em,0}{l}{
+\task{Task 2}{15 em,0}{bl}{
 \only<2->{
@@
 \frametitle{Hurd Filesystem}
+\tikzstyle{dirent}=[draw=black,thick,anchor=west]
+\tikzstyle{selected}=[draw=red,fill=red!30]
+\tikzstyle{optional}=[dashed,fill=gray!50]
+\tikzstyle{link}=[red,dotted,->,thick]
+\tikzstyle{grow2}=[grow via three points={one child at (1.5,0) and two children at (1.5,0) and (1.5,-0.7)},
+  edge from parent path={(\tikzparentnode.east) -- +(.5,0) |- (\tikzchildnode.west)}]
+\begin{center}
+\begin{tikzpicture}
+  [grow via three points={one child at (0.5,-0.7) and
+  two children at (0.5,-0.7) and (0.5,-1.4)},
+  edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
+  \node[dirent] (root) {/}
+  child { node[dirent] {bin}
+    child [grow2] { node[dirent] (bash) {bash}}
+    child [grow2] { node[dirent] (date) {date}}
+    child [grow2] { node[dirent] (grep) {grep}}
+    child [grow2] { node[dirent] (ls) {ls}}
+  }
+  child { node[dirent] {etc}}
+  child { node[dirent] {lib}}
+  child { node[dirent] {tmp}}
+  child { node[dirent] (hurd) {hurd}}
+  child { node[dirent] (home) {home}}
+  child { node[dirent] (http) {http:}}
+  child { node[dirent] {servers}
+    child { node[dirent] {socket}
+      child { node[dirent] (inet) {inet}}
+    }
+  };
+
+\task{ext2 File Server}{20 em,-5 em}{l}{}
+\task{httpfs}{20 em, -12 em}{ls}{}
+\task{pfinet}{20 em, -17 em}{ls}{}
+
+\draw [link] (root) -- +(5,0) |- (ext2 File Server Port 7);
+\draw [link] (bash) -- +(1,0) |- (ext2 File Server Port 5);
+\draw [link] (date) -- +(1,0) |- (ext2 File Server Port 3);
+\draw [link] (grep) -- +(1.5,0) |- (ext2 File Server Port 1);
+\draw [link] (ls) -- +(2,0) |- (ext2 File Server Port -1);
+\draw [link] (hurd) -- +(2,0) -- (ext2 File Server Port -3);
+\draw [link] (home) -- +(3,0) -- (ext2 File Server Port -5);
+
+\draw [link] (http) -- +(2,0) |- (httpfs Port 0);
+\draw [link] (inet) -- (pfinet Port -1);
+
+\end{tikzpicture}
+\end{center}
 \end{frame}
@@
 
-\begin{frame}
-\frametitle{Example: Basic File I/O}
+%\begin{frame}
+%\frametitle{Example: Basic File I/O}
+%\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt rpctrace} of {\tt ls} (mapping the C library)}
+\begin{semiverbatim}
+\tiny
+  149(pid9403)->dir_lookup ("lib/i386-gnu/libc.so.0.3" 1 0) = 0 1 ""    155(pid9403)
+
+  155(pid9403)->io_read (-1 512) = 0 "\\x7fELF\\x01\\x01\\x01\\0\\0\\0\\0\\0\\0\\0\\0\\0\\x03\\0\\x03\\0\\x01\\0\\0\\0pE\\x05\\04\\0\\0\\0\\xfc\\x19''\\0\\0\\0\\0\\04\\0 \\0\\n\\0(\\0x\\0w\\0\\x06\\0\\0\\04\\0\\0\\04\\0\\0\\04\\0\\0\\0@\\x01\\0\\0@\\x01\\0\\0\\x05\\0\\0\\0"
+
+  155(pid9403)->io_stat () = 0 {23 5 0 158020 0 1513754748 0 33261 1 0 0 2239676 0 1515530623 0 1513435063 0 1513747658 0 8192 4384 0 0 0 0 0 0 0 0 0 0 0}
+
+  155(pid9403)->io_map () = 0    153(pid9403)  (null)
+
+task135(pid9403)->vm_map (0 2245448 0 1    153(pid9403) 0 32 5 7 1) = 0 17039360
+
+\end{semiverbatim}
 \end{frame}
@@
 
+\only<1>{
+  \draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 8 em);
   \draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 8 em);
-\draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 8 em);
+}
+\only<2>{
+  \draw [color=red, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
+  \draw [color=red, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);
+}
 
@@
 \begin{frame}
-\frametitle{Example: Shared Libraries}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{{\tt rpctrace} of {\tt ls} (mapping the C library)}
-\begin{semiverbatim}
-\tiny
-  149(pid9403)->dir_lookup ("lib/i386-gnu/libc.so.0.3" 1 0) = 0 1 ""    155(pid9403)
-
-  155(pid9403)->io_read (-1 512) = 0 "\\x7fELF\\x01\\x01\\x01\\0\\0\\0\\0\\0\\0\\0\\0\\0\\x03\\0\\x03\\0\\x01\\0\\0\\0pE\\x05\\04\\0\\0\\0\\xfc\\x19''\\0\\0\\0\\0\\04\\0 \\0\\n\\0(\\0x\\0w\\0\\x06\\0\\0\\04\\0\\0\\04\\0\\0\\04\\0\\0\\0@\\x01\\0\\0@\\x01\\0\\0\\x05\\0\\0\\0"
-
-  155(pid9403)->io_stat () = 0 {23 5 0 158020 0 1513754748 0 33261 1 0 0 2239676 0 1515530623 0 1513435063 0 1513747658 0 8192 4384 0 0 0 0 0 0 0 0 0 0 0}
-
-  155(pid9403)->io_map () = 0    153(pid9403)  (null)
-
-task135(pid9403)->vm_map (0 2245448 0 1    153(pid9403) 0 32 5 7 1) = 0 17039360
-
-\end{semiverbatim}
+\frametitle{Mach Memory Management: Shared Memory}
+%\begin{center}
+\begin{tikzpicture}[transform canvas={xshift=9.5em, yshift=2em}]
+
+\task{Task 1}{-5em,0}{}{
+\node (memory1) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
+}
+\task{Task 2}{6 em,0}{}{
+\node (memory2) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
+}
+\task{Task 3}{17 em,0}{}{
+\node (memory3) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
+}
+
+\node (kernel) [kernel] at (6 em, -10 em) {};
+\node (kmemory) [fill=orange!40, minimum size=1.5cm, thick, draw] at (6 em, -10 em) {};
+\node [below] at (kernel.north) {Kernel};
+
+\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory1);
+\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory2);
+\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory3);
+
+\end{tikzpicture}
+%\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Mach Memory Management: Distributed Shared Memory}
+
+\tikzstyle{emptymemory} = [fill=white, minimum size=1.5cm, dashed, thick, draw]
+\tikzstyle{readmemory} = [fill=orange!40, minimum size=1.5cm, thick, draw]
+\tikzstyle{writememory} = [fill=green!40, minimum size=1.5cm, thick, draw]
+
+\begin{tikzpicture}[transform canvas={xshift=12em, yshift=7em, scale=0.4}]
+
+\task{Task 1}{-5em,0}{u}{
+\only<1-2,9->{\node (memory1) [emptymemory] {\only<2>{\Huge R}};}
+\only<3-8> {\node (memory1) [readmemory, dashed] {};}
+}
+\task{Task 2}{6 em,0}{u}{
+\only<1-3,9->{\node (memory2) [emptymemory] {};}
+\only<4-8>{\node (memory2) [readmemory, dashed] {};}
+}
+\task{Task 3}{17 em,0}{u}{
+\only<1-3,9-13>{\node (memory3) [emptymemory] {\only<11->{\Huge W}};}
+\only<4-8>{\node (memory3) [readmemory, dashed] {};}
+\only<14->{\node (memory3) [writememory, dashed] {};}
+}
+
+\node (kernel) [kernel] at (6 em, -10 em) {};
+\only<1-2,9-13>{\node (kmemory) [fill=white, minimum size=1.5cm, dashed, thick, draw] at (6 em, -10 em) {};}
+\only<3-8>{\node (kmemory) [fill=orange!40, minimum size=1.5cm, thick, draw] at (6 em, -10 em) {};}
+\only<14>{\node (kmemory) [writememory] at (6 em, -10 em) {};}
+\node [below] at (kernel.north) {Kernel};
+
+\only<3-8>{\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory1);}
+\only<4-8>{\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory2);}
+\only<4-8>{\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory3);}
+\only<14->{\draw [green, ultra thick, dotted, ->] (kmemory) -- (memory3);}
+
+\node (server) [kernel, color=cyan!20, minimum width=20em] at (6 em, -20 em) {};
+\only<1-9>{\node (smemory) [readmemory] at (server) {};}
+\only<10-12,14>{\node (smemory) [emptymemory] at (server) {};}
+\only<13>{\node (smemory) [writememory] at (server) {};}
+\node [below] at (server.north) {Memory Object Server};
+
+\only<2>{\draw [red, ->] (kernel) -- node [midway, left] {\huge m\_o\_data\_request} (server);}
+\only<3>{\draw [red, <-] (kernel) -- node [midway, right] {\huge m\_o\_data\_supply} (server);}
+\only<8>{\draw [red, <-] (kernel) -- node [midway, right] {\huge m\_o\_lock\_request (flush)} (server);}
+\only<9>{\draw [red, ->] (kernel) -- node [midway, right] {\huge m\_o\_lock\_completed} (server);}
+\only<11>{\draw [red, ->] (kernel) -- node [midway, left] {\huge m\_o\_data\_request} (server);}
+\only<14>{\draw [red, <-] (kernel) -- node [midway, right] {\huge m\_o\_data\_supply} (server);}
+
+\node (kernel2) [kernel] at (6 em, -30 em) {};
+\only<1-5,13->{\node (kmemory2) [emptymemory] at (kernel2) {};}
+\only<6-9>{\node (kmemory2) [readmemory] at (kernel2) {};}
+\only<10-12>{\node (kmemory2) [writememory] at (kernel2) {};}
+\node [below] at (kernel2.north) {Kernel};
+
+\task{Task 4}{-5em,-40 em}{u}{
+\node (memory4) [emptymemory] {};
+}
+\task{Task 5}{6 em,-40 em}{u}{
+\only<1-5,13->{\node (memory5) [emptymemory] {\only<5>{\Huge R}};}
+\only<6-12>{\node (memory5) [readmemory, dashed] {};}
+}
+\task{Task 6}{17 em,-40 em}{u}{
+\only<1-9,13->{\node (memory6) [emptymemory] {\only<7-9>{\Huge W}};}
+\only<10-12>{\node (memory6) [writememory, dashed] {};}
+}
+
+\only<5>{\draw [red, ->] (kernel2) -- node [midway, left] {\huge m\_o\_data\_request} (server);}
+\only<6>{\draw [red, <-] (kernel2) -- node [midway, right] {\huge m\_o\_data\_supply} (server);}
+\only<7>{\draw [red, ->] (kernel2) -- node [midway, left] {\huge m\_o\_data\_unlock} (server);}
+\only<10>{\draw [red, <-] (kernel2) -- node [midway, right] {\huge m\_o\_lock\_request (+write)} (server);}
+\only<12>{\draw [red, <-] (kernel2) -- node [midway, right] {\huge m\_o\_lock\_request (flush)} (server);}
+\only<13>{\draw [red, ->] (kernel2) -- node [midway, left] {\huge m\_o\_data\_return} (server);}
+
+\only<6-12>{\draw [orange, ultra thick, dotted, ->] (kmemory2) -- (memory5);}
+\only<10-12>{\draw [green, ultra thick, dotted, ->] (kmemory2) -- (memory6);}
+
+\end{tikzpicture}
 \end{frame}
@@
 \frametitle{The {\tt netmsg} Server}
+\begin{tikzpicture}[transform canvas={xshift=12em, yshift=6em, scale=0.4}]
+
+\task{Task 1}{-5em,0}{u}{}
+\task{Task 2}{6 em,0}{u}{}
+\task{Task 3}{17 em,0}{u}{}
+
+\node (kernel) [kernel] at (6 em, -10 em) {};
+\node [above of=kernel] {\Huge netmsg};
+
+\node [cloud, draw,cloud puffs=10,cloud puff arc=120, aspect=2, inner ysep=1em, minimum width=20em, minimum height=10em]
+  at (6 em, -20 em) {};
+
+\node (kernel2) [kernel] at (6 em, -30 em) {};
+\node [above of=kernel2] {\Huge netmsg};
+
+\task{Task 4}{-5em,-40 em}{u}{}
+\task{Task 5}{6 em,-40 em}{u}{}
+\task{Task 6}{17 em,-40 em}{u}{}
+
+\draw [red] (Task 1.south) -- (kernel.center) -- (kernel2.center) -- (Task 4.north);
+\draw [red] (Task 2.south) -- (kernel.center) -- (kernel2.center) -- (Task 5.north);
+\draw [red] (Task 3.south) -- (kernel.center) -- (kernel2.center) -- (Task 6.north);
+
+\end{tikzpicture}
+\end{frame}
+
+\begin{frame}
+\frametitle{The {\tt netmsg} Server}
+\begin{itemize}
+\item All Mach messages serialized over a single TCP connection
+\item Transmit Mach messages (almost) unchanged
+\item Memory regions are appended at the end of the message
+\item Local port numbers are transmitted unchanged
+\item Remote port numbers have all their bits inverted
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{The {\tt netmsg} Server}
+\begin{center}
+\begin{tikzpicture}[transform canvas={xshift=0em, yshift=-3em, scale=0.4}]
+
+\fill[color=yellow!20] (-50 em, -10 em) rectangle (-8 em, 25 em);
+\node at (-21 em, 20em) {\Huge Node 1};
+\fill[color=yellow!20] (50 em, -10 em) rectangle (8 em, 25 em);
+\node at (21 em, 20em) {\Huge Node 2};
+
+\filldraw (-30 em, 15em) node[rectangle, name=queue 1, minimum width=1.8cm, minimum height=1cm, draw, pattern=vertical lines] {};
+% \task{Task 0}{-30 em,15 em}{ur}{}
+
+\task{Task 1}{-30 em,0}{r}{
+  \only<1-2>{
+    \node (sendright) [draw, fill=orange, inner xsep=5] {Send};
+    \node (message1) [fit=(sendright), message] {Message};
+    \only<1>{\node (sendright) [draw, fill=orange, inner xsep=5] {Send};}
+  }
+}
+\task{netmsg 1}{-15 em,0}{l}{
+  \only<2-3>{
+    \node (sendright2) [draw, fill=orange, inner xsep=5] {Send};
+    \node (message2) [fit=(sendright2), message] {Message};
+  }
+  \only<2->{\node (sendright2) [draw, fill=orange, inner xsep=5] {Send};}
+}
+
+\only<1>{\draw[red,dotted,->,ultra thick] (sendright) -- (queue 1);}
+\only<2>{\draw[ultra thick,->] (message1) -- (message2);}
+\only<2->{
+  \draw[red,dotted,->,ultra thick] (sendright2) -- (queue 1);
+}
+
+\node [cloud, draw,cloud puffs=10,cloud puff arc=120, aspect=2, inner ysep=1em, minimum width=15em, minimum height=8em]
+  at (0 em, 0 em) {};
+
+\only<4->{
+  \filldraw (15 em, 15em) node[rectangle, name=queue 2, minimum width=1.8cm, minimum height=1cm, draw, pattern=vertical lines] {};
+}
+
+\task{netmsg 2}{15 em,0}{r}{
+  \only<3-5>{
+    \node (sendright3) [draw, fill=orange, inner xsep=5] {Send};
+    \node (message3) [fit=(sendright3), message] {Message};
+    \only<4>{\node (sendright3) [draw, fill=orange, inner xsep=5] {Send};}
+  }
+  \only<4->{\node (rcvright3) [draw, fill=green, inner xsep=5] at (0,1.25) {Rcv};}
+}
+\task{Task 2}{30 em,0}{l}{
+  \only<5>{
+    \node (sendright4) [draw, fill=orange, inner xsep=5] {Send};
+    \node (message4) [fit=(sendright4), message] {Message};
+  }
+  \only<5-6>{\node (sendright4) [draw, fill=orange, inner xsep=5] {Send};}
+}
+
+\only<1>{\draw[ultra thick, dotted, ->] (Task 1) -- (Task 2);}
+
+\only<3>{\draw[ultra thick,->] (message2) -- (message3);}
+
+\only<4>{\draw[red,dotted,->,ultra thick] (sendright3.west) to [bend left=20] (queue 2);}
+\only<4->{\draw[green,dotted,->,ultra thick] (queue 2) -- (rcvright3);}
+
+\only<5->{
+
+\draw[red,dotted,->,ultra thick] (sendright4) -- (queue 2);
+}
+
+\only<5>{\draw[ultra thick,->] (message3) -- (message4);}
+
+\end{tikzpicture}
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt netmsg}: Translation of a send right for transmission}
+\begin{semiverbatim}
+\tiny
+if ((ports[i] == MACH_PORT_NULL) || (ports[i] == MACH_PORT_DEAD))
+  \{
+    continue;
+  \}
+else if (remote_ports_by_local.count(ports[i]) == 1)
+  \{
+    /* We're transmitting a send right that we earlier received over the network.  Convert to the
+     * remote's name space, and indicate to the remote, by bit flipping the port number, that we're
+     * sending it a port number in its own name space.  Also, destroy the send right we received
+     * with the message, which is a send right to ourself!  If it's the port's last send right, then
+     * we'll get a no senders notification and deallocate the receive right then, so all we destroy
+     * here is the send right.
+     */
+
+    assert(local_port_type.at(ports[i]) == MACH_MSG_TYPE_PORT_RECEIVE);
+    mach_call (mach_port_mod_refs (mach_task_self(), ports[i],
+                                   MACH_PORT_RIGHT_SEND, -1));
+
+    ports[i] = (~ remote_ports_by_local[ports[i]]);
+  \}
+else
+  \{
+    /* We've got a send right (note it), but no mapping to a remote port (the remote takes care of that) */
+
+    local_port_type[ports[i]] = MACH_MSG_TYPE_PORT_SEND;
+
+    /* request a DEAD NAME notification */
+
+    mach_call (mach_port_request_notification (mach_task_self (), ports[i],
+                                               MACH_NOTIFY_DEAD_NAME, 0,
+                                               notification_port, MACH_MSG_TYPE_MAKE_SEND_ONCE, &old));
+  \}
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt netmsg}: Translation of a send right on reception}
+\begin{semiverbatim}
+\tiny
+/* We're receiving a send right over the network.  We need to relay on a send
+ * right, which will be to a local receive right.
+ */
+
+if (local_ports_by_remote.count(port) == 1)
+  \{
+    const mach_port_t newport = local_ports_by_remote[port];
+
+    assert(local_port_type.at(newport) == MACH_MSG_TYPE_PORT_RECEIVE);
+
+    /* it already exists; create a new send right to relay on */
+    mach_call (mach_port_insert_right (mach_task_self (), newport, newport,
+                                       MACH_MSG_TYPE_MAKE_SEND));
+
+    return newport;
+  \}
+else
+  \{
+    mach_port_t newport;
+
+    /* create new receive port and a new send right that will be moved to the recipient */
+    mach_call (mach_port_allocate (mach_task_self (), MACH_PORT_RIGHT_RECEIVE, &newport));
+    mach_call (mach_port_insert_right (mach_task_self (), newport, newport,
+                                       MACH_MSG_TYPE_MAKE_SEND));
+
+    /* request notification when all send rights have been destroyed */
+    mach_port_t old;
+    mach_call (mach_port_request_notification (mach_task_self (), newport,
+                                               MACH_NOTIFY_NO_SENDERS, 0,
+                                               newport, MACH_MSG_TYPE_MAKE_SEND_ONCE, &old));
+    return newport;
+  \}
+\end{semiverbatim}
 \end{frame}
@@
 
+\begin{comment}
 \begin{frame}
@@
 \end{frame}
+\end{comment}
+
+\begin{frame}
+\frametitle{Difficulty tracking which clients have access}
+
+% Because of the kernel's ability to proactively flush pages (unless PRECIOUS is specified),
+% we can't be certain that a client on a page's ACCESSLIST actually possesses that page.
+% All we know for sure is that a client not on ACCESSLIST does not possess the page.
+
+% Examining the Mach kernel code for m_o_lock_request, we see that the kernel might block
+% while looping over the pages (and unlocks the memory object while doing so), so for a
+% multi-page lock request, we need to consider the possibility that part of the page range
+% might be processed, then other kernel operations happen (including VM page operations)
+% before the remaining part of the page range is processed and the lock request completes.
+
+\begin{itemize}
+\item five clients have read access to page 50 and are on its ACCESSLIST
+\item pager\_flush is called on pages 0-100
+\item page 50 flushes on client A
+\item client A re-requests access to page 50, and it is supplied (ACCESSLIST doesn't change)
+\item the flushes complete
+\item only client A now has access to page 50 and should be alone on its ACCESSLIST
+\end{itemize}
+
+% This scenario shows that ACCESSLIST, WAITLIST and a FLUSHING flag aren't enough; we have to track
+% which clients requested access during the flush to correctly compute ACCESSLIST at the end.
+
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt \_\_pthread\_cond\_timedwait\_internal}}
+\begin{semiverbatim}
+\tiny
+  /* Block the thread.  */
+  err = __pthread_timedblock (self, abstime, clock_id);
+
+  __pthread_spin_lock (&cond->__lock);
+  if (! self->prevp)
+    \{
+      /* Another thread removed us from the list of waiters, which means a
+         wakeup message has been sent.  It was either consumed while we were
+         blocking, or queued after we timed out and before we acquired the
+         condition lock, in which case the message queue must be drained.  */
+      if (! err)
+        drain = 0;
+      else
+        \{
+          assert (err == ETIMEDOUT);
+          drain = 1;
+        \}
+    \}
+  else
+    \{
+      /* We're still in the list of waiters.  Noone attempted to wake us up,
+         i.e. we timed out.  */
+      assert (err == ETIMEDOUT);
+      __pthread_dequeue (self);
+      drain = 0;
+    \}
+  __pthread_spin_unlock (&cond->__lock);
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Original {\tt get\_identity} code}
+\begin{semiverbatim}
+\tiny
+id_clean (void *cookie)
+\{
+  struct idspec *i = cookie;
+  pthread_mutex_lock (&idlock);
+  if (refcounts_hard_references(&i->pi.refcounts) == 0) \{
+      /* Nobody got a send right in between, we can remove from the hash.  */
+      hurd_ihash_locp_remove (&idhash, i->id_hashloc);
+      ports_port_deref_weak (&i->pi);
+  \}
+  pthread_mutex_unlock (&idlock);
+\}
+
+fshelp_get_identity (struct port_bucket *bucket, ino_t fileno, mach_port_t *pt)
+\{
+  pthread_mutex_lock (&idlock);
+
+  i = hurd_ihash_find (&idhash, (hurd_ihash_key_t) fileno);
+  if (i == NULL) \{
+      ports_create_port (idclass, bucket, sizeof (struct idspec), &i);
+      hurd_ihash_add (&idhash, (hurd_ihash_key_t) fileno, i);
+
+      /* Weak reference for the hash entry.  */
+      ports_port_ref_weak(&i->pi);
+
+      *pt = ports_get_right (i);
+      ports_port_deref (i);
+  \} else \{
+    *pt = ports_get_right (i);
+  \}
+
+  pthread_mutex_unlock (&idlock);
+\}
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{{\tt libports} dereference code}
+\begin{semiverbatim}
+\tiny
+void
+ports_port_deref (void *portstruct)
+\{
+  struct port_info *pi = portstruct;
+  struct references result;
+
+  if (pi->class->dropweak_routine)
+    \{
+      /* If we need to call the dropweak routine, we need to hold one
+         reference while doing so.  We use a weak reference for this
+         purpose, which we acquire by demoting our hard reference to a
+         weak one.  */
+      refcounts_demote (&pi->refcounts, &result);
+
+      if (result.hard == 0 && result.weak > 1)
+        (*pi->class->dropweak_routine) (pi);
+
+      refcounts_deref_weak (&pi->refcounts, &result);
+    \}
+  else
+    refcounts_deref (&pi->refcounts, &result);
+
+  if (result.hard == 0 && result.weak == 0)
+    _ports_complete_deallocate (pi);
+\}
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{First attempt to fix {\tt get\_identity} code}
+\begin{semiverbatim}
+\tiny
+id_drop_weak (void *cookie)
+\{
+  struct idspec *i = cookie;
+  pthread_mutex_lock (&idlock);
+  if (refcounts_hard_references(&i->pi.refcounts) == 0) \{
+      /* Nobody got a send right in between, we can remove from the hash.  */
+      hurd_ihash_locp_remove (&idhash, i->id_hashloc);
+      ports_port_deref_weak (&i->pi);
+  \}
+  pthread_mutex_unlock (&idlock);
+\}
+
+fshelp_get_identity (struct port_bucket *bucket, ino_t fileno, mach_port_t *pt)
+\{
+  pthread_mutex_lock (&idlock);
+
+  i = hurd_ihash_find (&idhash, (hurd_ihash_key_t) fileno);
+  if (i == NULL) \{
+      ports_create_port (idclass, bucket, sizeof (struct idspec), &i);
+      hurd_ihash_add (&idhash, (hurd_ihash_key_t) fileno, i);
+
+      /* Weak reference for the hash entry.  */
+      ports_port_ref_weak(&i->pi);
+
+      *pt = ports_get_right (i);
+      ports_port_deref (i);
+  \} else \{
+    *pt = ports_get_right (i);
+  \}
+
+  pthread_mutex_unlock (&idlock);
+\}
+
+\end{semiverbatim}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Correctly fixed {\tt get\_identity} code}
+\begin{semiverbatim}
+\tiny
+id_drop_weak (void *cookie)
+\{
+  struct idspec *i = cookie;
+  pthread_mutex_lock (&idlock);
+  if (refcounts_hard_references(&i->pi.refcounts) == 0 && i->id_hashloc != NULL) \{
+      /* Nobody got a send right in between, we can remove i from the hash.  */
+      hurd_ihash_locp_remove (&idhash, i->id_hashloc);
+      i->id_hashloc = NULL;
+      ports_port_deref_weak (&i->pi);
+  \}
+  pthread_mutex_unlock (&idlock);
+\}
+
+fshelp_get_identity (struct port_bucket *bucket, ino_t fileno, mach_port_t *pt)
+\{
+  pthread_mutex_lock (&idlock);
+
+  i = hurd_ihash_find (&idhash, (hurd_ihash_key_t) fileno);
+  if (i == NULL) \{
+      ports_create_port (idclass, bucket, sizeof (struct idspec), &i);
+      hurd_ihash_add (&idhash, (hurd_ihash_key_t) fileno, i);
+
+      /* Weak reference for the hash entry.  */
+      ports_port_ref_weak(&i->pi);
+
+      *pt = ports_get_right (i);
+      ports_port_deref (i);
+  \} else \{
+    *pt = ports_get_right (i);
+  \}
+
+  pthread_mutex_unlock (&idlock);
+\}
+
+\end{semiverbatim}
+\end{frame}
 
@@
 \begin{itemize}
+\item Finish multi-client {\tt libpager}
 \item 64-bit user space
@@
 \item Rewrite of {\tt netmsg}
+\begin{itemize}
+\item UDP or raw Ethernet frames instead of TCP
+\item OSPF-like protocol to manage port numbers over multiple hosts
+\item Synchronization of port closures (DEAD NAMES and NO SENDERS)
+\item raw PCI device access
+\end{itemize}
+\vskip 0.5in
+\item {\tt https://www.gnu.org/software/hurd/}
+\item {\tt https://github.com/BrentBaccala/hurd}
 \end{itemize}

### 2018-01-21 slides/cua.tex
rename 'papers' subdirectory -> 'slides'
slides - move tikz macros to a separate file

--- 2018-01-17 papers/cua.tex
@@
+% Slideshow, written by Brent Baccala, for a lecture at Catholic University
+%
+% A screencast of the original lecture is available on youtube:
+%
+% https://www.youtube.com/watch?v=JwsuAEF2FYE
+
 \documentclass{beamer}
@@
 
-\usepackage{xstring}
+\usepackage{hurd}
 
 \usepackage{tikz}
-\usetikzlibrary{positioning, fit, backgrounds, arrows, patterns, trees, shapes}
+\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes}
 
@@
 \frametitle{Processor families in TOP500}
+% https://upload.wikimedia.org/wikipedia/commons/thumb/e/ef/Processor_families_in_TOP500_supercomputers.svg/800px-Processor_families_in_TOP500_supercomputers.svg.png
 \includegraphics[width=\textwidth]{Processor_families_in_TOP500_supercomputers.png}
@@
 \frametitle{Intel Ivy Bridge Processor}
+% https://www.pcper.com/files/review/2012-04-22/slides03.jpg
 \includegraphics[width=\textwidth]{ivybridge.jpg}
@@
 \frametitle{Cisco 5108 Blade Server Chassis with eight B200 M2s}
+% https://drive.google.com/uc?export=download&id=1UVc0DbcYrSk5DNcCGneIU_RmkW83JQFA
 \includegraphics[width=\textwidth]{ourcisco.jpg}
@@
 \frametitle{How many Debian packages build on Hurd?}
+% https://buildd.debian.org/stats/graph-big.png
 \includegraphics[width=\textwidth]{graph-big.png}
 \end{frame}
-
-% tikz commands to draw task boxes to illustrate Mach IPC concepts
-
-\tikzstyle{task} = [draw, fill=blue!20, rounded corners]
-
-\tikzstyle{kernel} = [draw, fill=blue!20, rounded corners, minimum width=30em, minimum height=8em]
-
-\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=3em, text centered,
-  minimum width=5em, minimum height=2em, rounded corners]
-
-% \task - takes four arguments
-% 1. name of task (used to refer to it from tikz; maybe printed as a label as well)
-% 2. location of task box's center point
-% 3. "r" or "l", indicating if a set of tick marks representing
-%    ports should drawn on the left or right side of the task box
-%    "u" to label the task "Task" instead of labeling it with its name
-%    "b" to make a "big" node (I wish I knew how to size the tick marks automatically)
-%    "s" to make a "small" node
-% 4. contents of the task box - other nodes
-%
-% The tick marks will be named "#1 Port #" for future reference
-%
-% It's implemented as a tikz matrix with two cells - the task name in the top cell
-% and everything else in the bottom cell
-
-\newcommand\task[4]{
-\matrix (#1) [task, inner xsep=0] at (#2) { \node {\IfSubStr{#3}{u}{Task}{#1}}; \\ #4
-
-% print port tick marks in fill color if they are disabled, in order to make
-% them expand out the dimensions of the task box yet remain invisible
-
-\IfSubStr{#3}{b}{\edef\starti{-20}}{\IfSubStr{#3}{s}{\edef\starti{-1}}{\edef\starti{-7}}}
-\IfSubStr{#3}{s}{\edef\endi{1}}{\edef\endi{7}}
-
-\foreach \i in {\starti,...,\endi}
-{
-  \IfSubStr{#3}{l}{\edef\leftcolor{black}\edef\leftcoord{coordinate (#1 Port \i)}}
-                  {\edef\leftcolor{blue!20}\edef\leftcoord{}}
-  \IfSubStr{#3}{r}{\edef\rightcolor{black}\edef\rightcoord{coordinate (#1 Port \i)}}
-                  {\edef\rightcolor{blue!20}\edef\rightcoord{}}
-  \draw[xshift=-5em, ultra thick, color=\leftcolor] (0,0.2 * \i) \leftcoord -- (0.2,0.2 * \i);
-  \draw[xshift=5em, ultra thick, color=\rightcolor] (-0.2,0.2 * \i) -- (0,0.2 * \i) \rightcoord;
-}
-
-\\};
-}
-
-% draw a box with vertical lines to represent a message queue
-% arg 1 is the name of a port to position it to the left of
-% arg 2 is name to be given the box
-
-\newcommand\queue[2]{
-\filldraw (#1) +(-3.1,0) node[rectangle, name=#2, minimum width=1.8cm, minimum height=1cm, draw, pattern=vertical lines] {};
-}
 
@@
 \end{frame}
+
+\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=3em, text centered,
+  minimum width=5em, minimum height=2em, rounded corners]
 
@@
 
-%\begin{frame}
-%\frametitle{Example: Basic File I/O}
-%\end{frame}
-
 \begin{frame}[fragile]
@@
 \end{frame}
+
+\tikzstyle{kernel} = [draw, fill=blue!20, rounded corners, minimum width=30em, minimum height=8em]
 
@@
 \filldraw (-30 em, 15em) node[rectangle, name=queue 1, minimum width=1.8cm, minimum height=1cm, draw, pattern=vertical lines] {};
-% \task{Task 0}{-30 em,15 em}{ur}{}
 

### 2018-01-21 slides/cua.tex
slides - change calling convention for \task and \queue macros

@@
 
-\begin{tikzpicture}[transform canvas={scale=0.4, yshift=12em, xshift=-15em}]
-
-\task{Task 1 left}{0,0}{ru}{}
-\task{Task 2 left}{0,-11 em}{ru}{}
-\task{Task 3 left}{0,-22 em}{ru}{}
-\task{Task 4 left}{0,-33 em}{ru}{}
-\task{Task 1 right}{30 em,0}{lu}{}
-\task{Task 2 right}{30 em,-11 em}{lu}{}
-\task{Task 3 right}{30 em,-22 em}{lu}{}
-\task{Task 4 right}{30 em,-33 em}{lu}{}
-
-\queue{Task 1 right Port 5}{queue 1}
-\queue{Task 2 right Port -5}{queue 2a}
-\queue{Task 2 right Port 5}{queue 2b}
-\queue{Task 3 right Port 0}{queue 3}
-\queue{Task 4 right Port 0}{queue 4}
+% node distance is the distance that the queues are positioned from their task ports
+\begin{tikzpicture}[transform canvas={scale=0.4, yshift=12em, xshift=-15em}, node distance=6em]
+
+\task{Task 1 left}{at (0,0)}{ru}{}
+\task{Task 2 left}{at (0,-11em)}{ru}{}
+\task{Task 3 left}{at (0,-22em)}{ru}{}
+\task{Task 4 left}{at (0,-33 em)}{ru}{}
+\task{Task 1 right}{at (30 em,0)}{lu}{}
+\task{Task 2 right}{at (30 em,-11 em)}{lu}{}
+\task{Task 3 right}{at (30 em,-22 em)}{lu}{}
+\task{Task 4 right}{at (30 em,-33 em)}{lu}{}
+
+\queue[left = of Task 1 right Port 5]{queue 1}{}
+\queue[left = of Task 2 right Port -5]{queue 2a}{}
+\queue[left = of Task 2 right Port 5]{queue 2b}{}
+\queue[left = of Task 3 right Port 0]{queue 3}{}
+\queue[left = of Task 4 right Port 0]{queue 4}{}
 
@@
 
-\task{Task 1}{0,0}{r}{
+\task{Task 1}{at (0,0)}{r}{
 \node (message1) [message] {Message 1};
 }
-\task{Task 2}{15 em,0}{l}{
+\task{Task 2}{at (15 em,0)}{l}{
 \node (message2) [message] at (1 em, 1 em) {Message 1};
@@
 }
-\task{Task 3}{0,-11 em}{r}{
+\task{Task 3}{at (0,-11 em)}{r}{
 \node (message3) [message] {Message 2};
@@
 
-\task{Task 1}{0,0}{br}{
+\task{Task 1}{at (0,0)}{br}{
 % print two sendright nodes, one to fill the message1 box to the correct size,
@@
 
-\task{Task 2}{15 em,0}{bl}{
+\task{Task 2}{at (15 em,0)}{bl}{
 \only<3->{
@@
 
-\task{Task 2}{15 em,0}{bl}{
+\task{Task 2}{at (15 em,0)}{bl}{
 \only<2->{
@@
 
-\task{ext2 File Server}{20 em,-5 em}{l}{}
-\task{httpfs}{20 em, -12 em}{ls}{}
-\task{pfinet}{20 em, -17 em}{ls}{}
+\task{ext2 File Server}{at (20 em,-5 em)}{l}{}
+\task{httpfs}{at (20 em, -12 em)}{ls}{}
+\task{pfinet}{at (20 em, -17 em)}{ls}{}
 
@@
 
-\task{Filesystem Server}{0,0}{}{
+\task{Filesystem Server}{at (0,0)}{}{
 \node (memory1) [above left=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
 }
-\task{ls}{15 em,0}{}{
+\task{ls}{at (15 em,0)}{}{
 %\node (memory1) [above right=-0.5, fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
@@
 
-\task{Task 1}{-5em,0}{}{
+\task{Task 1}{at (-5em,0)}{}{
 \node (memory1) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
 }
-\task{Task 2}{6 em,0}{}{
+\task{Task 2}{at (6 em,0)}{}{
 \node (memory2) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
 }
-\task{Task 3}{17 em,0}{}{
+\task{Task 3}{at (17 em,0)}{}{
 \node (memory3) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
@@
 
-\task{Task 1}{-5em,0}{u}{
+\task{Task 1}{at (-5em,0)}{u}{
 \only<1-2,9->{\node (memory1) [emptymemory] {\only<2>{\Huge R}};}
@@
 }
-\task{Task 2}{6 em,0}{u}{
+\task{Task 2}{at (6 em,0)}{u}{
 \only<1-3,9->{\node (memory2) [emptymemory] {};}
@@
 }
-\task{Task 3}{17 em,0}{u}{
+\task{Task 3}{at (17 em,0)}{u}{
 \only<1-3,9-13>{\node (memory3) [emptymemory] {\only<11->{\Huge W}};}
@@
 
-\task{Task 4}{-5em,-40 em}{u}{
+\task{Task 4}{at (-5em,-40 em)}{u}{
 \node (memory4) [emptymemory] {};
 }
-\task{Task 5}{6 em,-40 em}{u}{
+\task{Task 5}{at (6 em,-40 em)}{u}{
 \only<1-5,13->{\node (memory5) [emptymemory] {\only<5>{\Huge R}};}
@@
 }
-\task{Task 6}{17 em,-40 em}{u}{
+\task{Task 6}{at (17 em,-40 em)}{u}{
 \only<1-9,13->{\node (memory6) [emptymemory] {\only<7-9>{\Huge W}};}
@@
 
-\task{Task 1}{-5em,0}{u}{}
-\task{Task 2}{6 em,0}{u}{}
-\task{Task 3}{17 em,0}{u}{}
+\task{Task 1}{at (-5em,0)}{u}{}
+\task{Task 2}{at (6 em,0)}{u}{}
+\task{Task 3}{at (17 em,0)}{u}{}
 
@@
 
-\task{Task 4}{-5em,-40 em}{u}{}
-\task{Task 5}{6 em,-40 em}{u}{}
-\task{Task 6}{17 em,-40 em}{u}{}
+\task{Task 4}{at (-5em,-40 em)}{u}{}
+\task{Task 5}{at (6 em,-40 em)}{u}{}
+\task{Task 6}{at (17 em,-40 em)}{u}{}
 
@@
 
-\task{Task 1}{-30 em,0}{r}{
+\task{Task 1}{at (-30 em,0)}{r}{
   \only<1-2>{
@@
 }
-\task{netmsg 1}{-15 em,0}{l}{
+\task{netmsg 1}{at (-15 em,0)}{l}{
   \only<2-3>{
@@
 
-\task{netmsg 2}{15 em,0}{r}{
+\task{netmsg 2}{at (15 em,0)}{r}{
   \only<3-5>{
@@
 }
-\task{Task 2}{30 em,0}{l}{
+\task{Task 2}{at (30 em,0)}{l}{
   \only<5>{

### 2018-01-22 slides/rpctrace.tex
add slides describing rpctrace "attach" operation

\documentclass{beamer}
\usetheme{Madrid}

\setbeamertemplate{footline}{}
\beamertemplatenavigationsymbolsempty

\usepackage{comment}
\usepackage{hurd}

\usetikzlibrary{positioning, calc}

\begin{document}

\tikzstyle{send} = [rectangle, fill=green, rounded corners] 
\tikzstyle{recv} = [rectangle, fill=red!50, rounded corners]
\tikzstyle{empty} = [minimum size=0, inner sep=0]
\tikzstyle{queue} = [draw, rectangle, minimum height=.5cm, minimum width=1cm, pattern=vertical lines]

\begin{frame}
\begin{tikzpicture}

\node (center point) {};

\task [left = of center point] {rpctrace}{}{lr}{}
\task [right = of center point] {PUT}{}{l}{}

\queue [above = of rpctrace] {queue}{}

\node (below rpctrace) [below = of rpctrace] {};
\node (below PUT) [below = of PUT] {};

% invisible at first, but present to keep the canvas sizing consistent across all slides
\only<1-4>{
  \queue [white] {queue2}{at ($(below rpctrace)!.5!(below PUT)$)}
}
\only<5->{
  \queue {queue2}{at ($(below rpctrace)!.5!(below PUT)$)}
}

% invisible receive right to match one from the next set of slides, so the canvas sizing is identical
\draw [white, thick] (queue.west) -- ++(-3,1) node [recv, white, anchor=south] {R};

\only<1,8->{\node (R1) at (PUT Port 0) [recv, anchor=west] {R};}
\only<2-7>{\node (R1) at (PUT Port 0) [recv, anchor=west, minimum size=0, inner sep=0] {};}
\node (r1) at (R1) {};

\only<1-3,7->{
  \node (S1) at (R1.east) [send, anchor=west] {S};
  \node (S2) at (S1.east) [send, anchor=west] {S};
}

\only<1-3>{
  \draw [violet, thick] (PUT Port 0) -- +(-1,0) |- (queue.east);
}

\draw [violet, thick] (queue.north) -- ++(-1,1) node [send, anchor=south] {S};
\draw [violet, thick] (queue.north) -- ++(1,1) node [send, anchor=south] {S};

\only<2->{
  \node (R2) at (rpctrace Left Port 0) [recv, anchor=west] {R};
  \draw [violet, thick] (queue.west) -- ++(-2,0) |- (R2);
}

\only<3-9>{
  \node (S3) at (R2.east) [send, anchor=west] {S};
}
\only<10->{
  \node (S3) at (R2.east) [empty] {};
}

\only<5-7>{
  \node (R3) at (rpctrace Right Port 0) [recv, anchor=east] {R};
}
\only<8->{
  \node (R3) at (rpctrace Right Port 0) [minimum size=0, inner sep=0] {};
}

\only<5->{
  \draw [violet, thick] (rpctrace Right Port 0) -| ($(rpctrace Right Port 0)!.5!(queue2.west)$) |- (queue2.west);
}

\only<6->{
  \node (S4) at (R3.west) [send, anchor=east] {S};
}

\only<7->{
  \draw [violet, thick] (queue2.east) -| ($(queue2.east)!.5!(PUT Port 0)$) |- (PUT Port 0);
}

\only<9->{
  % S3's location moves between slides, so starting this line at S4 keeps the dashes from morphing between slides
  \draw [violet, dashed, thick] (S4) -- (S3);
}

\end{tikzpicture}
\end{frame}

\begin{comment}
    (defun increment-number-at-point ()
      (interactive)
      (skip-chars-backward ``0-9'')
      (or (looking-at ``[0-9]+'')
          (error ``No number at point''))
      (replace-match (number-to-string (1+ (string-to-number (match-string 0))))))

(when (string-match ``[ \t]*$'' test-str)
    (message (concat ``[`` (replace-match " nil nil test-str) ``]'')))

     (while (re-search-forward ``foo[ \t]+bar'' nil t)
       (replace-match ``foobar''))

(defun ff ()
  "increment numbers in string"
  (let (mystr start)
    (setq str (match-string 0)
    (setq start 0)
    (while (string-match "[0-9]+" mystr start)
      (replace-match (number-to-string (1+ (string-to-number (match-string 0)))) nil nil (match-string 0))
  (replace-regexp-in-string ``_'' `` `` (match-string 1))
)

\end{comment}

\begin{frame}
\begin{tikzpicture}

\node (center point) {};

\task [left = of center point] {rpctrace}{}{lr}{}
\task [right = of center point] {PUT}{}{l}{}

\queue [above = of rpctrace] {queue}{}

\node (below rpctrace) [below = of rpctrace] {};
\node (below PUT) [below = of PUT] {};

% invisible at first, but present to keep the canvas sizing consistent across all slides
\only<1-3>{
  \queue [white] {queue2}{at ($(below rpctrace)!.5!(below PUT)$)}
}
\only<4->{
  \queue {queue2}{at ($(below rpctrace)!.5!(below PUT)$)}
}

\only<1-2,5->{
  \node (S1) at (PUT Port 0) [send, anchor=west] {S};
  \node (S2) at (S1.east) [send, anchor=west] {S};
}

\only<1-2>{
  \draw [violet, thick] (PUT Port 0) -- +(-1,0) |- (queue.east);
}

\draw [violet, thick] (queue.north) -- ++(-1,1) node [send, anchor=south] {S};
\draw [violet, thick] (queue.north) -- ++(1,1) node [send, anchor=south] {S};
\draw [violet, thick] (queue.west) -- ++(-3,1) node [recv, anchor=south] {R};

\only<2->{
  \node (S3) at (rpctrace Left Port 0) [send, anchor=west] {S};
  \draw [violet, thick] (queue.west) -- ++(-2,0) |- (S3);
}

\only<4->{
  \node (R3) at (rpctrace Right Port 0) [recv, anchor=east] {R};
}

\only<4->{
  \draw [violet, thick] (rpctrace Right Port 0) -| ($(rpctrace Right Port 0)!.5!(queue2.west)$) |- (queue2.west);
}

\only<5->{
  \draw [violet, thick] (queue2.east) -| ($(queue2.east)!.5!(PUT Port 0)$) |- (PUT Port 0);
}

\only<6->{
  \draw [violet, dashed, thick] (S3) -- (R3);
}

\end{tikzpicture}
\end{frame}

\end{document}

### 2018-01-26 slides/rpctrace.tex
rpctrace slide series - add slide series illustrating insert_right problem

@@
 
-\usetikzlibrary{positioning, calc}
+\usetikzlibrary{positioning, calc, arrows, fit}
 
@@
 \tikzstyle{queue} = [draw, rectangle, minimum height=.5cm, minimum width=1cm, pattern=vertical lines]
+\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=3em, text centered,
+                       minimum width=5em, minimum height=2em, rounded corners]
+
+\tikzstyle{port link} = [black, -triangle 90]
+\tikzstyle{message path} = [violet, thick, -triangle 90]
+\tikzstyle{right link} = [red, thick, dotted, ->]
+\tikzstyle{internal link} = [dashed, thick]
+
+% illustration of rpctrace problems with mach_port_insert_right()
+
+\begin{frame}[fragile]
+\begin{block}{}
+\begin{semiverbatim}
+\tiny
+ports\_create\_port (port\_class, port\_bucket, sizeof (struct port\_info), \&readport);
+readportname = ports\_get\_right (readport);
+mach\_port\_insert\_right (mach\_task\_self (), readportname, readportname, MACH\_MSG\_TYPE\_MAKE\_SEND);
+\end{semiverbatim}
+\end{block}
+
+\begin{block}{}
+\begin{semiverbatim}
+\tiny
+routine mach\_port\_insert\_right(
+                task            : ipc\_space\_t;
+                name            : mach\_port\_name\_t;
+                poly            : mach\_port\_poly\_t);
+
+\end{semiverbatim}
+\end{block}
+
+\vfill
+
+\begin{center}
+\begin{tikzpicture}
+
+\node (center point) {};
+
+\task [left = of center point] {rpctrace}{}{lr}{
+  \only<1-3>{
+    \node (rpctrace Message) [message, anchor=north, opacity=0] {\scriptsize insert\_right};
+  }
+  \only<4-5>{
+    \node (rpctrace Message) [message, anchor=north] {\scriptsize insert\_right};
+  }
+  \only<6>{
+    \node (rpctrace Message) [message, anchor=north, fill=none, dashed, text opacity=0] {\scriptsize insert\_right};
+  }
+}
+
+\only<4-5>{
+  \node (rpctrace sendright) at (rpctrace Message) [send, anchor=west] {S};
+  \node at (rpctrace sendright.west) [anchor=east, outer sep=0, inner sep=0] {(10,};
+  \node at (rpctrace sendright.east) [anchor=west, outer sep=0, inner sep=0] {)};
+}
+
+\task [right = of center point] {Traced Task}{}{lk}{
+  \only<1-2,5->{
+    \node (Traced Task Message) [message, anchor=north, opacity=0] {\scriptsize insert\_right};
+  }
+  \only<3>{
+    \node (Traced Task Message) [message, anchor=north] {\scriptsize insert\_right};
+  }
+  \only<4>{
+    \node (Traced Task Message) [message, anchor=north, fill=none, dashed, text opacity=0] {\scriptsize insert\_right};
+  }
+}
+
+\only<3>{
+  \node (Traced Task sendright) at (Traced Task Message) [send, anchor=west] {S};
+  \node at (Traced Task sendright.west) [anchor=east, outer sep=0, inner sep=0] {(10,};
+  \node at (Traced Task sendright.east) [anchor=west, outer sep=0, inner sep=0] {)};
+}
+
+\draw [port link] (rpctrace Port 7) -| ($(rpctrace Port 7)!.5!(Traced Task Task Port)$) |- (Traced Task Task Port);
+
+\node (S1) at (rpctrace Port 7) [send, anchor=east] {S};
+\node (R1) at (rpctrace Port 4) [recv, anchor=east] {R};
+
+\draw [internal link] (S1) -- +(-.5,0) |- (R1);
+
+\draw [port link] (Traced Task Port 4) -- (rpctrace Port 4);
+\node (S2) at (Traced Task Port 4) [send, anchor=west] {S};
+\node at (S2.east) [anchor=west] {\scriptsize mach\_task\_self()};
+
+\only<4>{
+  \draw [message path] (Traced Task Message.north) |- (S2);
+  \draw [message path] (Traced Task Port 4) -- (rpctrace Port 4);
+  \draw [message path] (R1) -| (rpctrace Message.north);
+}
+
+\only<6>{
+  \draw [message path] (rpctrace Message.north) |- (S1);
+  \draw [message path] (rpctrace Port 7) -| ($(rpctrace Port 7)!.5!(Traced Task Task Port)$) |- (Traced Task Task Port);
+  \draw [message path] (Traced Task Task Port) -| ($(Traced Task Task Port)!.5!(Kernel Message.west)$) |- (Kernel Message.west);
+}
+
+\only<2->{
+  \node (R2) at (Traced Task Port -4) [recv, anchor=west] {R};
+  \node at (R2.north) [anchor=south west, inner sep=2] {10};
+}
+
+\only<3>{
+  \draw [right link] (Traced Task sendright) -- (R2);
+}
+
+\only<4>{
+  \draw [right link] (rpctrace sendright) -- (R2);
+}
+
+\only<5->{
+  \node (S3) at (rpctrace Port -4) [send, anchor=east] {S};
+  \node (R3) at (rpctrace Port -1) [recv, anchor=east] {R};
+  \draw [internal link] (S3) -- +(-.5,0) |- (R3);
+  \draw [port link] (rpctrace Port -4) -- (Traced Task Port -4);
+}
+
+\only<5>{
+  \draw [right link] (rpctrace sendright) -- (R3);
+}
+
+\only<1-5>{
+  \node (Kernel Message) [message, anchor=south west, opacity=0] at (Traced Task.north) {\scriptsize insert\_right};
+}
+\only<6->{
+  \node (Kernel Message) [message, anchor=south west] at (Traced Task.north) {\scriptsize insert\_right};
+  \node (Kernel sendright) at (Kernel Message) [send, anchor=west] {S};
+  \node at (Kernel sendright.west) [anchor=east, outer sep=0, inner sep=0] {(10,};
+  \node at (Kernel sendright.east) [anchor=west, outer sep=0, inner sep=0] {)};
+  \draw [right link] (Kernel sendright) -- (R3);
+}
+
+\end{tikzpicture}
+\end{center}
+
+\end{frame}
+
+%% illustration of attach operation (two frames)
 
@@
 
-\only<1,8->{\node (R1) at (PUT Port 0) [recv, anchor=west] {R};}
-\only<2-7>{\node (R1) at (PUT Port 0) [recv, anchor=west, minimum size=0, inner sep=0] {};}
-\node (r1) at (R1) {};
+\only<1,5->{\node (R1) at (PUT Port 0) [recv, anchor=west] {R};}
+\only<2-4>{\node (R1) at (PUT Port 0) [empty] {};}
 
-\only<1-3,7->{
+\only<1-3,6->{
   \node (S1) at (R1.east) [send, anchor=west] {S};
@@
 
-\only<3-9>{
+\only<3-8>{
   \node (S3) at (R2.east) [send, anchor=west] {S};
 }
-\only<10->{
+\only<9->{
   \node (S3) at (R2.east) [empty] {};
@@
 
-\only<5-7>{
-  \node (R3) at (rpctrace Right Port 0) [recv, anchor=east] {R};
-}
-\only<8->{
-  \node (R3) at (rpctrace Right Port 0) [minimum size=0, inner sep=0] {};
+\only<7->{
+  \node (S4) at (rpctrace Right Port 0) [send, anchor=east] {S};
+  \draw [violet, thick] (rpctrace Right Port 0) -| ($(rpctrace Right Port 0)!.5!(queue2.west)$) |- (queue2.west);
 }
@@
 \only<5->{
-  \draw [violet, thick] (rpctrace Right Port 0) -| ($(rpctrace Right Port 0)!.5!(queue2.west)$) |- (queue2.west);
-}
-
-\only<6->{
-  \node (S4) at (R3.west) [send, anchor=east] {S};
-}
-
-\only<7->{
   \draw [violet, thick] (queue2.east) -| ($(queue2.east)!.5!(PUT Port 0)$) |- (PUT Port 0);
@@
 
-\only<9->{
+\only<8->{
   % S3's location moves between slides, so starting this line at S4 keeps the dashes from morphing between slides

### 2018-01-26 slides/rpctrace.tex
slides - update elisp code to modify \only macros (now runs, but is buggy)

@@
 \usetikzlibrary{positioning, calc, arrows, fit}
+
+\begin{comment}
+
+;; ELISP code to insert and remove slides from beamer \only macros
+;;
+;; XXX still have to deal with case where \only<2> needs to become \only<2-3> (inserting new frame 3)
+;;     or \only<5> has to become \only<0> (deleting frame 5)
+
+(defun incr-only-func (min)
+  "increment all numbers in string greater than or equal to a minimum value"
+  (let ((mystr (match-string 0)) (start 0) value)
+       (save-match-data (while (string-match "[0-9]+" mystr start)
+                               (setq value (string-to-number (match-string 0 mystr)))
+                               (when (>= value min)
+                                     (setq mystr (replace-match (number-to-string (1+ value))
+                                                                nil nil mystr)))
+                               (string-match "[0-9]+" mystr start)
+                               (setq start (match-end 0))))
+       mystr))
+
+(defun incr-only (min)
+  "increment an only macro"
+  (interactive "NSlide number to insert: ")
+  (query-replace-regexp "\\only<[^>]+>" `(replace-eval-replacement replace-quote (incr-only-func ,min))))
+
+(defun decr-only-func (min)
+  "decrement all numbers in string greater than a minimum value"
+  (let ((mystr (match-string 0)) (start 0) value)
+       (save-match-data (while (string-match "[0-9]+" mystr start)
+                               (setq value (string-to-number (match-string 0 mystr)))
+                               (when (> value min)
+                                     (setq mystr (replace-match (number-to-string (1- value))
+                                                                nil nil mystr)))
+                               (string-match "[0-9]+" mystr start)
+                               (setq start (match-end 0))))
+       mystr))
+
+(defun decr-only (min)
+  "decrement an only macro"
+  (interactive "NSlide number to remove: ")
+  (query-replace-regexp "\\only<[^>]+>" `(replace-eval-replacement replace-quote (decr-only-func ,min))))
+
+\end{comment}
 
@@
 
-\begin{comment}
-    (defun increment-number-at-point ()
-      (interactive)
-      (skip-chars-backward ``0-9'')
-      (or (looking-at ``[0-9]+'')
-          (error ``No number at point''))
-      (replace-match (number-to-string (1+ (string-to-number (match-string 0))))))
-
-(when (string-match ``[ \t]*$'' test-str)
-    (message (concat ``[`` (replace-match " nil nil test-str) ``]'')))
-
-     (while (re-search-forward ``foo[ \t]+bar'' nil t)
-       (replace-match ``foobar''))
-
-(defun ff ()
-  "increment numbers in string"
-  (let (mystr start)
-    (setq str (match-string 0)
-    (setq start 0)
-    (while (string-match "[0-9]+" mystr start)
-      (replace-match (number-to-string (1+ (string-to-number (match-string 0)))) nil nil (match-string 0))
-  (replace-regexp-in-string ``_'' `` `` (match-string 1))
-)
-
-\end{comment}
-
 \begin{frame}

### 2018-01-28 slides/rpctrace.tex
rpctrace slides - add slide illustrating basic operation of rpctrace

@@
 \tikzstyle{internal link} = [dashed, thick]
+
+\begin{frame}
+\begin{tikzpicture}
+
+\task {rpctrace}{}{lr}{}
+
+\task [right = of rpctrace] {Traced Task 2}{}{sl}{}
+\task [above = of Traced Task 2] {Traced Task 1}{}{sl}{}
+\task [below = of Traced Task 2] {Traced Task 3}{}{sl}{}
+
+\task [above = of rpctrace] {Unknown Task}{}{sl}{}
+
+\draw [port link] (rpctrace Left Port 0) -- +(-1,0) |- (Unknown Task Port 0);
+\draw [port link] (Traced Task 2 Port 0) -- (rpctrace Right Port 0);
+\draw [port link] (Traced Task 1 Port 0) -| ($(Traced Task 1 Port 0)!.5!(rpctrace Right Port 5)$) |- (rpctrace Right Port 5);
+\draw [port link] (Traced Task 3 Port 0) -| ($(Traced Task 3 Port 0)!.5!(rpctrace Right Port -5)$) |- (rpctrace Right Port -5);
+
+\node (Task 1 sendright) at (Traced Task 1 Port 0) [send, anchor=west] {S};
+\node (Task 2 sendright) at (Traced Task 2 Port 0) [send, anchor=west] {S};
+\node (Task 3 sendright) at (Traced Task 3 Port 0) [send, anchor=west] {S};
+
+\node (Unknown Task recvright) at (Unknown Task Port 0) [recv, anchor=west] {R};
+
+\node (R1) at (rpctrace Right Port 5) [recv, anchor=east] {R};
+\node (R2) at (rpctrace Right Port 0) [recv, anchor=east] {R};
+\node (R3) at (rpctrace Right Port -5) [recv, anchor=east] {R};
+
+\node (S1) at (rpctrace Left Port 0) [send, anchor=west] {S};
+
+\draw [internal link] (S1) -- (R2);
+\draw [internal link] ($(S1)!.5!(R2)$) -- (R1);
+\draw [internal link] ($(S1)!.5!(R2)$) -- (R3);
+
+\end{tikzpicture}
+\end{frame}
 

### 2018-01-29 slides/rpctrace.tex
slides - elisp code to modify \only macros appears to run properly

@@
 ;; ELISP code to insert and remove slides from beamer \only macros
-;;
-;; XXX still have to deal with case where \only<2> needs to become \only<2-3> (inserting new frame 3)
-;;     or \only<5> has to become \only<0> (deleting frame 5)
 
-(defun incr-only-func (min)
-  "increment all numbers in string greater than or equal to a minimum value"
-  (let ((mystr (match-string 0)) (start 0) value)
-       (save-match-data (while (string-match "[0-9]+" mystr start)
-                               (setq value (string-to-number (match-string 0 mystr)))
-                               (when (>= value min)
-                                     (setq mystr (replace-match (number-to-string (1+ value))
-                                                                nil nil mystr)))
-                               (string-match "[0-9]+" mystr start)
-                               (setq start (match-end 0))))
+(defun slide-func (slide)
+  "add or remove a slide number from only macros"
+  (let ((mystr (match-string 0)) (start 0) value1 value2 newstr (sign (/ slide (abs slide))))
+       (save-match-data (while (string-match "\\([0-9]+\\)\\(-\\([0-9]+\\)?\\)?" mystr start)
+                               (setq value1 (string-to-number (match-string 1 mystr)))
+                               (setq value2 (cond ((match-string 3 mystr) (string-to-number (match-string 3 mystr)))
+                                                  ((match-string 2 mystr) 1000000)
+                                                  (t value1)))
+                               (if (> value1 (abs slide)) (setq value1 (+ value1 sign)))
+                               (if (>= value2 (abs slide)) (setq value2 (+ value2 sign)))
+                               (setq newstr (cond ((> value1 value2) "0")
+						  ((eq value1 value2) (format "%d" value1))
+                                                  ((>= value2 999999) (format "%d-" value1))
+                                                  (t (format "%d-%d" value1 value2))))
+                               (setq mystr (replace-match newstr nil nil mystr))
+                               (setq start (+ (match-beginning 0) (length newstr)))))
        mystr))
 
-(defun incr-only (min)
-  "increment an only macro"
-  (interactive "NSlide number to insert: ")
-  (query-replace-regexp "\\only<[^>]+>" `(replace-eval-replacement replace-quote (incr-only-func ,min))))
+(defun slide-duplicate (slide)
+  "remove slide numbers in only macros"
+  (interactive "NSlide number to duplicate: ")
+  (query-replace-regexp "\\only<[0-9-,]+>" `(replace-eval-replacement replace-quote (slide-func ,slide))))
 
-(defun decr-only-func (min)
-  "decrement all numbers in string greater than a minimum value"
-  (let ((mystr (match-string 0)) (start 0) value)
-       (save-match-data (while (string-match "[0-9]+" mystr start)
-                               (setq value (string-to-number (match-string 0 mystr)))
-                               (when (> value min)
-                                     (setq mystr (replace-match (number-to-string (1- value))
-                                                                nil nil mystr)))
-                               (string-match "[0-9]+" mystr start)
-                               (setq start (match-end 0))))
-       mystr))
-
-(defun decr-only (min)
-  "decrement an only macro"
+(defun slide-remove (slide)
+  "remove slide numbers from only macros"
   (interactive "NSlide number to remove: ")
-  (query-replace-regexp "\\only<[^>]+>" `(replace-eval-replacement replace-quote (decr-only-func ,min))))
+  (query-replace-regexp "\\only<[0-9-,]+>" `(replace-eval-replacement replace-quote (slide-func ,(- slide)))))
 

### 2019-05-08 libpager/paper.tex
paper.tex now compiles as a LaTeX file

\documentclass{article}

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}

% I like smaller margins than the default
\setmargnohfrb{0.5in}{0.5in}{0.5in}{0.5in}

% I prefer unindented paragraphs with a blank line between them
\parindent 0pt
\parskip \baselineskip

% Redefine underscore to print as a normal character
\catcode`\_=\active
\def_{\_}

\title{Building an SSI Cluter with GNU/Hurd}
\author{Brent Baccala}
\date{}

\pdfinfo{
  /Title	(Building an SSI Cluter with GNU/Hurd)
  /Author	(Brent Baccala)
}

% Shortcuts just for this document
\def\libpager{{\tt libpager\ }}

\begin{document}

\maketitle

\section{Introduction}

Operating system designers have tried for years to leverage
the power of multiple computers and develop cluster
operating systems.  Notable attempts include VAX/VMS, various cluster
Linux systems [MOSIX, refs], Beowulf, Plan 9, ScaleMP, and
Hadoop.

Some of these designs, like Beowulf and Hadoop, are built using
libraries that allow programs which use those libraries to run
distributed over multiple, otherwise independent, computers.  Others,
such as VAX/VMS, Plan 9, and ScaleMP, aspire to present the cluster as
if it were a single machine -- a Single System Image (SSI) -- allowing
unmodified programs to exploit cluster parallelism.

GNU/Hurd is an operating system based on the Mach kernel, which was
designed for use in a cluster environment, making it a natural choice
to build a POSIX-compatible SSI cluster.  Development on Hurd has
progressed slowly, but a Debian port has been available since 2013,
and approximately three quarters of the Debian packages compile and
run on Hurd.

To support a multi-node Hurd cluster, the libpager used by filesystem
translators needs to support multiple clients (i.e, multiple kernels
mmap'ing files).  Hurd's original libpager, written in the mid-1990s,
only supported a single client, and was thus suitable for single node
operation.

This is a crucial library.  It provides the cache coherence logic
necessary to implement inter-node shared memory, whether disk-based or
not.  Its use in the tmpfs translator, which is in turn used by
Hurd's POSIX shared memory implementation

Hurd's netmsg and libpaper have been developed to the point where Hurd
could be usable as a cluster operating system, though the current
Debian distribution doesn't use this functionality.  At the time of
this writing (2019), the major barriers to building a Hurd cluster are
the 32-bit address space and the lack of SMP support, enhancements
that are beyond the scope of the work described in this paper.

CAP THEOREM

The CAP theorem states that we can't have all three of: consistency,
availability, partitioning.  Right now, we sacrifice availability - if
the network partitions, libpager can deadlock.  If we introduce a
timeout on lock requests, then we're sacrificing consistency.

\section{Introduction to Mach Port-Based IPC}

In addition to opaque data (generally, but not necessarily interpreted
as remote procedure calls), Mach messages can contain two more
sophisticated structures.  The first are port rights.

A Mach message can transport send or receive rights for Mach ports.
This is basically the only way in which a process can obtain Mach
ports since, much like POSIX file descriptors, port numbers are local
to a process and interpreted by the kernel, so there is no way that a
process can fabricate another process's port number, just as a POSIX
process can not fabricate another process's file descriptor.

All Hurd authentication is based on port rights.  For example, a file
system server may give out a send right which allows access to a
particular file, simply by answering RPC messages transmitted to that
port.  The process receiving the send right is free to pass it to
another process, which would then obtain the same rights to access the
file.  Note that since the first process would be free to proxy any
requests to access the file, this isn't a security flaw.  In fact, the
process {\it can} proxy the requests by creating a new port of its
own, handling out a send right to that port, and relaying message to
the file system server, which would allow it to filter access to the
file.  This is roughly how file systems are implemented in Hurd.  The
file system server has read/write access to an entire disk partition,
but gives out send rights that only allow access to individual files.

The second additional structure in a Mach message is out-of-band data.
A Mach message can include pointer/length pairs to regions of memory,
along with a flag indicating whether those regions should be
deallocated when the message is sent.  If a region is not deallocated,
then it behaves much like a large block of opaque data copied in the
Mach message.  However, if a region is deallocated, then there is no
need to actually copy the data in the region, as hardware memory
management tables can be used to simply map it into the receiving
task's memory space.  This allows large blocks of memory to be
efficiently transferred in Mach messages, and forms the basis for
Mach's memory management system.

\section{Introduction to Mach Memory Management}

Mach's memory management was specifically designed to support
multi-node clusters.  Mach's model of memory management is that memory
is managed by user-space processes, and system memory is a cache of
that memory.  Any user-space process that implements Mach's memory
management protocol can be mapped into a process's address space.
When the process attempts to access memory, a page fault is triggered
which causes Mach to request the page from the user-space process.
The faulting thread remains suspended until the page has been supplied
to the kernel.

Mach's ability to attach out-of-band memory to messages is thus a
crucial component in the efficient implementation of its memory
management.  A page is typically supplied to kernel by attaching it to
a memory_object_data_supply message.  No copying needs be done, as the
page is transferred to kernel control simply by adjusting bits in the
hardware memory management tables.  Likewise, transferring a modified
page from the kernel to a user-space memory manager is done by
attaching it to a Mach message in a similar way.

Of course, it can be inefficient to keep a copy of a page in user
space while its also in use by the kernel.  Pages can be flagged
``precious'', indicating to the kernel that they must be returned to
the user-space process and can not be simply discarded.

The flexibility of allowing user-space processes to implement the
memory management protocol does create unique failure and abuse
scenarios.  A program can implement the memory management protocol,
map it into a region of memory, transfer that memory to another
process, then trigger a fault by responding with
memory_object_data_error to any access attempt.  The other program, if
not careful, could fail with a SIGSEGV just for trying to read the
memory it received in a message!  Thus, in Hurd, user-space programs,
especially system servers, must exhibit the same kind of care when
receiving memory regions in Mach messages as Linux kernel authors are
used to considering when a kernel routine receives memory from a
user-space program.

\section{The NETMSG Server}

To implement distributed virtual memory in Hurd, we need to transport
Mach messages across the network.  Several programs have been written
over the years to transport Mach messages over a network connection;
all collectively go under the name of the ``netmsg server'', or just
``netmsg''.

I wrote Hurd's current netmsg server in 2016.  It operates over a
single unsecured TCP/IP connection and has many limitations, but it
works well enough to implement shared virtual memory.  In server mode,
it presents a single node, by default the root of the filesystem, to
any clients that connect.  In client mode, it acts as a translator,
allowing a remote server's exported node to be mounted in the
local filesystem.

Out-of-band memory is simply
copied across the network connection, then deallocated.

\section{LIBPAGER}

Hurd's implementation of the Mach memory management protocol is
implemented in a library called ``libpager''.  The original libpager,
written in the mid-1990s, only supported a single client (i.e, a
single Mach kernel), and was thus suitable for single node operation.
Its API, however, is flexible enough to allow a more sophisticated,
multi-client libpager as a drop-in replacement.  For dynamically
linked servers, all that is required is to update the system shared
library to obtain multi-node functionality.

pager_read_page() is expected to allocate memory (unless it returns an
error) and pass it to libpager, which is responsible for deallocating
it (actually by passing it to the kernel).

pager_write_page() does no memory allocation or deallocation.  It
receives allocated memory and leaves it allocated, with libpager
responsible for dellocation (or passing it to the kernel).

pager_flush() and pager_flush_some() discard dirty pages!

pager_return() and pager_return_some() flush and return dirty (and precious) pages.

pager_sync() and pager_sync_some() return dirty (and precious) pages but don't flush.

The current kernel implementation seems to request pages one at
a time, as the process attempts to access them.

If we're servicing a page read-only to multiple clients, where do we
get the data from?  If our local kernel has a copy, we almost
certainly want to get it there, since it's just a manner of flipping
bits in a hardware page table to obtain a new "copy" of the data.
Otherwise, we probably want to get the data from the local disk, but
if the disk is very busy and we've got fast network connections, we
might want to get that data from another node in the cluster.

Mach's current memory object API makes this a lot more difficult.  You
can't just request that the kernel hand you back an unmodified page,
except by flagging it "precious".  That's not too bad if the kernel
and the memory manager are both local on one node, but if you do this
over the network, you're sending unnecessary copies of the memory
pages around.

How do we figure out which client is the local kernel?  One possible
solution is for netmsg to listen on /servers/netmsg for an RPC that
accepts a send right and identifies which node currently hosts the
corresponding receive right.

[[ the kernel uses different control ports for different memory objects ]]

  If the kernel uses the same control port for all of its pager
  operations, then we wouldn't have to make this query for every memory
  object; a single ext2fs translator, for example, could globally
  identify the location of all of the control ports, one for each node,
  and this information could be shared amongst all of that translator's
  pager objects.

DISTRIBUTED LIBPAGER.  We could use a shim process on remote nodes, to
allow requests for an existing page go to a node that holds a copy of
that page, instead of to the node with the disk.  Two ways I can
imagine this.  Either ext2fs forks off multiple processes on different
nodes, or a mechanism is developed to detect when a port is remote,
figure out which node it (currently) resides on, and then libpager can
fork a process and push it to the other node, invisible to ext2fs
proper.

CONFIGURATION OPTIONS

- do we write a page back every time we handle it?
  perhaps always call write_page, and add a flag to write_page indicating that we still have a copy?
- how long do we let clients hold a page if other clients are waiting for it?
  currently: 0 (immediately send lock requests)
  could create deadlock if the kernel returns pages without making any forward progress
- how do we obtain pages when we can read them from multiple sources?
  only applies to READ access; local kernel should be preferred
- handle network partitions / failed nodes / malfunctioning kernels
  timeout while waiting for a client?
  CAP theorem - timeouts would sacrifice consistency to achieve availability

\section{The Pagemap}

Hurd's original \libpager had a fairly simple pagemap.  The pagemap was
an array of 16-bit shorts; each page had a single 16-bit entry, allocated
as follows:

\begin{verbatim}
#define PM_WRITEWAIT  0x0200   /* queue wakeup once write is done */
#define PM_INIT       0x0100   /* data has been written */
#define PM_INCORE     0x0080   /* kernel might have a copy */
#define PM_PAGINGOUT  0x0040   /* being written to disk */
#define PM_PAGEINWAIT 0x0020   /* provide data back when write done */
#define PM_INVALID    0x0010   /* data on disk is irrevocably wrong */

#define PM_ERROR(byte) (((byte) & 0xc) >> 2)
#define PM_NEXTERROR(byte) ((byte) & 0x3)

enum page_errors
{
  PAGE_NOERR,
  PAGE_ENOSPC,
  PAGE_EIO,
  PAGE_EDQUOT,
};
\end{verbatim}

Four bits are allocated for error code.  (all other errors get collapsed into EIO)

The remaining bits indicate if the page is in core, if we're waiting
for it to be paged in or out (typically to or from disk).

Clearly such a simple pagemap is inadequate for multi-client
operation.  For one thing, we could have a potentially unlimited
number of clients with read-only copies of the page, as well as a
potentially unlimited number of clients waiting on the page, so
a fixed-size structure seems inadequate.

Nor is a separate structure for each page desirable if there are
a large number of pages.  Since the state of one page is often
identical to the state of many others, a pagemap of pointers
suggests itself.

The new \libpager is coded in C++, so there were several implementation options:

\begin{itemize}

\item  A straight-up array of pointers

Easy enough to understand, but then we have to
code reference counting to know when those dynamic structures can be
deallocated.

\item An array of {\tt shared_ptr}'s

Simplifies the code, because
it does all the reference counting for us, but at the cost of making
the array four times bigger than it needs to be.

\item A custom class with a clever copy constructor

We just assign into the array and the copy constructor takes care of
either finding an existing pagemap structure or creating a new one, at
the cost of obscuring the fact that {\tt pagemap[i] = new_client_list;} is
far more complex than a simple assignment.

\end{itemize}

I used the clever copy constructor.  The pagemap is an
array of pointers to unique pagemap structures kept in a C++ {\tt std::set}
called {\tt pagemap_set}.  When working on a page, the main library
code copies a pagemap entry into a temporary structure ({\tt
  tmp_pagemap_entry}), modifies it, then assigns it back into the
pagemap using an {\tt operator=} that either finds a matching entry in
{\tt pagemap_set} and uses a pointer to it, or move-inserts {\tt
  tmp_pagemap_entry} into {\tt pagemap_set} and uses a pointer to it.

All the queues and lists are private, and we've got a bunch of methods
to access them, allowing us to change the pagemap structure later if
we want.

\begin{verbatim}
struct pagemap_entry {
   /* ACCESSLIST
    * A list of clients that currently have access, along
    * with an indication if this is read-only or read-write access.
    */
   port_t clients_with_access[];
   bool write_access_granted;

   /* WAITLIST
    * A list of the clients waiting for access, in order of arrival,
    * and what kind of access they are waiting for (read or write)
    */
   port_t clients_waiting[];
   boolean clients_waiting_for_write[];

   /* PAGINGOUT
    * A flag that we set when we get a data return and start writing it
    * to backing store, and clear when the write is finished.
    */
   bool pagingout;

   /* ERROR
    * the last error code returned from a backing store write operation
    * (hopefully KERN_SUCCESS)
    */
   kern_return_t error;
};

Global structure {
   /* NEXTERROR list
    * A linked list of outstanding (error code, page, client) tuples.
    * They'll be transmitted to the client that requested them.
    */

   /* WRITEWAIT list
    * A linked list of pages waiting to be written out to disk
    */

    std::map<std::pair<vm_offset_t, vm_size_t>, outstanding_lock> outstanding_locks;

    std::set<outstanding_change_request> outstanding_change_requests;

   ERROR code - a kern_return_t, so we're not limited to a small number of errors

   /* Flags */
   PAGINGOUT - set this flag when we get a data return, clear it after the write is finished
   INVALID - set if the data on disk is invalid because we got an error writing the page back to the filesystem
      If some clients still have copies of the data, we may be able to operate for a while, then retry the write operation
}
\end{verbatim}

How to maintain the pagemap for a large number of clients?  Each page
has a pointer to a structure indicating which clients currently hold
copies of the page.  We anticipate that many pages (say, all the pages
in a single shared library) will have the same set of clients, so
there's one structure for each combination of clients, with multiple
pages pointing to it.  Pagemaps can be potentially large, so a pagemap
entry should be a single pointer.

Each file corresponds to a separate memory object, and separate memory
objects get separate control ports in the kernel.  A future
enhancement could be for the kernel to use a single control port for
all of its memory objects, allowing libpager to reduce the number of
these combination structures, if different files have the same usage
pattern (a program and some of its shared libraries, for example).

Each pagemap entry has an ACCESSLIST and a WAITLIST.  ACCESSLIST is
unordered; WAITLIST is ordered.  ACCESSLIST is stored sorted.

What happens when either list gets a new client (or loses one)?  We
have to search in the structures for the new combination, and create a
new structure if the desired combination doesn't exist yet.

\begin{comment}

We currently use a C++ std::set and let it do the searching for us.

Other possibilities:

  How can we organize the structures?  First, by the number of clients
  on each list.  We'll always know how many clients are in the
  ACCESSLIST we're searching for, so if we've got six ACCESSLIST clients
  and are moving the only client on WAITLIST to ACCESSLIST, we'll search
  in the seven-client ACCESSLIST, zero-client WAITLIST tables.  We can
  use a hash table or a tree.

  Because ACCESSLIST is sorted, we can binary search to find an
  ACCESSLIST that matches the one we want, then linear search forward
  and backward to find the one we want.

  or... Keep ACCESSLIST and WAITLIST completely separate for each other.
  Each pagemap entry has an ACCESSLIST pointer and a WAITLIST pointer,
  as well as some flags, including WRITE-ACCESS, which indicates if
  we've given out write access to a single client, in which case
  ACCESSLIST isn't a pointer, but a single client port number.  If
  WRITE-ACCESS is false, then ACCESSLIST points to a NULL-terminated
  list of client port numbers.  The disadvantage here is that the
  pagemap entries require more space (at least two pointers).

  WAITLISTs are ordered and include a flag for each client indicating
  whether it's waiting for READ access or WRITE access.  A special case,
  indicated by a flag, could be a single client on a WAITLIST, in which
  case we don't need a pointer to a WAITLIST, but rather can use the
  port number instead of the pointer.

The pagemap structures do not count the number of pages pointing to
them.  If we did that, it would facilitate deallocation when the count
drops to zero, but would interfere with the "add table" pointer
optimization described below.

POINTER OPTIMIZATION (UNIMPLEMENTED)

To speed this process, each structure can maintain pointers to
accelerate this mapping.  Each structure would have an "add table"
would maintain mappings like "add client 123 -> structure 0x80fadde0".
If the "add table" doesn't have an entry, then we need to search all
the structures to find one.  A problem with this is the need to
invalidate those pointers when a structure is deallocated.  Another
problem is that it requires the structures to resize when these
pointers are added. Keep this idea for a future enhancement.

single client pointer optimizations
  1. no access ->2 ->3
    pagemap_entry * put_client_1_on_waitlist_for_READ
    pagemap_entry * put_client_1_on_waitlist_for_WRITE
  2. waiting for read ->4
    pagemap_entry * move_first_WAITLIST_client_to_ACCESSLIST
  3. waiting for write ->5
    ditto (move_first_WAITLIST_client_to_ACCESSLIST)
  4. has read ->6
    ditto (put_client_1_on_WAITLIST_for_WRITE)
  5. has write
  6. has read; waiting for write ->5
    ditto (move_first_WAITLIST_client_to_ACCESSLIST)

PAGEMAP FOR A SMALL NUMBER OF CLIENTS (UNIMPLEMENTED)

I've thought of a possible optimization for small numbers of clients
that would largely avoid the need for separate pagemap structures, and
would transition to the unlimited size structures described above when
more clients connect that it can handle.

First, steal a bit from the pointer by aligning the structure on a
word boundary.  Then, if the least significant bit is 1, interpret the
pagemap entry as a bitfield rather than as a pointer to a more
complicated structure.

A 32 bit field (err... make that a 31 bit field) can handle up to seven
clients.  The new pagemap entries have an ACCESSLIST, which is an
unsorted set of clients (those who currently have access), and a
WAITLIST, which is a sorted list of clients that have requested
access, and a two boolean flags.

We number our clients 0 through 6, and I'm thinking above moving their
ports into high port number space to easily identify them.  So, the
first client on the first memory object would be on port 0x8000000,
the second client would be on 0x80000001, etc.  The second memory
object would have its ports on 0x80000008 through 0x8000000f.  So we
can just strip off the lowest 3 bits to figure out our "client
number".

We need a paging out flag, and a bit to indicate if the granted access
is read-only or read-write.  Assume that the {\tt error} field is
always {\tt KERN_SUCCESS}.  The ACCESSLIST requires one bit per client
to indicate if access is granted.  The WAITLIST is sorted (client
requests are processed FIFO) and requires four bits per slot -- three
to identify the client (all ones indicates an empty slot), and one to
indicate if the request is for read or write access.  Thus, we need
seven bits for the ACCESSLIST, two more flag bits (paging out and
read/write access), one bit to indicate that this isn't a pointer,
leaving room for five WAITLIST slots.

Once we hit our seventh client, or some oddball condition (like an
error return other than the expected KERN_SUCCESS, or a sixth client
on the WAITLIST), we shift to a pointer that points to a more
complicated, dynamically allocated structure.

Likewise, a 64-bit pointer (wishful thinking at the moment, since Hurd
is currently a 32-bit system) would allow fifteen clients, with an
eleven client WAITLIST.  Both the 32-bit and the 64-bit scheme have
two bits left over, perhaps they could be used to allow EIO, ENOSPC,
and EDQUOT, in addition to KERN_SUCCESS, since these seem to be the
most likely error codes.

\end{comment}

ACCESS CONTENTION

Putting a client on an empty queue triggers lock requests to all
clients in the working set, with the exception of the requesting
client itself, if it's already got read access and is requesting write
access.  Once the working set size drops to zero (or one), then the
original request can be satisfied.

In the meantime, additional clients can be added to the queue.  Each
client should be allowed access to a page for a period of time before
a lock request is sent.  Is the easiest way to do this to use the
timeout parameter on mach_msg?

Current implementation waits until the kernel returns a page, then
writes it to disk.  If a request for the page comes in before the
write completes, we wait for the write to finish, then send it right
back out without needing a read.

The new implementation will be handling pages as they're passed from
one client to another.  Should they be written out each time they are
handled?  Not write them at all until they're finally released?  Have
some kind of timer or counter to write them at a controlled pace?
Initially, we'll not write them at all until they're finally released,
which mimics the current behavior.

So when the last client returns a page, we look to see if anything is
in the wait queue.  If so, the first client(s) in the queue get the
pages.  If not, we flag that a write is in progress and start it.
When it completes, we check the queue again to see if anything is now
waiting.  If so, we service it.  If not, we discard the data and
notify the user that the page has been evicted.

\section{Pseudocode}

\begin{verbatim}
service_waitlist: (pass in a pagetable pointer, a data pointer, a read/write flag, a deallocate flage)
  XXX: data length must be page_size, as this logic is for a single page
  call with pager locked
  use m_o_data_supply to supply first set of WAITLIST clients
    all but last specify deallocate=false; last one specifies deallocation flag as passed in
  remove those clients from WAITLIST and move them to ACCESSLIST
  what if WAITLIST still has clients on it after sending messages?
     [ ] use a timeout before sending more lock requests?
     [X] send lock requests right away to everything on ACCESSLIST
\end{verbatim}

\begin{verbatim}
send_error_to_WAITLIST (pass in an error and offset):
  XXX: data length must be page_size
    ( XXX this causes a data_error to be sent for each page, when we could consolidate them, )
    ( but hopefully data_error is a corner case.  For ENOSPC or EDQUOT, though, not really. )
  send m_o_data_error's to everything on WAITLIST and clear WAITLIST
    anything on WAITLIST that's also on ACCESSLIST should get a lock instead of data_error
      and be put on NEXTERROR list
  (such a client sent an unlock request and is waiting for a lock, not data_error)

  [ ] flush everything else on ACCESSLIST
  [X] do nothing to ACCESSLIST
  [ ] do something more clever with stuff on ACCESSLIST
     (we got a read error, but other clients have copies of the data)
\end{verbatim}

\begin{verbatim}
finalize_unlock: (pass in a page number and an error code)
  if page not flagged ERROR:
     send m_o_lock_request to first client on WAITLIST (no reply), remove this client from WAITLIST,
     and upgrade its ACCESSLIST entry to reflect WRITE access
     if there are additional clients on WAITLIST:
        [ ] use a timeout before sending a lock request to this client
        [X] send a lock request right away to this client
            (XXX note that we just sent a lock request to answer the unlock; really want a timeout here)
  else (page flagged ERROR):
     send m_o_lock_request to first client on WAITLIST (internal flush variant)
     remove this client from WAITLIST
     add this client and error to NEXTERROR list
     (wait for reply and don't remove from ACCESSLIST until we get it)
     (other clients on WAITLIST will be processed when the lock is answered)
\end{verbatim}

\begin{verbatim}
m_o_data_request: (kernel requesting a single page)

  if this client is on NEXTERROR list and WRITE ACCESS WAS REQUESTED, send a m_o_data_error message,
        move NEXTERROR to ERROR, remove the NEXTERROR list entry, and return
  if the page is PAGINGOUT:
    if WAITLIST is empty and ACCESSLIST is not, send lock request (internal flush variant) to ACCESSLIST client
       ASSERT: there should only be one WRITE client on ACCESSLIST - the client that returned the data we're paging out
    add requesting client to WAITLIST and return
  if this client is already on ACCESSLIST, it flushed and is trying to re-aquire access,
     so check to see if WAITLIST is empty
     if so, remove client from ACCESSLIST and proceed (kernel flushed without being asked)
     if not, we're flushing, so add client to WAITLIST, run internal_lock_completed on this page, and return
  if WAITLIST is not empty, add requesting client to WAITLIST and return
  if another client has WRITE access (WAITLIST is empty), add requesting client to WAITLIST,
     send lock request (internal flush variant) and return
  if WRITE access is requested and ACCESSLIST is not empty (other clients have access),
     add requesting client to WAITLIST (it's empty),
     send lock requests (internal flush variant) and return
     (we could give out the page with READ access and wait for a unlock request, but I think not)
  else
     (ACCESSLIST is empty or READ access is requested and only READ clients are on ACCESSLIST)
     if page is flagged INVALID, send m_o_data_error to this client and return
     add this client to WAITLIST (it's empty)
     unlock pager
     read the page with pager_read_page()
     relock pager
     if pager_read_page() returned an error, mark page with ERROR and call send_error_to_WAITLIST
     if no error, service_waitlist(deallocate=true) (read/write status supplied by filesystem), clear ERROR
\end{verbatim}

\begin{verbatim}
m_o_lock_completed:
  m_o_lock_completed could have been sent in response to an internal lock request, or a fs flush, sync, return
  find client/offset/length in list of lock requests and decrement locks_pending
  if locks_pending is zero:
    if internal_lock_outstanding, and WAITLIST is not empty, then call internal_lock_completed
    remove client/offset/length from list
  if no more clients for this offset/length, wake up any waiting thread
\end{verbatim}

\begin{verbatim}
internal_lock_completed: (internal flush variant)
  use alloca() to allocate an array of flags, one set (in a uint8) for each page in current message:
    PAGEIN, UNLOCK, ERROR
  for all pages covered by message:
    [ if this client had WRITE access, print error (should have used m_o_data_return instead of m_o_lock_completed) ]
    if this client isn't on ACCESSLIST, do nothing (m_o_data_return already processed the data)
    if WAITLIST is empty, do nothing (m_o_data_request already called internal_lock_completed)
    else:
      remove client from ACCESSLIST
      if ACCESSLIST is empty but WAITLIST is not:
        if INVALID is not set: set PAGEIN flag
        else (INVALID set): call send_error_to_WAITLIST
          XXX: assumes that ERROR is correctly set whenever INVALID is set

      if ACCESSLIST has a single client with READ access and it's also the first client on WAITLIST requesting WRITE access,
        set UNLOCK flag
      otherwise, if ACCESSLIST has multiple clients, or a single client that's not first on WAITLIST,
        then we're still waiting for them to answer their locks, so do nothing
  unlock pager
  use alloca() to allocate an array of pointers and write lock flags, one of each for each page in current message
  for each page flagged PAGEIN:
    read the page with pager_read_page(), saving resulting pointer and write lock flag
    if pager_read_page() returned an error, set flag ERROR
    (a client with no access requested WRITE access and had to wait for other clients to flush)
  for each page flagged UNLOCK:
    call pager_unlock_page(), and set ERROR based on return value
    (a client with READ access requested WRITE access and had to wait for other clients with READ access to flush)
  relock pager
  for each page flagged PAGEIN:
    if page not flagged ERROR:
       service_waitlist(deallocate=true) (read/write status supplied by filesystem), clear ERROR
       XXX: this causes multi-page operations to broken up into single page ops
    else (page flagged ERROR): call send_error_to_WAITLIST
  for each page flagged UNLOCK: call finalize_unlock
\end{verbatim}

\begin{verbatim}
m_o_data_unlock: (kernel requesting write access when it's already got read access)
  lock pager
  alloca() an array of UNLOCK and ERROR flags, one pair for each page in message
  for all pages:
    ASSERT: this client already on ACCESSLIST with read access
    ASSERT: PAGINGOUT flag is not set (only should be set if a client had WRITE access)
    if WAITLIST contains at least one client waiting for WRITE access, do nothing and return
      (we're either already trying to flush everything on ACCESSLIST, or will, including this client)
      (if it saw the flush request before sending the unlock, it would never have sent the unlock)
      (so it hasn't processed the flush yet, and will interpret it as a response to the unlock)
    else if WAITLIST contains only clients waiting for READ access, add client to WAITLIST for WRITE and return
      (ACCESSLIST contains only READ clients and WAITLIST contains only READ clients)
      (so we're waiting for a page in to finish, as it will trigger a flush that will answer this unlock)
    otherwise, if there's other clients on ACCESSLIST, add client to WAITLIST for WRITE (it's empty)
      and send lock requests to other clients (internal flush variant) and return
    otherwise, we're the only client on ACCESSLIST and WAITLIST is empty:
      add client to WAITLIST for WRITE
      set UNLOCK flag
  unlock pager
  for all pages with UNLOCK flag set:
    call pager_unlock_page() and set ERROR based on return value
  relock pager
  for all pages with UNLOCK flag set: call finalize_unlock
\end{verbatim}

\begin{verbatim}
lock_object: (several types: flush, sync, return, response to an unlock request)
  NEW MODE NEEDED: asynchronous, but reply requested
  wrapper around m_o_data_lock message
  if sync requested, add to lock_requests list and increment count
     after m_o_lock_request message sent, wait for locks_pending and pending_writes to drop to zero,
     then decrement threads_waiting and remove lock_requests if zero

  lock_object() called by pager_ calls to flush, sync, and return data from kernel,
    as well as m_o_data_unlock

  if a m_o_data_return comes in while waiting for a synchronous lock to complete,
    we wait for the write to finish before returning from the lock
    the data return could be triggered by the lock request (i.e, a sync or a return)
  if a lock request comes in while a write is happening, we don't do anything to synchronize them
\end{verbatim}

\begin{verbatim}
m_o_data_return: (kernel returning pages)
  (this can not be used in liu of a lock completed message for a WRITE,
     because it doesn't indicate if the kernel still maintains WRITE access)
  npages = length / __vm_page_size;
  pm_entries points to 'npages' entries in pagemap
  lock pager
  ASSERT: this client should be all the page's ACCESSLISTs

  if kernel_copy is false:

    remove this client from all of these page's ACCESSLISTs
    [ ASSERT: these ACCESSLISTs should now be empty ]
    [ ASSERT: this client is not on any of the WAITLISTs (it had WRITE access, so there's nothing for it to wait for) ]
    [ not necessarily - if the pages are precious, they might be coming back with only READ access ]
    alloca() an array of UNLOCK, NOTIFY, and ERROR flags, one set for each page in message
    for all pages:
      if ACCESSLIST is empty and WAITLIST is not empty:
        service_waitlist(deallocate=false)
        (a client with no access requested WRITE access and the last client we were waiting for flushed)
      if ACCESSLIST has a single client with READ access and it's also the first client on WAITLIST requesting WRITE access,
        set UNLOCK flag
        (a client with READ access requested WRITE access and had to wait for other clients with READ access to flush)
      if both ACCESSLIST and WAITLIST are empty and pages are not DIRTY:
        set NOTIFY flag
        ( DIRTY pages will get their notifications done after their pager_write_page() calls )
    if any pages required UNLOCK or NOTIFY:
      ( this can't happen if pages are dirty )
      ASSERT: ! pages_dirty
      XXX: should this code be moved down further?
      unlock pager
      for all pages with UNLOCK set:
        call pager_unlock_page() and set ERROR based on return value
      for all pages with NOTIFY set:
        call pager_notify_evict()
        clear INVALID, ERROR, and NEXTERROR for this page
      relock pager
      for all pages with UNLOCK set: call finalize_unlock

  if pages are dirty:

    ASSERT: if kernel_copy = true, then this client is the only thing on ACCESSLIST
    ASSERT: if kernel_copy = false, then ACCESSLIST is empty
      (this client was the only thing on ACCESSLIST, because it had WRITE access to create a dirty page)

    [ ] for all pages with empty WAITLISTs and ACCESSLISTs:
        (XXX pager_sync needs to trigger a write even for pages with active clients)
    [X] for all pages in message:
         set PAGINGOUT in pagemap

    [ ] if PAGINGOUT was already set for any page, add this message, including kernel_copy flag
        and the pointer to the last lock incremented, to WRITEWAIT list and return
        ( allow simultaneous pager_write_page() calls for different pages )
    [X] if WRITEWAIT list is not empty, add this message (and kernel_copy and pointer) to WRITEWAIT list and return
        if WRITEWAIT list is empty, add this message (and kernel_copy and pointer) to WRITEWAIT list
           and process everything on WRITEWAIT list
        ( allow no simultaneous pager_write_page() calls )

  else (pages are not dirty):
    munmap() data
\end{verbatim}

\begin{verbatim}
service_first_WRITEWAIT_entry:
  use alloca() to allocated an array of flags, one set (in a uint8) for each page in current message:
    NOTIFY, ERROR, and PAGEOUT
    ( perhaps put this is a subroutine so that this array is deallocated after every message is processed )
  for all pages in current message:
    search WRITEWAIT list for a matching page
    if no match, set PAGEOUT flag
    ( this avoids writing a page if we've got more recent data waiting to write )
    ( this could be triggered by a rapid succession of pager_sync() calls on a busy page and a slow disk )
  unlock the pager
  call pager_write_page() on all pages flagged PAGEOUT
    set ERROR on any error return
  relock the pager
  for each page we just wrote:
    search WRITEWAIT list for matching pages
    ( can't use PAGEOUT flag because WRITEWAIT list might have changed while pager was unlocked )
    set INVALID equal to ERROR
    if something found on WRITEWAIT:
      ASSERT: kernel_copy is true (else how could there be a later write?)
      do nothing
    else if none found and WAITLIST is not empty:
      clear PAGINGOUT
      (this client had WRITE access because the page was dirty; so everything on WAITLIST is a data request)
      if kernel_copy is false: service_waitlist (deallocate = false)
      (if kernel_copy is true then we're still waiting for a flush to complete)
    else (none found and WAITLIST is empty):
      clear PAGINGOUT
      if kernel_copy is false, set NOTIFY flag and clear ERROR and NEXTERROR

  munmap() current message

  unlock the pager
  if client requested notify_on_evict, call pager_notify_evict() for any page flaged NOTIFY
    also, clear INVALID, ERROR, and NEXTERROR for these page
    ( the kernel didn't keep a copy and it didn't get shipped back out in an m_o_data_supply message )
  relock the pager (this can be cleverly wrapped into previous unlock/lock cycle)

  signal wakeup on anything waiting on this WRITEWAIT entry
  remove this WRITEWAIT entry from WRITEWAIT list
\end{verbatim}

\begin{verbatim}
pager_sync:
  only needs to do anything if a client has WRITE access
  might need to do multiple locks if different clients have WRITE access to different pages
\end{verbatim}

\begin{verbatim}
pager_return:
  scan pagelist for specified range and form the union of all clients on the ACCESSLISTs
  send multiple locks for multiple clients
  set (or increment) locks_pending by the number of lock requests sent
\end{verbatim}

\begin{verbatim}
pager_flush:
  scan pagelist for specified range and form the union of all clients on the ACCESSLISTs
  send multiple locks for multiple clients
  set (or increment) locks_pending by the number of lock requests sent
\end{verbatim}

\begin{verbatim}
m_o_terminate:
  mach_port_destroy() control and name ports passed in
    (we had send rights and should have just received receive rights in the m_o_terminate message)
\end{verbatim}

\begin{verbatim}
pager_shutdown:
  set terminating flag
  send pager_return (flush, return all, sync) to all active clients
  in the meantime:
    ignore all m_o_data_request and m_o_unlock requests
    m_o_lock_completed messages: don't do "internal" processing, but do notify threads
    process m_o_data_return messages, but don't generate any lock requests or data supply messages as a result
    [ ] handle new m_o_init_object messages by adding them to client list, but not sending m_o_ready
        that way, they'll get m_o_destroy in response
    [X] ignore m_o_init_object messages
  [ "De-allocating the abstract memory object port also has this effect" - Mach Kernel Principles p. 43 ]
  [ so... we don't have to do this section ]
  [ send m_o_destroy (error = ENODEV) to all active clients
  [   now handle new m_o_init_object messages by sending m_o_destroy in response and adding to client list
  [ wait for m_o_terminate replies from all clients
  call ports_destroy_right, which will blow away our receive right, and ultimately call the destructor,
    which will then call drop_client on any remaining clients
\end{verbatim}

SHUTDOWN CODE

Old libpager would flush everything (no special flags set), then
destroy the receive right and ignore any final messages.  This created
a minor race condition where clients could request a page while the
flush was processing, which would then get dropped.

BUG IN OLD CODE (NOT):
  client has WRITE access and page is being actively changed
  pager_write() is relatively slow
  pager_sync() is called asynchronously three times in rapid succession
  client returns data three times
  first m_o_return_data starts a pager_write and sets PAGINGOUT
  next two m_o_return_data's both see PAGINGOUT and set WRITEWAIT
  first pager_write finishes, sees WRITEWAIT, and wakes other two threads
  which one goes first?  coin flip
  thus, there's a chance that the second and third m_o_return_data's will be processed in reverse order
  actually, current implementation avoids this problem
    The second and third m_o_return_data's are both queued by the demuxer and don't
    get serviced until the first one finishes.  Then they are processed in order.

BUG IN OLD CODE:
  m_o_data_supply is called in data-return.c with precious = 0, instead of notify_on_evict

BUG IN OLD CODE:
  in pager_shutdown(), pthread_mutex_unlock() is called after ports_destroy_right(), which
  might have free'd the struct pager

CAVEAT IN OLD CODE:
  if we get a write error during a m_o_data_return, and we're flagged PAGEINWAIT because of an outstanding READ request,
  then the INVALID flag doesn't get set, but after we hand out the data, we won't get it back from the kernel
  (assuming that notify_on_evict is FALSE and thus pages are not flagged PRECIOUS)
  then we've dropped data without flagging anything; next read request will read old data from the disk

BUG IN OLD CODE (NOT):
  page 1 is dirty
  pager_sync() is called
  page 1 is returned and begins PAGINGOUT
  pages 1 and 2 are changed and become dirty
  pager_return() is called
  pages 1 and 2 are returned; m_o_data_return sees that page 1 is PAGINGOUT, flags it WRITEWAIT and sleeps
  page 2 is requested by the kernel
  page 2 has no flags set, so a pager_read_page() is started, even though more recent data is waiting to be written
  currently, serialization in demux prevents this problem, too, as the page 2 data_request won't start
    processing until the earlier data_return's complete

ISSUE:
  client on WAITLIST could have sent an unlock request and be waiting for a lock, not data_error
  a unlock request will come from a client on ACCESSLIST (for READ) and will have it put on WAITLIST (for WRITE)
  there could be an outstanding flush if something is already on WAITLIST for WRITE

RESOLVED ISSUE:
  data_request logic could cause multiple overlapping calls to pager_read_page() if a bunch of clients
    request READ access near simultaneously

RACE CONDITION:
  client requests access
  client is put on WAITLIST and pager_read_page() begins
  another client requests contradictory access via m_o_data_request
  NOT [ we send a lock request to original client, even though it hasn't gotten the data yet! ]
  NOT [ then pager_read_page() completes, and we send the data to the original client! ]
  YES [ no messages are sent since WAITLIST is not empty ]
  YES [ new client is put on the end of WAITLIST ]

RACE CONDITION:
  client B has READ access
  client A requests WRITE access (m_o_data_request)
  lock (flush) request is sent to client B and client A is added to WAITLIST
  nearly simultaneously, client B requests WRITE access (unlock)
  do nothing, since client B wouldn't have sent unlock request if it had already received lock request,
    so it will interpret the lock request as an answer to the unlock request

RACE CONDITION:
  client A has WRITE access
  pager_return()  (pager_sync would have kernel_copy = true; pager_flush never returns data)
  client A is sent a lock request
  client B requests WRITE access
  client A is sent another lock request, and client B goes on WAITLIST
  client A sends back page (kernel_copy = false)
  client B gets serviced
  client A sends completions for both lock requests
  RESULT: pager_write_page() is never called; data just moved from client A to client B
  ( old code would always call write_page(), then set PAGEINWAIT flag on request, then service request after page out finished)

Q: when could a client on WAITLIST be waiting for an lock, not a data_supply?
A: when that client already had READ access and sent an unlock request to get WRITE access
  if a client sent an unlock request, then it already had READ access
  how could this happen if there's (other) clients on WAITLIST?
  1. a lock/flush message has already been sent to service the clients on WAITLIST,
       the client sent the unlock request before processing the lock/flush,
       so essentially the unlock has already been answered and can be ignored
     this happens if a client is waiting for WRITE access when other clients had access (of some kind)
  2. the clients are on WAITLIST because we're waiting for paging (in or out) to complete
     in this case, we need to send a lock

RACE CONDITION:
  client A has READ access
  client B requested READ access, and we start paging in (client B goes on WAITLIST)
  client A requested WRITE access (via unlock), so it goes on WAITLIST
  client B's pager_read_page() returned an error, so we want to send data_error to B
     and lock to client A, with a queued NEXTERROR for client A

RACE CONDITION: (not sure about this - idea is to trigger something very similar to what is matched in above code)
  client A has WRITE access
  client B requests READ access
  lock request is sent to client A
  client A returns page with data_return, and client B gets the data and goes on ACCESSLIST
  client B returns the data and is removed from ACCESSLIST
  client C requests READ access, goes on WAITLIST and a pager_read_page() starts
  client A's lock completed message is processed, triggering a second pager_read_page(), since ACCESSLIST is empty

RESOLVED ISSUE: if multiple clients with READ access attempt to unlock near simultaneously,
  data_unlock logic will trigger multiple lock requests (a set for each client attempting to unlock)

ISSUES:
  before pager_write_page() returns, we can't call pager_read_page() or pager_write_page()
  after pager_write_page() returns, we might have added additional clients to WAITLIST,
    but did we add them while waiting for pager_write_page() or were they there before?
    if ACCESSLIST clients have READ access, and first WAITLIST clients are requested READ access, they are new
    if ACCESSLIST clients have READ access, and first WAITLIST client is requesting WRITE access,
      they may or may not be new
    if ACCESSLIST client has WRITE access, and WAITLIST isn't empty,
      they may or may not be new
    if we wait until pager_write_page() returns to send lock requests, we avoid having to figure out
      if we've already sent lock requests, and introduce some delay to let clients use the data

HOW ABOUT:
  before calling pager_write_page(), service the first batch of clients on WAITLIST
     and move them to ACCESSLIST, but send no lock requests
  after pager_write_page(), service any compatible WAITLIST clients and move them to ACCESSLIST
  if there are still outstanding WAITLIST clients, send lock requests to everyone on ACCESSLIST

RACE CONDITION:
  client A has WRITE access
  pager_return() is called; a lock request is sent to A
  client B requests access; a lock request is sent to A
  client A answers the pager_return's lock request with a data_return (kernel_copy = false; dirty = true)
  data_return code services WAITLIST and sends data to B
  second lock requests answers and does nothing since A is no longer on ACCESSLIST
  should be OK

RACE CONDITION:
  client A has WRITE access
  pager_sync() is called asynchronously
  client A syncs with a m_o_data_return, but keeps a copy of the data
  near simultaneously, client B requests access and libpager requests a flush
  client A answers with a new m_o_data_return containing data different from the first
  libpager sees the first m_o_data_return and hands it out to client B
  RESULT: client B does not have the most recent changes to the data
  [ ] SOLUTION 1: if m_o_data_return indicates kernal_copy=true, we can't hand it out to other clients
         PROBLEM: m_o_data_return doesn't indicate if the kernel has released WRITE access,
                  so this doesn't work if client B is requesting READ access and the lock
                  request is only to revoke WRITE access
  [X] SOLUTION 2: when we request a flush, wait for a lock_completed message

LOCK SEQUENCE:

client had READ access
FLUSH LOCK
NON-FLUSH LOCK (sync)
FLUSH LOCK COMPLETED (waits for non-flush lock before processing)
DATA REQUEST from flushed client for READ access - it's on ACCESSLIST, but has already flushed
  "if this client is already on ACCESSLIST, print error message and return"
  if other clients had READ access, pager_read_page() gets called
NON-FLUSH LOCK COMPLETED
  only now does the flush get processed
  if client on ACCESSLIST, stuff happens

client had READ access
lock_request (FLUSH)
flush occurs internally to the kernel, but the lock_request requires more processing
there's another access to the page
DATA REQUEST from flushed client for READ access - it's on ACCESSLIST, but has already flushed
  "if this client is already on ACCESSLIST, print error message and return"
  change to: if this client is already on ACCESSLIST for WRITE access, print error message and return
                XXX: what if kernel had WRITE access, but page wasn't dirty, so it didn't get returned?
             if this client is already on ACCESSLIST for READ access, it flushed and is trying to re-aquire access,
                so check to see if we've got an outstanding lock w/ flush flag set
                if so, [X] add client to WAITLIST and return (flush code will run later)
                       [ ] clear flush flag and run flush code now (more complicated but a bit faster)
                otherwise, remove client from ACCESSLIST and proceed (kernel flushed without being asked)
  if other clients had READ access, pager_read_page() gets called
lock_completed message
  only now does the flush get processed
  if client on ACCESSLIST, stuff happens

multi-page flush (pager_flush or pager_return)
lock_request sent to a client on a page where that client has no access
lock_request processed on that page
client requests access to that page

multi-page flush (pager_flush or pager_return)
lock_request sent to a client on a page where that client has no access
client requests access to that page
lock_request processed on that page

FLUSHING:

Because of the kernel's ability to proactively flush pages (unless PRECIOUS is specified),
we can't be certain that a client on a page's ACCESSLIST actually possesses that page.
All we know for sure is that a client not on ACCESSLIST does not possess the page.

Examining the Mach kernel code for m_o_lock_request, we see that the kernel might block
while looping over the pages (and unlocks the memory object while doing so), so for a
multi-page lock request, we need to consider the possibility that part of the page range
might be processed, then other kernel operations happen (including VM page operations)
before the remaining part of the page range is processed and the lock request completes.

Scenario A
- five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
- pager_flush is called on pages 0-100
- page 50 flushes on client A
- client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
- the flushes complete
- only client A now has access to page 50 and should be alone on its ACCESSLIST

Scenario A shows that ACCESSLIST, WAITLIST and a FLUSHING flag aren't enough; we have to track
which clients requested access during the flush to correctly compute ACCESSLIST at the end.

PRECIOUS PAGES

Flagging a page precious (so the kernel always returns it to us) isn't that big a deal
for a local client, because "returning the page" just means flipping bits in the VM tables,
but it's a lot heavier operation for a remote client, since the page will be transmitted
across the network unnecessarily.

Flagging all of our pages "precious" is tempting.  It provides us with a mechanism
for retrieving a local, in-memory page that we want to serve out over the network
without having to read it again from disk.  It also simplifies the lock completed
logic, since we don't have to worry about whether the page got returned or not
when we get the lock completed message - precious pages always get returned.

initial client goes on WAITLIST only when:
1. client has requested any access and page is PAGINGOUT, or
2. client has requested any access and another client has WRITE access
3. client has requested WRITE access (data request or unlock) and another client has any access
4. client has requested any access and is waiting for the page to page in
     READ access requested - only READ clients on ACCESSLIST
     WRITE access requested - ACCESSLIST empty

additional clients go on WAITLIST when they request almost any access
additional clients DON'T go on WAITLIST when:
1. we're sending an immediate m_o_data_error
2. they're trying to unlock and we anticipate that they'll get a flush soon by virtue of being on ACCESSLIST

Theorem: if WAITLIST is not empty, and both ACCESSLIST and WAITLIST contain only READ clients,
         then we're waiting for a page in to complete
Proof: ???
Converse: false; while waiting for a page in, we could get a request for WRITE access

Theorem: there can be at most one client on both ACCESSLIST and WAITLIST simultaneously

Theorem: if anything is on WAITLIST, we've got a flush request outstanding

Theorem: if WAITLIST is empty, we've got no flush requests outstanding

PROBLEM CONDITION:
client A has READ access
client B requests READ access
page in starts; client A on ACCESSLIST; client B goes on WAITLIST (for READ)
client A requests WRITE access via unlock request
WAITLIST is not empty, but we're not trying to flush
--- what should happen? ---
add client A to WAITLIST for WRITE
wait for page in to finish; client B gets data and moves to ACCESSLIST
now A and B are on ACCESSLIST with READ access, and A is on WAITLIST for WRITE
client B is sent a flush request, after an optional delay
once client B flushes, client A is unlocked and is sent a lock message
now A is on ACCESSLIST with WRITE access

...vs...

client A has READ access
client B requests WRITE access
flush sent to client A; client A on ACCESSLIST; client B goes on WAITLIST (for WRITE)
client A requests WRITE access via unlock request
--- what should happen? ---
nothing - client A will process the flush as an answer to its unlock request
after flush is acknowledged, page in starts
after page in finishes, client B gets data and moves from WAITLIST to ACCESSLIST

if a client is requesting unlock, then it's already on ACCESSLIST for READ
if anything is on WAITLIST for WRITE, then we're trying to flush  (right?)
if the only things on WAITLIST are for READ, then we're waiting for a page in

client A has READ access
client B requests READ access
page in starts; client A on ACCESSLIST; client B goes on WAITLIST (for READ)
client C requests WRITE access via data request
cleint C gets added to WAITLIST (for WRITE)
client A requests WRITE access via unlock request
WAITLIST is not empty, and there's a client on it waiting for WRITE, but we're not trying to flush
--- what should happen? ---
add client A to WAITLIST for WRITE
wait for page in to finish; client B gets data and moves to ACCESSLIST
now A and B are on ACCESSLIST with READ access, and C and A are on WAITLIST for WRITE
clients A and B are sent a flush request, after an optional delay
the flush requests answers A's unlock, so it comes off the WAITLIST
once clients A and B flush, client C gets the data
client A should re-request WRITE access via data request

... or ...

client A has READ access
client B requests READ access
page in starts; client A on ACCESSLIST; client B goes on WAITLIST (for READ)
client C requests WRITE access via data request
cleint C gets added to WAITLIST (for WRITE)
client A requests WRITE access via unlock request
WAITLIST is not empty, and there's a client on it waiting for WRITE, but we're not trying to flush
--- what should happen? ---
nothing - client A will soon get a flush
wait for page in to finish; client B gets data and moves to ACCESSLIST
now A and B are on ACCESSLIST with READ access, and C and A are on WAITLIST for WRITE
clients A and B are sent a flush request, after an optional delay
once clients A and B flush, client C gets the data
client A should re-request WRITE access via data request

PROBLEM

client A has READ access
client B requests WRITE access, so a flush request gets sent to client A and client B goes on WAITLIST
   ACCESSLIST: A(read)   WAITLIST: B(for write)
client A requests WRITE access (unlock request) and goes on WAITLIST
client A processes flush and requests WRITE access, so goes on WAITLIST again with a data request
client A is now on WAITLIST twice

\section{Random Test Code}

I wrote a test program that excercises libpager by making a
pseudo-random string of pager requests.  The idea was to run it
repeatedly using various seed values, then if it detected a bug, it
could be re-run with the same seed value in order to reproduce the
bug.  It never actually found any bugs.

There are five possible client operations:

\begin{enumerate}
\item request READ access
\item request WRITE access
\item request unlock READ $\to$ WRITE
\item flush page
\item service a message from the memory manager
\end{enumerate}

If the client has no access to a page, then 1, 2, or 5 are possible.
If the client has read access, then 3, 4, or 5 are allowed.  With
write access, 4 or 5 are permissible.

translator operations:
1. discard request
2. return request
3. sync request
4. complete a pending pager_read, pager_write, or pager_unlock (which one? there could be several)

1 2 3 can be synchronous or asynchronous

The test program should check to ensure that following conditions are met:

[ ] write access to a page is exclusive to a single client
[p] pages get written back in-order (though skips are possible)
[ ] if a page is modified, nobody ever sees an earlier version of it
[ ] all data requests are properly answered with a data supply or a data error (or data unavailable)
[ ] all data unlock requests are properly answered with a data lock
[ ] if a data lock requests a reply, make sure one is received

p = partially done

Test for BUG 1:
  A2 T4 (the read completes) A5 (the data_supply is processed)
    T3 T3 T3 (the syncs) A5 A5 A5 (the syncs are processed) T4 T4 T4 (the writes complete)

client_request_read_access(client, start_page, end_page)
  ASSERT: page absent in client
  sends m_o_data_request
client_request_write_access(client, start_page, end_page)
  ASSERT: page absent in client
  sends m_o_data_request
client_request_unlock(client, start_page, end_page)
  ASSERT: page present in client with read access
  sends m_o_data_unlock
client_flush(client, start_page, end_page)
  ASSERT: pages present
  if pages have WRITE access, modify data and data_return
  if pages have READ access and are precious, data_return
client_service_message(client)
  read message
  data_supply - save pointers, read/write and precious (reply if requested)
  data_lock - send messages, update pointers, reply if requested
  data_error - ??

Test for BUG 3: (one client, two pages A and B)
  AB2 T4 (the read completes) TA3 A5 (the sync is processed) TAB2 A5 (the sync is processed)
    B2 (page request) T4 (the read completes)
    (now we need to close things out and see that data got discarded)

data should always be modified if possible (whenever a client's got WRITE access)

\section{POSIX Shared Memory}

Hurd's System V shared memory implementation, in glibc's
sysdeps/mach/hurd, operates by mapping "files" in the /run/shm portion
of the filesystem tree.  /run/shm is typically implemented using the
tmpfs translator, which is linked with libpager.  Thus, a multi-client
libpager enables the use of inter-node shared memory.  What is
required to make this happen is a shared /run/shm - shared with full
Mach IPC semantics, not just NFS - so netmsg is required.  unionfs can
possibly be used to overlay the various /run/shm's on each other.

\section{Distributed Tasks and Process Migration}

In the Hurd architecture, processes are localized to a node, but our
implementation of distributed shared memory allows cross-node forks
with full POSIX semantics.  A straightforward modification to
Hurd's C library would allow fork calls to spawn processes
on other nodes.

How can a process share threads between nodes?  Two main problems:
sharing Mach ports, and sharing memory.  Mach allows the allocation
of un-backed memory pages.  We need to attach a memory manager
to the un-backed pages to allow them to be distributed across nodes.
This requires relatively minor changes to Mach.

This would allow cross-node forks, but not cross-node threads, which
would be difficult to implement using Mach's current design.  Any
thread in a task can listen for messages on a Mach port, and messages
are only delivered once, so balancing the need to copy messages to
multiple nodes (since any thread could be the only one listening to a
given port) while guaranteeing unique delivery is quite problematic.

This suggests a natural division: processes are local to each node.  A
multi-threaded program can fork multiple times, and each forked
process can spawn multiple threads.  The processes are distributed
between nodes, while each process is local to a node and supports
multiple threads.

Linux's NUMA implementation uses a zonelist for memory allocation.
Memory allocation requests are attempted locally, and if this fails,
then we fall back into the zonelist, which gives us a list of other
nodes to try for memory allocation.

The linux scheduler (sched-domains.txt) occasionally balances
load between scheduling domains.

\section{Current Status and Future Work}

Both the netmsg server and the new libpager have been successfully
tested in a virtual machine running Hurd.

\begin{itemize}

\item 64-bit addressing and SMP

This is the most important work needed on Hurd.  We've got all of the
major pieces of a working cluster operating system, but nobody cares
since it can only use a single processor on each node and individual
processes are limited to a 4 GB address space.

\item improved netmsg

Use a reliable datagram delivery protocol, not TCP.

Direct access to network hardware.  Memory-mapped PCI hardware
accessed with a library.  This would require a network card
to be dedicated to netmsg, but with Cisco virtual NICs, that's
not such a big problem.

\item Useful Mach kernel changes

\begin{itemize}

\item allow memory manager to attach to unmanaged pages

  This is necessary to implement cross-node {\tt fork}'s with full POSIX semantics.

\item m_o_data_return could specify what permissions the kernel still holds on the page

  Then
  we could easily downgrade from WRITE to READ access without waiting for a lock completed
  message, or (like the current code), never downgrading from WRITE to READ.

\item m_o_lock_completed message could specify what kind of lock was completed

  This would avoid the
  need to match reply ports to lock messages unless (like the current code), we just
  wait for all outstanding locks to complete before continuing with any of them.

\item the ability to request a copy of a unmodified, un-precious page from the kernel

  This would allow \libpager to serve out the page to other nodes without reading it again from the disk.

\item ability to notify memory manager when a non-precious page has been evicted

    (over the network, precious is heavy weight just to get notification that a page has been evicted)

\item Use a kernel name port common across all memory objects

    this would allow sharing pagemaps between memory objects with the same usage pattern
    (i.e, kernels A and B both map files C and D)

\end{itemize}

\item distributed filesystem

\end{itemize}

\end{document}

### 2019-05-12 libpager/paper.tex
working on paper.tex

@@
 \usepackage{comment}
+
+\usepackage{../slides/hurd}
+\usepackage{tikz}
+\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes}
 
@@
 
-CAP THEOREM
-
-The CAP theorem states that we can't have all three of: consistency,
-availability, partitioning.  Right now, we sacrifice availability - if
-the network partitions, libpager can deadlock.  If we introduce a
-timeout on lock requests, then we're sacrificing consistency.
+The CAP theorem states that a distributed system can't have all of
+three things: consistency, availability, partitioning.  Right now, we
+sacrifice availability -- if the network partitions, libpager can
+deadlock.  If we introduce a timeout on lock requests, then we're
+sacrificing consistency.
 
@@
 
-In addition to opaque data (generally, but not necessarily interpreted
-as remote procedure calls), Mach messages can contain two more
-sophisticated structures.  The first are port rights.
-
-A Mach message can transport send or receive rights for Mach ports.
+Mach inter-process communications is primarily done using
+variable-sized datagram messages that are transmitted reliably and
+in-order to queues from which tasks can receive them.  In addition to
+opaque data (generally, but not necessarily interpreted as remote
+procedure calls), Mach messages can contain two more sophisticated
+structures.  The first are port rights, which come in two basic
+flavors.  Send rights give a task the ability to transmit messages to
+a queue, and receive rights give a task the ability to receive
+messages from a queue.  A Mach message can transport send or receive
+rights for Mach ports.
+
+\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=3em, text centered,
+  minimum width=5em, minimum height=2em, rounded corners]
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[node distance = .5cm]
+
+\task{Task 1}{at (0,0)}{br}{
+% print two sendright nodes, one to fill the message1 box to the correct size,
+% and another one to actually draw it, since message1 will overwrite the
+% first sendright
+
+\node (sendright) [draw, fill=orange] {Send 41};
+\node (message1) [fit=(sendright), message] {Message 1};
+\node (sendright) [draw, fill=orange] {Send 41};
+
+\node (message3) at (0, -7em) [message] {Message 2};
+}
+
+\task{Task 2}{at (15 em,0)}{bl}{
+
+\node (sendright2) [draw, fill=orange] {Send 9};
+\node (message2) [fit=(sendright2), message] {Message 1};
+\node (sendright2) [draw, fill=orange] {Send 9};
+
+\node (message4) at (0, -7em) [message] {Message 2};
+}
+
+\node (port 41) [left of=Task 1 Port -10, anchor=south] {41};
+
+\draw [thick, color=red, dotted, ->] (node cs:name=sendright, angle=-20) -- (port 41);
+
+\node [left of=Task 1 Port 5, anchor=south] {27};
+\node [right of=Task 2 Port 5, anchor=south] {52};
+%% \only<1,2>{\draw [thick, -triangle 90] (Task 1 Port 5) -- (Task 2 Port 5);}
+
+\draw [color=violet, thick, -triangle 90] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) -- (Task 2 Port 5)
+    -- +(.5,0) |- (node cs:name=message2, angle=160);
+
+\node [right of=Task 2 Port -10, anchor=south] {9};
+
+\draw [thick, triangle 90-] (Task 1 Port -10) -- (Task 2 Port -10);
+
+\draw [color=violet, thick, triangle 90-] (node cs:name=message3,angle=20) -- +(.3,0) |- (Task 1 Port -10) -- (Task 2 Port -10)
+    -- +(.5,0) |- (node cs:name=message4, angle=160);
+
+\end{tikzpicture}
+\end{center}
+\caption{Transferring a send right}
+\label{Transferring a send right}
+\end{figure}
+
+In Figure \ref{Transferring a send right}, Process 1 sends Message 1
+to Process 2 containing a send right to the queue known to Process 1
+as port 41, which could have been created by Process 1, or received
+from another task.  The Mach kernel translates this into a send right
+to a previously unused port 9 in Process 2's port number space.
+Process 2 can now send a message to its port 9, which will be received
+by Process 1 on port 41.
+
 This is basically the only way in which a process can obtain Mach
@@
 process can fabricate another process's port number, just as a POSIX
-process can not fabricate another process's file descriptor.
+process can not fabricate another process's file descriptor.  A
+process is provided a few standard ports when it is initialized, one
+of which being a port with send rights that represents the root of the
+file system.  By sending filename lookup requests to this port, the
+process can traverse the entire file system tree, each node of which
+is represented by a different port.  The hurd file system is thus a
+hierarchy of ports, most of which recognize a standard set of RPCs
+performing file operations.  The primary way for a program to
+advertise its services in Hurd is for it to listen for messages on a
+port that it attachs to a name in the file system.
 
@@
 When the process attempts to access memory, a page fault is triggered
-which causes Mach to request the page from the user-space process.
-The faulting thread remains suspended until the page has been supplied
-to the kernel.
+which causes Mach to request the page from the user-space process
+using a well documented protocol.  The faulting thread remains
+suspended until the page has been supplied to the kernel.
 
@@
 ``precious'', indicating to the kernel that they must be returned to
-the user-space process and can not be simply discarded.
+the user-space process and can not be simply discarded.  This frees
+the user-space memory manager from needing to keep a copy of a page
+that has been supplied to the kernel.  Otherwise, non-``precious''
+pages can be freely discarded by the kernel when they are no longer
+needed.
 
@@
 especially system servers, must exhibit the same kind of care when
-receiving memory regions in Mach messages as Linux kernel authors are
-used to considering when a kernel routine receives memory from a
-user-space program.
+receiving memory regions in Mach messages as Linux kernel routines
+must consider when receiving memory from a user-space program.
 
@@
 
-pager_read_page() is expected to allocate memory (unless it returns an
-error) and pass it to libpager, which is responsible for deallocating
-it (actually by passing it to the kernel).
-
-pager_write_page() does no memory allocation or deallocation.  It
-receives allocated memory and leaves it allocated, with libpager
-responsible for dellocation (or passing it to the kernel).
-
-pager_flush() and pager_flush_some() discard dirty pages!
-
-pager_return() and pager_return_some() flush and return dirty (and precious) pages.
-
-pager_sync() and pager_sync_some() return dirty (and precious) pages but don't flush.
+On one side, \libpager implements the Mach memory management protocol
+interacting with the Mach kernel.  On the other side, \libpager
+presents an API to the server program, which is responsible for
+producing the pages, as follows:
+
+\begin{itemize}
+
+\item {\tt pager_read_page()}
+
+This routine is expected to allocate memory (unless it returns an
+error) containing the requested data and pass it to \libpager, which
+is responsible for deallocating it (typically by passing it to the
+kernel).  A common use scenario is that {\tt pager_read_page()}
+reads a page from disk and returns it.  {\tt pager_read_page()}
+can be called from multiple threads to service multiple paging
+requests, and must implement its own thread safety.
+
+\item {\tt pager_write_page()}
+
+This routine is expected to save a modified memory page just before
+\libpager releases its last copy of the data.  It does no memory
+allocation or deallocation.  It receives allocated memory and leaves
+it allocated, with \libpager responsible for dellocation, or passing
+it to the kernel, if a new request has arrived in the time required
+for {\tt pager_write_page()} to complete
+
+\item {\tt pager_unlock_page()}
+
+\item {\tt pager_sync()} and {\tt pager_sync_some()}
+
+This routine requests that copies of modified pages be obtained
+and submitted to {\tt pager_write_page()}.  Doesn't otherwise
+interfere with \libpager operation.
+
+\item {\tt pager_flush()} and {\tt pager_flush_some()}
+
+discard dirty pages!
+
+\item {\tt pager_return()} and {\tt pager_return_some()}
+
+ flush and return dirty (and precious) pages.
+
+\item {\tt pager_get_error()}
+
+\end{itemize}
 
@@
 
-CONFIGURATION OPTIONS
-
-- do we write a page back every time we handle it?
-  perhaps always call write_page, and add a flag to write_page indicating that we still have a copy?
-- how long do we let clients hold a page if other clients are waiting for it?
-  currently: 0 (immediately send lock requests)
-  could create deadlock if the kernel returns pages without making any forward progress
-- how do we obtain pages when we can read them from multiple sources?
-  only applies to READ access; local kernel should be preferred
-- handle network partitions / failed nodes / malfunctioning kernels
-  timeout while waiting for a client?
-  CAP theorem - timeouts would sacrifice consistency to achieve availability
+\section{Configuration Options}
+
+Several implementation choices exist, and perhaps should be configuration options.
+
+\begin{itemize}
+
+\item How long do we let clients hold a page if other clients are waiting for it?
+
+  Currently we send a lock request immediately after supplying the
+  page, which could create deadlock if the kernel returns pages
+  without making any forward progress.  If processes on two different
+  nodes are both trying to write into a page, we might spend all of
+  our time ping-ponging the page and forth without anything actually
+  getting done.
+
+  Check Mach kernel source to see what happens in this case.  My guess
+  is that if the process can be started immediately, it is, and this
+  is only an issue if system load is high enough to prevent that.
+
+\item Do we write a page back every time we handle it?
+
+  Currently, we don't call {\tt pager_write_page()} until a page is
+  completely freed or sync'ed, but this can cause modified data to sit
+  indefinitely in RAM when we'd rather save it to disk periodically.
+
+  Note that if the server wishes to periodically sync data to disk, it
+  can call {\tt pager_sync()} on a timer thread.  The question is
+  whether we want to more proactively call {\tt pager_write_page()}
+  anytime we're handling a page that needs to be transferred between
+  two clients.
+
+\item How do we obtain pages when we can read them from multiple sources?
+
+  Currently, we read it again with {\tt pager_read_page()}, but if the
+  local kernel has a copy, it's probably faster to ask the kernel for
+  a copy.  We don't have a good way of detecting which client is the
+  local kernel, and the Mach memory management API doesn't provide a
+  way to request a copy of an unmodified, non-``precious'' page
+  from the kernel.
+
+  A corner case is what happens when the read fails with an error.
+  Currently, we propagate that error to everything waiting on the
+  page, even though we could request a copy of the page from one
+  of the clients that still has it.
+
+\item How to handle network partitions / failed nodes / malfunctioning kernels
+
+  Currently we deadlock if a client doesn't respond; perhaps we should
+  timeout while waiting for a client?  How to pick the timeout value?
+
+\item How to handle failed writes to the backing store?
+
+  Do we keep the page around in memory to service future requests, and
+  hope that maybe another write will succeed in the future, or discard
+  the page?  Perhaps keep it for a time?  How to decide when to
+  discard?  Maybe only keep it if EAGAIN is returned?
+
+  If we do discard the data, how do we handle future requests for that
+  page?  Currently, we return errors with no mechanism to recover.
+
+  If we kept failing writes around indefinitely, that would provide a
+  simple mechanism to implement a ramdisk -- the read operation always
+  returns an empty page and the write operation always fails.
+
+\end{itemize}
 
@@
 
-Hurd's original \libpager had a fairly simple pagemap.  The pagemap was
-an array of 16-bit shorts; each page had a single 16-bit entry, allocated
-as follows:
+All Mach memory management operations operate on memory pages, so
+\libpager needs to maintain a table that tracks the state of each
+page, called the pagemap.  Hurd's original \libpager had a fairly
+simple pagemap.  The pagemap was an array of 16-bit shorts; each page
+had a single 16-bit entry, allocated as follows:
 
@@
 
-I used the clever copy constructor.  The pagemap is an
-array of pointers to unique pagemap structures kept in a C++ {\tt std::set}
+I used the clever copy constructor.  The pagemap is an array of
+pointers to unique pagemap structures kept in a C++ {\tt std::set}
 called {\tt pagemap_set}.  When working on a page, the main library
@@
   tmp_pagemap_entry} into {\tt pagemap_set} and uses a pointer to it.
+No attempt is (currently) made to free unused entries in {\tt
+  pagemap_set}.
 
@@
    /* ERROR
-    * the last error code returned from a backing store write operation
+    * the last error code returned from a backing store operation
     * (hopefully KERN_SUCCESS)
@@
    kern_return_t error;
+
+   /* INVALID
+    * indicates that the backing store data is invalid because we got an
+    * error return from a write attempt
+    *
+    * error returns on read or unlock operations do not set "invalid"
+    */
+   bool invalid;
 };
@@
    /* WRITEWAIT list
-    * A linked list of pages waiting to be written out to disk
+    * A linked list of pages waiting to be written to backing store
     */
@@
     std::set<outstanding_change_request> outstanding_change_requests;
-
-   ERROR code - a kern_return_t, so we're not limited to a small number of errors
-
-   /* Flags */
-   PAGINGOUT - set this flag when we get a data return, clear it after the write is finished
-   INVALID - set if the data on disk is invalid because we got an error writing the page back to the filesystem
-      If some clients still have copies of the data, we may be able to operate for a while, then retry the write operation
 }
@@
 copies of the page.  We anticipate that many pages (say, all the pages
-in a single shared library) will have the same set of clients, so
-there's one structure for each combination of clients, with multiple
-pages pointing to it.  Pagemaps can be potentially large, so a pagemap
-entry should be a single pointer.
+in a single shared library) will exist in the same state, so there's
+one structure for each combination of its entries, with multiple pages
+pointing to it.  Pagemaps can be potentially large, so a pagemap entry
+should be a single pointer.
 
@@
 
+We currently use a C++ std::set and let it do the searching for us.
+
 \begin{comment}
-
-We currently use a C++ std::set and let it do the searching for us.
 
@@
 notify the user that the page has been evicted.
+
+ERROR HANDLING
+
+The three backing store callback functions can all return an error
+code.
+
+Errors when reading from backing store are fairly straightforward --
+we got an error, so all we can do is propagate the error to the
+clients.  Actually, if there are outstanding copies of the page we
+tried to read, we could obtain a copy from another client, but we
+don't do that currently.  Errors when writing are a bit more
+complicated -- we've still got a copy of the page, do we keep it
+around and hope that the write will complete successfully at some
+point in the future?  We don't do that currently, either.  Errors
+returned from an unlock operation should probably only affect the
+unlock request.  Should they also affect future unlock operations on
+this page, or should failing unlocks be retried?  Currently, a failing
+unlock is propagated to everything waiting on the page (both read and
+write requests); probably should only be sent to write requests.
+
+Also, error returns on unlock requests are inconsistently recorded in
+the pagemap.  The error return code in the pagemap is only really used
+if a write operation failed, in which case the page is discarded and
+future requests all return the error code.  An exception is if a
+kernel kept a copy of the page whose write failed (i.e, there was a
+sync request that triggered the write), in which case future
+operations fail until something causes the kernel to send us the page
+again, when we retry the write.
 
@@
 
-BUG IN OLD CODE:
-  m_o_data_supply is called in data-return.c with precious = 0, instead of notify_on_evict
-
-BUG IN OLD CODE:
-  in pager_shutdown(), pthread_mutex_unlock() is called after ports_destroy_right(), which
-  might have free'd the struct pager
-
 CAVEAT IN OLD CODE:
@@
 Scenario A
-- five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
-- pager_flush is called on pages 0-100
-- page 50 flushes on client A
-- client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
-- the flushes complete
-- only client A now has access to page 50 and should be alone on its ACCESSLIST
+\begin{itemize}
+\item five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
+\item pager_flush is called on pages 0-100
+\item page 50 flushes on client A
+\item client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
+\item the flushes complete
+\item only client A now has access to page 50 and should be alone on its ACCESSLIST
+\end{itemize}
 
@@
 
-translator operations:
-1. discard request
-2. return request
-3. sync request
-4. complete a pending pager_read, pager_write, or pager_unlock (which one? there could be several)
-
-1 2 3 can be synchronous or asynchronous
-
-The test program should check to ensure that following conditions are met:
-
-[ ] write access to a page is exclusive to a single client
-[p] pages get written back in-order (though skips are possible)
-[ ] if a page is modified, nobody ever sees an earlier version of it
-[ ] all data requests are properly answered with a data supply or a data error (or data unavailable)
-[ ] all data unlock requests are properly answered with a data lock
-[ ] if a data lock requests a reply, make sure one is received
-
-p = partially done
+There are four possible translator operations:
+
+\begin{enumerate}
+\item discard request
+\item return request
+\item sync request
+\item complete a pending pager_read, pager_write, or pager_unlock (which one? there could be several)
+\end{enumerate}
+
+1, 2, and 3 can be synchronous or asynchronous.
+
+The test program should (but currently doesn't) check to ensure that
+following conditions are met:
+
+\begin{itemize}
+\item write access to a page is exclusive to a single client
+\item pages get written back in-order (though skips are possible) (partially done)
+\item if a page is modified, nobody ever sees an earlier version of it
+\item all data requests are properly answered with a data supply or a data error (or data unavailable)
+\item all data unlock requests are properly answered with a data lock
+\item if a data lock requests a reply, make sure one is received
+\end{itemize}
 
@@
 
-  Then
-  we could easily downgrade from WRITE to READ access without waiting for a lock completed
-  message, or (like the current code), never downgrading from WRITE to READ.
+  Then we could easily downgrade from WRITE to READ access without
+  waiting for a lock completed message, or (like the current code),
+  never downgrading from WRITE to READ.
 
@@
 
-  This would avoid the
-  need to match reply ports to lock messages unless (like the current code), we just
-  wait for all outstanding locks to complete before continuing with any of them.
-
-\item the ability to request a copy of a unmodified, un-precious page from the kernel
+  This would avoid the need to match reply ports to lock messages
+  unless (like the current code), we just wait for all outstanding
+  locks to complete before continuing with any of them.
+
+\item ability to request a copy of a unmodified, un-precious page from the kernel
 
@@
 
-    (over the network, precious is heavy weight just to get notification that a page has been evicted)
+  Over the network, flagging a page ``precious'' is awfully heavy
+  weight just to get notification that a page has been evicted, since
+  a ``precious'' page will be sent back to the memory manager.
+  Knowing that a page has been evicted may eliminate the need for a
+  future lock request.
 
@@
 
-    this would allow sharing pagemaps between memory objects with the same usage pattern
+  This would allow sharing pagemaps between memory objects with the same usage pattern
   (i.e, kernels A and B both map files C and D)
 
+\item Add the ability for a kernel to respond to a lock request by
+  sending the page directly to another kernel, rather than back to the
+  memory manager.
+
+  This would cut our network utilization roughly in half for pages
+  dominated by contention between nodes.
+
 \end{itemize}

### 2019-05-19 libpager/paper.tex
a checkpoint of work on "Building an SSI Cluster" paper

@@
 \usepackage{comment}
+\usepackage{xspace}
+\usepackage{hyperref}
+\usepackage{mdframed}
+\usepackage{xcolor}
+
+\definecolor{lightblue}{rgb}{0.8,0.8,1.0}
 
 \usepackage{../slides/hurd}
-\usepackage{tikz}
-\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes}
+%\usepackage{tikz}
+\usepackage{tkz-graph}
+\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes, arrows.meta}
 
@@
 \author{Brent Baccala}
-\date{}
+\date{May 2019}
 
@@
 % Shortcuts just for this document
-\def\libpager{{\tt libpager\ }}
+\def\libpager{{\tt libpager}\xspace}
+\def\netmsg{{\tt netmsg}\xspace}
+\def\pagersync{{\tt pager_sync()}\xspace}
+\def\pagerwritepage{{\tt pager_write_page()}\xspace}
+\def\pagerreadpage{{\tt pager_read_page()}\xspace}
+\def\moinit{{\tt memory_object_init}\xspace}
+\def\modatasupply{{\tt memory_object_data_supply}\xspace}
+\def\modataerror{{\tt memory_object_data_error}\xspace}
 
@@
 
-Operating system designers have tried for years to leverage
-the power of multiple computers and develop cluster
-operating systems.  Notable attempts include VAX/VMS, various cluster
-Linux systems [MOSIX, refs], Beowulf, Plan 9, ScaleMP, and
-Hadoop.
+Operating system designers have tried for years to leverage the power
+of multiple computers and develop cluster operating systems.  Notable
+attempts include VAX/VMS, MOSIX, Beowulf, Plan 9, ScaleMP, and Hadoop.
 
@@
 GNU/Hurd is an operating system based on the Mach kernel, which was
-designed for use in a cluster environment, making it a natural choice
-to build a POSIX-compatible SSI cluster.  Development on Hurd has
-progressed slowly, but a Debian port has been available since 2013,
-and approximately three quarters of the Debian packages compile and
-run on Hurd.
-
-To support a multi-node Hurd cluster, the libpager used by filesystem
-translators needs to support multiple clients (i.e, multiple kernels
-mmap'ing files).  Hurd's original libpager, written in the mid-1990s,
-only supported a single client, and was thus suitable for single node
-operation.
-
-This is a crucial library.  It provides the cache coherence logic
+designed for use in a cluster environment.  Hurd is designed around a
+set of servers that communicate primarily by remote procedure calls
+mediated by the Mach kernel, making it a natural choice to build a
+POSIX-compatible SSI cluster.  Development on Hurd has progressed
+slowly, but a Debian port has been available since 2013, and
+approximately three quarters of the Debian packages compile and run on
+Hurd.
+
+Hurd needs several enhancements to operate as an SSI cluster.  First,
+Mach messages must be transported over network links, a function
+provided by yet another Hurd server, called \netmsg.  Next,
+distributed shared memory must be implemented, without which we can't
+achieve full POSIX compatibility.  Fortunately, Mach's memory
+management subsystem was designed to support distributed shared
+memory, and Hurd's interface to Mach's memory management subsystem
+is compartmentalized in a C library called \libpager.
+
+\libpager is a crucial library.  It provides the cache coherence logic
 necessary to implement inter-node shared memory, whether disk-based or
-not.  Its use in the tmpfs translator, which is in turn used by
-Hurd's POSIX shared memory implementation
-
-Hurd's netmsg and libpaper have been developed to the point where Hurd
-could be usable as a cluster operating system, though the current
-Debian distribution doesn't use this functionality.  At the time of
-this writing (2019), the major barriers to building a Hurd cluster are
-the 32-bit address space and the lack of SMP support, enhancements
-that are beyond the scope of the work described in this paper.
+not.  It is used, for example, by the filesystem servers to implement
+{\tt mmap()}.  Hurd's original \libpager, written in the mid-1990s,
+only supported a single client (a client being a Mach kernel), and was
+thus suitable for single node operation, but its API was designed to
+eventually support multiple clients.  To support a multi-node Hurd
+cluster, a new \libpager was written to support multiple clients (i.e,
+multiple Mach kernels).  This allows both inter-node {\tt mmap()}
+as well as inter-node POSIX shared memory.
+
+Finally, Hurd's C library needs to be modified to allow cross-node
+forks, and Hurd's authentication mechanism needs to be extended
+as well.
+
+Hurd's \netmsg server and the \libpager library have been developed
+to the point where Hurd could be usable as a cluster operating system,
+though the current Debian distribution doesn't use this functionality.
+At the time of this writing (2019), the major barriers to building a
+Hurd cluster are the 32-bit address space and the lack of SMP support,
+enhancements that are beyond the scope of the work described in this
+paper.
 
@@
 three things: consistency, availability, partitioning.  Right now, we
-sacrifice availability -- if the network partitions, libpager can
+sacrifice availability -- if the network partitions, \libpager can
 deadlock.  If we introduce a timeout on lock requests, then we're
@@
 
+\begin{figure}
+\begin{center}
+
+\tikzset{
+    state/.style={
+        draw, ellipse
+    },
+    >=latex
+}
+
+\def\notm{$\overline{\rm m}$}
+\begin{tikzpicture}[node distance=3cm]
+
+\node (e) [state] {e};
+\node (wait-r) [state, above left=of e] {wait-r};
+\node (wait-w) [state, above right=of e] {wait-w};
+\node (ww) [state, right=of wait-w] {\notm,w,w};
+\node (rw) [state, above right=of wait-w] {\notm,r,w};
+\node (nw) [state, below right=of wait-w] {\notm,n,w};
+\node (nr) [state, left=of wait-r] {\notm,n,r};
+\node (rr) [state, above left=of wait-r] {\notm,r,r};
+\node (nn) [state, below left=of wait-r] {\notm,n,n};
+
+\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(read)} (wait-r);
+\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(write)} (wait-w);
+\draw [->] (wait-r) -- (wait-w);
+\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nr);
+\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rr);
+\draw [->] (wait-r) to[out=45, in=135] node[midway, above]{{\bf data_supply}} node[midway, below]{(write access)} (ww);
+
+\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nw);
+\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rw);
+\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(write access)} (ww);
+
+\draw [->] (wait-r) to[out=-135, in=-135] node[midway]{{\bf data_error}} (e);
+\draw [->] (wait-w) to[out=-45, in=-45] node[midway]{{\bf data_error}} (e);
+
+\end{tikzpicture}
+\end{center}
+\caption{Page-in transitions.  Adapted from figure 6 in [1]}
+\label{Mach memory}
+\end{figure}
+
+Each virtual memory page in the Mach kernel can be in one of fifteen
+different states: empty (e), waiting for either read or write access,
+and twelve different states for pages in memory, depending on
+whether the page has been modified or not, the current access, and the
+desired access.  Figure \ref{Mach memory} shows the transitions
+involved in paging in; [1] uses five similar diagrams to completely
+illustrate all of the Mach memory manager's possible transitions.
+
 Mach's ability to attach out-of-band memory to messages is thus a
@@
 management.  A page is typically supplied to kernel by attaching it to
-a memory_object_data_supply message.  No copying needs be done, as the
+a \modatasupply message.  No copying needs be done, as the
 page is transferred to kernel control simply by adjusting bits in the
@@
 page from the kernel to a user-space memory manager is done by
-attaching it to a Mach message in a similar way.
+attaching it to a Mach message from the kernel.
 
@@
 needed.
+
+Each file corresponds to a separate memory object, and separate memory
+objects get separate control ports in the kernel.  A future
+enhancement could be for the kernel to use a single control port for
+all of its memory objects, allowing \libpager to reduce the number of
+these combination structures, if different files have the same usage
+pattern (a program and some of its shared libraries, for example).
 
@@
 
-\section{The NETMSG Server}
-
-To implement distributed virtual memory in Hurd, we need to transport
-Mach messages across the network.  Several programs have been written
-over the years to transport Mach messages over a network connection;
-all collectively go under the name of the ``netmsg server'', or just
-``netmsg''.
-
-I wrote Hurd's current netmsg server in 2016.  It operates over a
-single unsecured TCP/IP connection and has many limitations, but it
-works well enough to implement shared virtual memory.  In server mode,
-it presents a single node, by default the root of the filesystem, to
-any clients that connect.  In client mode, it acts as a translator,
-allowing a remote server's exported node to be mounted in the
-local filesystem.
-
-Out-of-band memory is simply
-copied across the network connection, then deallocated.
-
-\section{LIBPAGER}
+\section{Introduction to Hurd}
+
+The Free Software Foundation chose Mach in the early 1990s as the
+kernel for what was envisioned as its flagship operating system, Hurd.
+The name is an allusion to the collection of servers required to
+implement the core functionality usually found in an operating system
+kernel, but which are transferred to user space in the Mach design.
+Linux ultimately eclipsed Hurd with a faster development process and a
+simpler, more traditional design that packed functionality into a
+large, monolithic kernel.  Hurd development has continued, however,
+and it now available as a Debian port, though without the bevy of
+features and device drivers that have been packed into Linux over the
+years.
+
+For example, the file system, usually implemented as part of the
+kernel, is mostly decoupled in Hurd.  In principle, a server called
+{\tt pci-arbiter} is responsible for mapping PCI devices, such as
+a SCSI controller, into the memory space of user space programs that
+manage them.  In reality, GNU Mach (the version of Mach used by Hurd)
+contains a significant number of device drivers ripped from an older
+version of the Linux kernel, so we do have some device drivers
+(including our disk controllers) in the kernel.
+
+Aside from reading and writing raw disk blocks, however, the file
+system implementation is completely done in user-space servers, called
+{\it translators} in Hurd nomenclature.  The most important file
+server is {\tt ext2fs}, which interprets and manages an ext2 file
+system.  The program interprets {\tt dir_lookup} RPCs and typically
+responds to them by creating a port (in its own port space) that will
+represent a file handle, then replies to the {\tt dir_lookup} message
+with a send right to the port.  The file handle port handles messages
+like {\tt io_stat}, {\tt io_read}, and {\tt io_write} to carry out
+basic I/O, as well as messages like {\tt file_exec} (if the file is
+executable) and {\tt file_chown} to change the file's ownership, and
+some more exotic messages like {\tt io_async}, which requests that the
+fileserver send messages to the process when I/O is available
+(typically used for sockets rather than files), and {\tt
+  io_reauthenticate}, which tells the file server to drop this port's
+association with a user and perform an authentication exchange.
+
+One of the advantages of this design is that mounting a loopback
+device is a non-privileged operation.  All that is required is to run
+another copy of {\tt ext2fs}, pass it the name of the file containing
+the loopback filesystem, and attach it to a name in the directory
+hierarchy.  Voila!  All future {\tt dir_lookup} calls into that part
+of the filesystem will be relayed to the new {\tt ext2fs}.  Likewise,
+an {\tt iso9660fs} translator is available that allows CD and DVD
+images to be loopback mounted without special permissions.
+
+A great deal of Hurd's POSIX compatible API is implemented in the
+standard C library.  It is here, for example, that {\tt read()} is
+implemented using an {\tt io_read} RPC exchange over a Mach port
+linked to a Hurd server.  In addition to a program's standard threads
+of execution, each program also has a thread, managed by code in the C
+library, that listens on a message port.  POSIX signals, for example,
+are delivered by sending message to this port.  Also, while the Linux
+kernel can be queried to find a lot of information about a process,
+such as the command string used to invoke it, the user and group
+permissions it is running with, its environment, and its open file
+descriptors, among other things, Hurd does not track any of this
+information in the kernel.  In Hurd, much of this information is
+obtained by querying the process directly, using its message port,
+though this has the disadvantage of depending on the correct operation
+of the program.  In particular, it is fairly easy to experience a
+``hung'' {\tt ps} in Hurd, caused by a deadlocked program that is not
+responding to its message port.
+
+A sampling of some major Hurd servers include the following:
+
+\begin{description}
+\item[{\tt exec}]
+
+is responsible for executing programs.  It
+parses the ELF binary format and performs the necessary relocations
+and memory mappings to prepare a new Mach task and make it ready to
+run.  If the program begins with the hash-bang sequence ({\tt \#!}),
+{\tt exec} finds a suitable interpreter to run the program.  Users can
+specify their own {\tt exec} servers, so, for example, if a user
+wanted to use {\tt qemu} to emulate a different architecture, no
+kernel changes would be required, only a custom {\tt exec} server.
+
+\item[{\tt auth}]
+
+implements POSIX authentication based on users and groups.  A Hurd
+process will be passed a send right to the {\tt auth} server when it
+starts, and it is the possession of this send right that allows the
+process access to protected resources.  The naive idea of simply
+passing this send right to a server would grant the server all of the
+process's access rights, so this is not done.  Instead, the process
+creates a new port when it desires to authenticate to a server and
+passes send rights to that port to both the server and the {\tt auth}
+server.  The server then passes its copy of the send right to the {\tt
+  auth} server, which matches the two against each other and confirms
+that the desired access is permitted.  Matching the send rights is
+trivial on a single node, since Mach guarantees that rights to the
+same port will always appear on the same port number.
+
+\item[{\tt proc}]
+
+tracks processes in a POSIX-compatible way.  For
+example, the Mach kernel maintains no sense of a parent/child
+relationship between processes, or any concept of process ownership by
+a user or group.  Killing a process (or sending any signal to a
+process) requires an authentication step to obtain the process's
+message port; this authentication is provided by the {\tt proc}
+server.  Also, some information about a process (like its command name
+and environment) needs to be provided without any authentication to
+allow a program like {\tt ps} to operate; the {\tt proc} server
+provides this functionality, as well.
+
+\item[{\tt pfinet}]
+
+is the Hurd's TCP/IP networking stack.  Networking functions in
+the C library trigger RPC calls to this server.
+
+\end{description}
+
+\section{\netmsg}
+
+Due to its heavy reliance on message passing between user-space
+servers, Hurd is a natural choice to build an SSI cluster, if we can
+transport Mach messages over a network.  Several programs have been
+written over the years to do this; all collectively go under the
+name of the ``netmsg server'', or just \netmsg.
+
+I wrote Hurd's current \netmsg server in 2016.  It operates over a
+single unsecured TCP/IP connection and has many limitations, but works
+well enough to implement distributed shared virtual memory.  In server
+mode, it presents a single node, by default the root of the
+filesystem, to any clients that connect.  In client mode, it acts as a
+translator, allowing a remote server's exported node to be mounted in
+the local filesystem.
+
+EXAMPLE: a \netmsg session
+
+Thus, we have a remote file system, but it's important to realize that
+this remote file system supports full Mach messaging.  Opening a file,
+for example, is done by sending a Mach io_? message that is relayed
+across the network, then received, processed, and answered by the
+remote Hurd server managing that file.  This is in contrast to remote
+file systems like NFS that don't transport Mach messages, and only
+allow a subset of Hurd RPCs which can translated into NFS messages.
+In particular, since Mach virtual memory is implemented using the Mach
+messaging protocol, \netmsg allows a remote file to be {\tt mmap}'ed
+into local memory.
+
+\netmsg works by copying Mach messages almost verbatim across the
+network connection.  Only the Mach port numbers are changed.
+Out-of-band memory is simply copied across the network connection,
+then deallocated.
+
+While usable, this version of \netmsg has many flaws.  Transporting
+all of the Mach messages over a single TCP/IP session provides
+reliable, in-order delivery as guaranteed by Mach, but at the cost of
+serializing all messages between two hosts.  Also, using the system's
+TCP/IP stack requires context switches between \netmsg and the
+TCP/IP server (a user-space program, like so much else in Hurd) for
+every Mach message being relayed.  The current scheme doesn't do a
+good job of handling more than two nodes, because there's no mechanism
+to detect a cycle, which could result in a lot of unnecessary network
+chatter.  Finally, it's become clear that the simplistic means of
+managing ports creates unavoidable race conditions.
+
+For example, consider (again) what happens when node A drops its last
+send right to a port on node B.  Node A sends a NO SENDERS message to
+node B and forgets about the port.  It's possible that node B has sent
+node A another send right for the same port, and that it's in transit
+when node A sends the NO SENDERS message.  In that case, node A would
+actually have an outstanding send right when node B receives the NO
+SENDERS message, which would cause node B to incorrectly drop the send
+right linked to node A.  I can see no way around this except a more
+complicated handshake where the NO SENDERS message from node A to node
+B would include a sequence number allowing node B to determine that it
+hadn't sent any additional send rights, and a reply message from node
+B to node A finalizing the port deallocation.  (what about SEQNOS?)
+
+\section{\libpager}
 
 Hurd's implementation of the Mach memory management protocol is
-implemented in a library called ``libpager''.  The original libpager,
+implemented in a library called ``libpager''.  The original \libpager,
 written in the mid-1990s, only supported a single client (i.e, a
@@
 Its API, however, is flexible enough to allow a more sophisticated,
-multi-client libpager as a drop-in replacement.  For dynamically
+multi-client \libpager as a drop-in replacement.  For dynamically
 linked servers, all that is required is to update the system shared
@@
 is responsible for deallocating it (typically by passing it to the
-kernel).  A common use scenario is that {\tt pager_read_page()}
-reads a page from disk and returns it.  {\tt pager_read_page()}
-can be called from multiple threads to service multiple paging
-requests, and must implement its own thread safety.
+kernel).  A common use scenario is that {\tt pager_read_page()} reads
+a page from disk and returns it.  {\tt pager_read_page()} can be
+called from multiple threads to service multiple paging requests, and
+must implement its own thread safety.  However, it is free to
+serialize itself with a global mutex, and this will not prevent {\tt
+  libpager} from servicing other requests.
 
@@
 it to the kernel, if a new request has arrived in the time required
-for {\tt pager_write_page()} to complete
+for {\tt pager_write_page()} to complete.
 
@@
 How do we figure out which client is the local kernel?  One possible
-solution is for netmsg to listen on /servers/netmsg for an RPC that
+solution is for \netmsg to listen on /servers/netmsg for an RPC that
 accepts a send right and identifies which node currently hosts the
-corresponding receive right.
+corresponding receive right.  Mach 3 Kernel Interfaces defined the
+{\tt norma_port_location_hint} RPC that might be useful.
 
@@
 nodes, or a mechanism is developed to detect when a port is remote,
-figure out which node it (currently) resides on, and then libpager can
+figure out which node it (currently) resides on, and then \libpager can
 fork a process and push it to the other node, invisible to ext2fs
@@
 
-\section{Configuration Options}
-
-Several implementation choices exist, and perhaps should be configuration options.
+\subsection{Configuration Options}
+
+Several \libpager implementation choices exist, and perhaps should be configuration options.
 
@@
 
+  Another possible future enhancement is to track how many calls
+  to \pagerreadpage are outstanding, and transition away from calling
+  this routine if it appears to be overloaded.
+
 \item How to handle network partitions / failed nodes / malfunctioning kernels
@@
 
+\item Are the pager_ functions thread-safe?  Do we call them from
+  another thread before they return?
+
+  {\tt pager_write_page()} is called in a protected manner -- only
+  from a single thread (but not always the same thread), and \libpager
+  waits until it returns before calling it again.
+
+  Both {\tt pager_read_page()} and {\tt pager_unlock_page()} may be
+  called reentrantly while another thread is waiting for another call
+  to finish, but never on the same page.
+
 \end{itemize}
 
-\section{The Pagemap}
+\subsection{The Pagemap}
 
@@
 \libpager needs to maintain a table that tracks the state of each
-page, called the pagemap.  Hurd's original \libpager had a fairly
+page, called the pagemap.  Hurd's original \libpager had a relatively
 simple pagemap.  The pagemap was an array of 16-bit shorts; each page
@@
 
+\begin{figure}[h]
+\begin{mdframed}[backgroundcolor=lightblue]
 \begin{verbatim}
@@
 \end{verbatim}
-
-Four bits are allocated for error code.  (all other errors get collapsed into EIO)
-
-The remaining bits indicate if the page is in core, if we're waiting
-for it to be paged in or out (typically to or from disk).
+\end{mdframed}
+\caption{Pagemap structure in {\it original} {\tt libpager}}
+\end{figure}
+
+Two bits were allocated to each of two error codes.  {\tt ENOSPC} and
+{\tt EDQUOT} were preserved; all other errors get collapsed into {\tt
+  EIO}.  The remaining bits indicate if the page is in core, if we're
+waiting for it to be paged in or out to the backing store.
 
@@
 
-Nor is a separate structure for each page desirable if there are
-a large number of pages.  Since the state of one page is often
-identical to the state of many others, a pagemap of pointers
-suggests itself.
+Nor is a separate structure for each page desirable if there are a
+large number of pages.  We anticipate that many pages (say, all the
+pages in a single shared library) will exist in the same state, so
+there's one structure for each combination of its entries, with
+multiple pages pointing to it.  Pagemaps can be quite large, so a
+pagemap of pointers suggests itself.
 
@@
 };
-
+\end{verbatim}
+\end{mdframed}
+\caption{The new pagemap structure}
+\end{figure}
+
+\begin{verbatim}
 Global structure {
@@
 
-How to maintain the pagemap for a large number of clients?  Each page
-has a pointer to a structure indicating which clients currently hold
-copies of the page.  We anticipate that many pages (say, all the pages
-in a single shared library) will exist in the same state, so there's
-one structure for each combination of its entries, with multiple pages
-pointing to it.  Pagemaps can be potentially large, so a pagemap entry
-should be a single pointer.
-
-Each file corresponds to a separate memory object, and separate memory
-objects get separate control ports in the kernel.  A future
-enhancement could be for the kernel to use a single control port for
-all of its memory objects, allowing libpager to reduce the number of
-these combination structures, if different files have the same usage
-pattern (a program and some of its shared libraries, for example).
-
-Each pagemap entry has an ACCESSLIST and a WAITLIST.  ACCESSLIST is
-unordered; WAITLIST is ordered.  ACCESSLIST is stored sorted.
-
-What happens when either list gets a new client (or loses one)?  We
-have to search in the structures for the new combination, and create a
-new structure if the desired combination doesn't exist yet.
-
-We currently use a C++ std::set and let it do the searching for us.
-
-\begin{comment}
-
-Other possibilities:
-
-  How can we organize the structures?  First, by the number of clients
-  on each list.  We'll always know how many clients are in the
-  ACCESSLIST we're searching for, so if we've got six ACCESSLIST clients
-  and are moving the only client on WAITLIST to ACCESSLIST, we'll search
-  in the seven-client ACCESSLIST, zero-client WAITLIST tables.  We can
-  use a hash table or a tree.
-
-  Because ACCESSLIST is sorted, we can binary search to find an
-  ACCESSLIST that matches the one we want, then linear search forward
-  and backward to find the one we want.
-
-  or... Keep ACCESSLIST and WAITLIST completely separate for each other.
-  Each pagemap entry has an ACCESSLIST pointer and a WAITLIST pointer,
-  as well as some flags, including WRITE-ACCESS, which indicates if
-  we've given out write access to a single client, in which case
-  ACCESSLIST isn't a pointer, but a single client port number.  If
-  WRITE-ACCESS is false, then ACCESSLIST points to a NULL-terminated
-  list of client port numbers.  The disadvantage here is that the
-  pagemap entries require more space (at least two pointers).
-
-  WAITLISTs are ordered and include a flag for each client indicating
-  whether it's waiting for READ access or WRITE access.  A special case,
-  indicated by a flag, could be a single client on a WAITLIST, in which
-  case we don't need a pointer to a WAITLIST, but rather can use the
-  port number instead of the pointer.
-
-The pagemap structures do not count the number of pages pointing to
-them.  If we did that, it would facilitate deallocation when the count
-drops to zero, but would interfere with the "add table" pointer
-optimization described below.
-
-POINTER OPTIMIZATION (UNIMPLEMENTED)
-
-To speed this process, each structure can maintain pointers to
-accelerate this mapping.  Each structure would have an "add table"
-would maintain mappings like "add client 123 -> structure 0x80fadde0".
-If the "add table" doesn't have an entry, then we need to search all
-the structures to find one.  A problem with this is the need to
-invalidate those pointers when a structure is deallocated.  Another
-problem is that it requires the structures to resize when these
-pointers are added. Keep this idea for a future enhancement.
-
-single client pointer optimizations
-  1. no access ->2 ->3
-    pagemap_entry * put_client_1_on_waitlist_for_READ
-    pagemap_entry * put_client_1_on_waitlist_for_WRITE
-  2. waiting for read ->4
-    pagemap_entry * move_first_WAITLIST_client_to_ACCESSLIST
-  3. waiting for write ->5
-    ditto (move_first_WAITLIST_client_to_ACCESSLIST)
-  4. has read ->6
-    ditto (put_client_1_on_WAITLIST_for_WRITE)
-  5. has write
-  6. has read; waiting for write ->5
-    ditto (move_first_WAITLIST_client_to_ACCESSLIST)
-
-PAGEMAP FOR A SMALL NUMBER OF CLIENTS (UNIMPLEMENTED)
-
-I've thought of a possible optimization for small numbers of clients
-that would largely avoid the need for separate pagemap structures, and
-would transition to the unlimited size structures described above when
-more clients connect that it can handle.
-
-First, steal a bit from the pointer by aligning the structure on a
-word boundary.  Then, if the least significant bit is 1, interpret the
-pagemap entry as a bitfield rather than as a pointer to a more
-complicated structure.
-
-A 32 bit field (err... make that a 31 bit field) can handle up to seven
-clients.  The new pagemap entries have an ACCESSLIST, which is an
-unsorted set of clients (those who currently have access), and a
-WAITLIST, which is a sorted list of clients that have requested
-access, and a two boolean flags.
-
-We number our clients 0 through 6, and I'm thinking above moving their
-ports into high port number space to easily identify them.  So, the
-first client on the first memory object would be on port 0x8000000,
-the second client would be on 0x80000001, etc.  The second memory
-object would have its ports on 0x80000008 through 0x8000000f.  So we
-can just strip off the lowest 3 bits to figure out our "client
-number".
-
-We need a paging out flag, and a bit to indicate if the granted access
-is read-only or read-write.  Assume that the {\tt error} field is
-always {\tt KERN_SUCCESS}.  The ACCESSLIST requires one bit per client
-to indicate if access is granted.  The WAITLIST is sorted (client
-requests are processed FIFO) and requires four bits per slot -- three
-to identify the client (all ones indicates an empty slot), and one to
-indicate if the request is for read or write access.  Thus, we need
-seven bits for the ACCESSLIST, two more flag bits (paging out and
-read/write access), one bit to indicate that this isn't a pointer,
-leaving room for five WAITLIST slots.
-
-Once we hit our seventh client, or some oddball condition (like an
-error return other than the expected KERN_SUCCESS, or a sixth client
-on the WAITLIST), we shift to a pointer that points to a more
-complicated, dynamically allocated structure.
-
-Likewise, a 64-bit pointer (wishful thinking at the moment, since Hurd
-is currently a 32-bit system) would allow fifteen clients, with an
-eleven client WAITLIST.  Both the 32-bit and the 64-bit scheme have
-two bits left over, perhaps they could be used to allow EIO, ENOSPC,
-and EDQUOT, in addition to KERN_SUCCESS, since these seem to be the
-most likely error codes.
-
-\end{comment}
-
-ACCESS CONTENTION
-
-Putting a client on an empty queue triggers lock requests to all
-clients in the working set, with the exception of the requesting
-client itself, if it's already got read access and is requesting write
-access.  Once the working set size drops to zero (or one), then the
-original request can be satisfied.
-
-In the meantime, additional clients can be added to the queue.  Each
-client should be allowed access to a page for a period of time before
-a lock request is sent.  Is the easiest way to do this to use the
-timeout parameter on mach_msg?
-
-Current implementation waits until the kernel returns a page, then
-writes it to disk.  If a request for the page comes in before the
-write completes, we wait for the write to finish, then send it right
-back out without needing a read.
-
-The new implementation will be handling pages as they're passed from
-one client to another.  Should they be written out each time they are
-handled?  Not write them at all until they're finally released?  Have
-some kind of timer or counter to write them at a controlled pace?
-Initially, we'll not write them at all until they're finally released,
-which mimics the current behavior.
-
-So when the last client returns a page, we look to see if anything is
-in the wait queue.  If so, the first client(s) in the queue get the
-pages.  If not, we flag that a write is in progress and start it.
-When it completes, we check the queue again to see if anything is now
-waiting.  If so, we service it.  If not, we discard the data and
-notify the user that the page has been evicted.
-
-ERROR HANDLING
+\subsection{Access Contention}
+
+The primary data structures for mediating contention between clients
+are the ACCESSLIST and the WAITLIST in the pagemap.
+
+Putting a client on an empty WAITLIST triggers lock requests to all
+clients on the ACCESSLIST, with the exception of the requesting client
+itself, if it's already got read access and is requesting write
+access.  Once the ACCESSLIST drops to zero (or one, in the exceptional
+case just described), then the original request can be satisfied.
+
+In the meantime, additional clients can be added to the WAITLIST,
+which basically triggers nothing.  The rule is that if the WAITLIST is
+not empty, then we're already trying to revoke access to the clients
+on ACCESSLIST.  Remember, there's currently no guarantee that a client
+will have access for any period of time at all.
+
+When a client with write access returns a page, we look to see if
+anything is on the WAITLIST.  If so, the first client(s) on WAITLIST
+get the pages.  If not, we flag that a write is in progress and start
+it.  When it completes, we check the WAITLIST again to see if anything
+is now waiting.  If so, we service it.  If not, we discard the data
+and notify the user that the page has been evicted.
+
+If multiple clients with READ access attempt to unlock near
+simultaneously, will the data_unlock logic will trigger multiple lock
+requests (a set for each client attempting to unlock)?
+
+initial client goes on WAITLIST only when:
+1. client has requested any access and page is PAGINGOUT, or
+2. client has requested any access and another client has WRITE access
+3. client has requested WRITE access (data request or unlock) and another client has any access
+4. client has requested any access and is waiting for the page to page in
+     READ access requested - only READ clients on ACCESSLIST
+     WRITE access requested - ACCESSLIST empty
+
+\subsection{Error Handling}
 
@@
 
-\section{Pseudocode}
+Handling error on unlock requests is somewhat tricky.  {\it Mach 3
+  Kernel Interfaces} states, in its {\bf NOTES} on {\bf
+  memory_object_lock_request}:
+
+\begin{quote}
+When a running thread requires an access that is currently prohibited,
+the kernel issues a {\bf memory_object_data_unlock} call specifying
+the access required. The memory manager can then use {\bf
+  memory_object_lock_request} to relax its access restrictions on the
+data.
+
+To indicate that an unlock request is invalid (that is, requires
+permission that can never be granted), the memory manager must first
+flush the page. When the kernel requests the data again with the
+higher permission, the memory manager can indicate the error by
+responding with a call to {\bf memory_object_data_error}.
+\end{quote}
+
+Thus, when {\tt pager_unlock_page()} returns an error, we can't simply
+relay that error code in a reply to the kernel.  Instead, we must
+flush the page and wait for a {\bf memory_object_data_request} from
+that client, at which point we return the error code from {\tt
+  pager_unlock_page()} rather than processing the data request
+normally.  This is the function of the NEXTERROR list.  In the
+original \libpager design, it was encoded in the pagemap with two
+bits, since there was only one client, and all errors except ENOSPC
+and EDQUOT were collapsed into EIO.  In the new design, NEXTERROR is
+maintained completely separate from the pagemap, largely due to the
+expectation that it will seldom be used.
+
+For example, consider what happens if client A has read access to
+page, and is thus alone on ACCESSLIST.  Client B now requests read
+access to the same page, so it goes on WAITLIST and we
+call \pagerreadpage.  Before this call completes, client A requests
+write access via an unlock request, so it goes on the WAITLIST.
+Finally, \pagerreadpage returns an error.  What do we do?
+We send an \modataerror to client B, and a lock request to
+client A, with a queued NEXTERROR for client A.
+
+\subsection{Pseudocode}
 
@@
      if not, we're flushing, so add client to WAITLIST, run internal_lock_completed on this page, and return
+     XXX: check to see if client is already on WAITLIST?
   if WAITLIST is not empty, add requesting client to WAITLIST and return
+     XXX: check to see if client is already on WAITLIST?
   if another client has WRITE access (WAITLIST is empty), add requesting client to WAITLIST,
@@
 
-Old libpager would flush everything (no special flags set), then
+Old \libpager would flush everything (no special flags set), then
 destroy the receive right and ignore any final messages.  This created
@@
 
-BUG IN OLD CODE (NOT):
-  client has WRITE access and page is being actively changed
-  pager_write() is relatively slow
-  pager_sync() is called asynchronously three times in rapid succession
-  client returns data three times
-  first m_o_return_data starts a pager_write and sets PAGINGOUT
-  next two m_o_return_data's both see PAGINGOUT and set WRITEWAIT
-  first pager_write finishes, sees WRITEWAIT, and wakes other two threads
-  which one goes first?  coin flip
-  thus, there's a chance that the second and third m_o_return_data's will be processed in reverse order
-  actually, current implementation avoids this problem
-    The second and third m_o_return_data's are both queued by the demuxer and don't
-    get serviced until the first one finishes.  Then they are processed in order.
-
-CAVEAT IN OLD CODE:
-  if we get a write error during a m_o_data_return, and we're flagged PAGEINWAIT because of an outstanding READ request,
-  then the INVALID flag doesn't get set, but after we hand out the data, we won't get it back from the kernel
-  (assuming that notify_on_evict is FALSE and thus pages are not flagged PRECIOUS)
-  then we've dropped data without flagging anything; next read request will read old data from the disk
-
-BUG IN OLD CODE (NOT):
-  page 1 is dirty
-  pager_sync() is called
-  page 1 is returned and begins PAGINGOUT
-  pages 1 and 2 are changed and become dirty
-  pager_return() is called
-  pages 1 and 2 are returned; m_o_data_return sees that page 1 is PAGINGOUT, flags it WRITEWAIT and sleeps
-  page 2 is requested by the kernel
-  page 2 has no flags set, so a pager_read_page() is started, even though more recent data is waiting to be written
-  currently, serialization in demux prevents this problem, too, as the page 2 data_request won't start
-    processing until the earlier data_return's complete
-
-ISSUE:
-  client on WAITLIST could have sent an unlock request and be waiting for a lock, not data_error
-  a unlock request will come from a client on ACCESSLIST (for READ) and will have it put on WAITLIST (for WRITE)
-  there could be an outstanding flush if something is already on WAITLIST for WRITE
-
-RESOLVED ISSUE:
-  data_request logic could cause multiple overlapping calls to pager_read_page() if a bunch of clients
-    request READ access near simultaneously
-
-RACE CONDITION:
-  client requests access
-  client is put on WAITLIST and pager_read_page() begins
-  another client requests contradictory access via m_o_data_request
-  NOT [ we send a lock request to original client, even though it hasn't gotten the data yet! ]
-  NOT [ then pager_read_page() completes, and we send the data to the original client! ]
-  YES [ no messages are sent since WAITLIST is not empty ]
-  YES [ new client is put on the end of WAITLIST ]
-
-RACE CONDITION:
-  client B has READ access
-  client A requests WRITE access (m_o_data_request)
-  lock (flush) request is sent to client B and client A is added to WAITLIST
-  nearly simultaneously, client B requests WRITE access (unlock)
-  do nothing, since client B wouldn't have sent unlock request if it had already received lock request,
-    so it will interpret the lock request as an answer to the unlock request
-
-RACE CONDITION:
-  client A has WRITE access
-  pager_return()  (pager_sync would have kernel_copy = true; pager_flush never returns data)
-  client A is sent a lock request
-  client B requests WRITE access
-  client A is sent another lock request, and client B goes on WAITLIST
-  client A sends back page (kernel_copy = false)
-  client B gets serviced
-  client A sends completions for both lock requests
-  RESULT: pager_write_page() is never called; data just moved from client A to client B
-  ( old code would always call write_page(), then set PAGEINWAIT flag on request, then service request after page out finished)
-
-Q: when could a client on WAITLIST be waiting for an lock, not a data_supply?
-A: when that client already had READ access and sent an unlock request to get WRITE access
-  if a client sent an unlock request, then it already had READ access
-  how could this happen if there's (other) clients on WAITLIST?
-  1. a lock/flush message has already been sent to service the clients on WAITLIST,
-       the client sent the unlock request before processing the lock/flush,
-       so essentially the unlock has already been answered and can be ignored
-     this happens if a client is waiting for WRITE access when other clients had access (of some kind)
-  2. the clients are on WAITLIST because we're waiting for paging (in or out) to complete
-     in this case, we need to send a lock
-
-RACE CONDITION:
-  client A has READ access
-  client B requested READ access, and we start paging in (client B goes on WAITLIST)
-  client A requested WRITE access (via unlock), so it goes on WAITLIST
-  client B's pager_read_page() returned an error, so we want to send data_error to B
-     and lock to client A, with a queued NEXTERROR for client A
-
-RACE CONDITION: (not sure about this - idea is to trigger something very similar to what is matched in above code)
-  client A has WRITE access
-  client B requests READ access
-  lock request is sent to client A
-  client A returns page with data_return, and client B gets the data and goes on ACCESSLIST
-  client B returns the data and is removed from ACCESSLIST
-  client C requests READ access, goes on WAITLIST and a pager_read_page() starts
-  client A's lock completed message is processed, triggering a second pager_read_page(), since ACCESSLIST is empty
-
-RESOLVED ISSUE: if multiple clients with READ access attempt to unlock near simultaneously,
-  data_unlock logic will trigger multiple lock requests (a set for each client attempting to unlock)
-
-ISSUES:
-  before pager_write_page() returns, we can't call pager_read_page() or pager_write_page()
-  after pager_write_page() returns, we might have added additional clients to WAITLIST,
-    but did we add them while waiting for pager_write_page() or were they there before?
-    if ACCESSLIST clients have READ access, and first WAITLIST clients are requested READ access, they are new
-    if ACCESSLIST clients have READ access, and first WAITLIST client is requesting WRITE access,
-      they may or may not be new
-    if ACCESSLIST client has WRITE access, and WAITLIST isn't empty,
-      they may or may not be new
-    if we wait until pager_write_page() returns to send lock requests, we avoid having to figure out
-      if we've already sent lock requests, and introduce some delay to let clients use the data
-
-HOW ABOUT:
-  before calling pager_write_page(), service the first batch of clients on WAITLIST
-     and move them to ACCESSLIST, but send no lock requests
-  after pager_write_page(), service any compatible WAITLIST clients and move them to ACCESSLIST
-  if there are still outstanding WAITLIST clients, send lock requests to everyone on ACCESSLIST
-
-RACE CONDITION:
-  client A has WRITE access
-  pager_return() is called; a lock request is sent to A
-  client B requests access; a lock request is sent to A
-  client A answers the pager_return's lock request with a data_return (kernel_copy = false; dirty = true)
-  data_return code services WAITLIST and sends data to B
-  second lock requests answers and does nothing since A is no longer on ACCESSLIST
-  should be OK
-
-RACE CONDITION:
-  client A has WRITE access
-  pager_sync() is called asynchronously
-  client A syncs with a m_o_data_return, but keeps a copy of the data
-  near simultaneously, client B requests access and libpager requests a flush
-  client A answers with a new m_o_data_return containing data different from the first
-  libpager sees the first m_o_data_return and hands it out to client B
-  RESULT: client B does not have the most recent changes to the data
-  [ ] SOLUTION 1: if m_o_data_return indicates kernal_copy=true, we can't hand it out to other clients
-         PROBLEM: m_o_data_return doesn't indicate if the kernel has released WRITE access,
-                  so this doesn't work if client B is requesting READ access and the lock
-                  request is only to revoke WRITE access
-  [X] SOLUTION 2: when we request a flush, wait for a lock_completed message
-
-LOCK SEQUENCE:
-
-client had READ access
-FLUSH LOCK
-NON-FLUSH LOCK (sync)
-FLUSH LOCK COMPLETED (waits for non-flush lock before processing)
-DATA REQUEST from flushed client for READ access - it's on ACCESSLIST, but has already flushed
-  "if this client is already on ACCESSLIST, print error message and return"
-  if other clients had READ access, pager_read_page() gets called
-NON-FLUSH LOCK COMPLETED
-  only now does the flush get processed
-  if client on ACCESSLIST, stuff happens
-
-client had READ access
-lock_request (FLUSH)
-flush occurs internally to the kernel, but the lock_request requires more processing
-there's another access to the page
-DATA REQUEST from flushed client for READ access - it's on ACCESSLIST, but has already flushed
-  "if this client is already on ACCESSLIST, print error message and return"
-  change to: if this client is already on ACCESSLIST for WRITE access, print error message and return
-                XXX: what if kernel had WRITE access, but page wasn't dirty, so it didn't get returned?
-             if this client is already on ACCESSLIST for READ access, it flushed and is trying to re-aquire access,
-                so check to see if we've got an outstanding lock w/ flush flag set
-                if so, [X] add client to WAITLIST and return (flush code will run later)
-                       [ ] clear flush flag and run flush code now (more complicated but a bit faster)
-                otherwise, remove client from ACCESSLIST and proceed (kernel flushed without being asked)
-  if other clients had READ access, pager_read_page() gets called
-lock_completed message
-  only now does the flush get processed
-  if client on ACCESSLIST, stuff happens
-
-multi-page flush (pager_flush or pager_return)
-lock_request sent to a client on a page where that client has no access
-lock_request processed on that page
-client requests access to that page
-
-multi-page flush (pager_flush or pager_return)
-lock_request sent to a client on a page where that client has no access
-client requests access to that page
-lock_request processed on that page
-
-FLUSHING:
-
-Because of the kernel's ability to proactively flush pages (unless PRECIOUS is specified),
-we can't be certain that a client on a page's ACCESSLIST actually possesses that page.
-All we know for sure is that a client not on ACCESSLIST does not possess the page.
+\subsection{Flushing Pages}
+
+Mach kernels are allowed to proactively flush pages unless the
+``precious'' flag is specified, so we can't be certain that a client
+on a page's ACCESSLIST actually possesses that page.  All we know for
+sure is that a client not on ACCESSLIST does not possess the page.
 
@@
 
-Scenario A
+Consider the following scenario:
+
 \begin{itemize}
@@
 
-Scenario A shows that ACCESSLIST, WAITLIST and a FLUSHING flag aren't enough; we have to track
-which clients requested access during the flush to correctly compute ACCESSLIST at the end.
+This scenario shows that ACCESSLIST, WAITLIST and a FLUSHING flag
+aren't enough; we have to track which clients requested access during
+the flush to correctly compute ACCESSLIST at the end.
 
@@
 Flagging a page precious (so the kernel always returns it to us) isn't that big a deal
-for a local client, because "returning the page" just means flipping bits in the VM tables,
+for a local client, because ``returning the page'' just means flipping bits in the VM tables,
 but it's a lot heavier operation for a remote client, since the page will be transmitted
@@
 
-Flagging all of our pages "precious" is tempting.  It provides us with a mechanism
+Flagging all of our pages ``precious'' is tempting.  It provides us with a mechanism
 for retrieving a local, in-memory page that we want to serve out over the network
@@
 logic, since we don't have to worry about whether the page got returned or not
-when we get the lock completed message - precious pages always get returned.
-
-initial client goes on WAITLIST only when:
-1. client has requested any access and page is PAGINGOUT, or
-2. client has requested any access and another client has WRITE access
-3. client has requested WRITE access (data request or unlock) and another client has any access
-4. client has requested any access and is waiting for the page to page in
-     READ access requested - only READ clients on ACCESSLIST
-     WRITE access requested - ACCESSLIST empty
-
-additional clients go on WAITLIST when they request almost any access
-additional clients DON'T go on WAITLIST when:
-1. we're sending an immediate m_o_data_error
-2. they're trying to unlock and we anticipate that they'll get a flush soon by virtue of being on ACCESSLIST
-
-Theorem: if WAITLIST is not empty, and both ACCESSLIST and WAITLIST contain only READ clients,
-         then we're waiting for a page in to complete
-Proof: ???
-Converse: false; while waiting for a page in, we could get a request for WRITE access
-
-Theorem: there can be at most one client on both ACCESSLIST and WAITLIST simultaneously
-
-Theorem: if anything is on WAITLIST, we've got a flush request outstanding
-
-Theorem: if WAITLIST is empty, we've got no flush requests outstanding
-
-PROBLEM CONDITION:
-client A has READ access
-client B requests READ access
-page in starts; client A on ACCESSLIST; client B goes on WAITLIST (for READ)
-client A requests WRITE access via unlock request
-WAITLIST is not empty, but we're not trying to flush
---- what should happen? ---
-add client A to WAITLIST for WRITE
-wait for page in to finish; client B gets data and moves to ACCESSLIST
-now A and B are on ACCESSLIST with READ access, and A is on WAITLIST for WRITE
-client B is sent a flush request, after an optional delay
-once client B flushes, client A is unlocked and is sent a lock message
-now A is on ACCESSLIST with WRITE access
-
-...vs...
-
-client A has READ access
-client B requests WRITE access
-flush sent to client A; client A on ACCESSLIST; client B goes on WAITLIST (for WRITE)
-client A requests WRITE access via unlock request
---- what should happen? ---
-nothing - client A will process the flush as an answer to its unlock request
-after flush is acknowledged, page in starts
-after page in finishes, client B gets data and moves from WAITLIST to ACCESSLIST
-
-if a client is requesting unlock, then it's already on ACCESSLIST for READ
-if anything is on WAITLIST for WRITE, then we're trying to flush  (right?)
-if the only things on WAITLIST are for READ, then we're waiting for a page in
-
-client A has READ access
-client B requests READ access
-page in starts; client A on ACCESSLIST; client B goes on WAITLIST (for READ)
-client C requests WRITE access via data request
-cleint C gets added to WAITLIST (for WRITE)
-client A requests WRITE access via unlock request
-WAITLIST is not empty, and there's a client on it waiting for WRITE, but we're not trying to flush
---- what should happen? ---
-add client A to WAITLIST for WRITE
-wait for page in to finish; client B gets data and moves to ACCESSLIST
-now A and B are on ACCESSLIST with READ access, and C and A are on WAITLIST for WRITE
-clients A and B are sent a flush request, after an optional delay
-the flush requests answers A's unlock, so it comes off the WAITLIST
-once clients A and B flush, client C gets the data
-client A should re-request WRITE access via data request
-
-... or ...
-
-client A has READ access
-client B requests READ access
-page in starts; client A on ACCESSLIST; client B goes on WAITLIST (for READ)
-client C requests WRITE access via data request
-cleint C gets added to WAITLIST (for WRITE)
-client A requests WRITE access via unlock request
-WAITLIST is not empty, and there's a client on it waiting for WRITE, but we're not trying to flush
---- what should happen? ---
-nothing - client A will soon get a flush
-wait for page in to finish; client B gets data and moves to ACCESSLIST
-now A and B are on ACCESSLIST with READ access, and C and A are on WAITLIST for WRITE
-clients A and B are sent a flush request, after an optional delay
-once clients A and B flush, client C gets the data
-client A should re-request WRITE access via data request
-
-PROBLEM
-
-client A has READ access
-client B requests WRITE access, so a flush request gets sent to client A and client B goes on WAITLIST
-   ACCESSLIST: A(read)   WAITLIST: B(for write)
-client A requests WRITE access (unlock request) and goes on WAITLIST
-client A processes flush and requests WRITE access, so goes on WAITLIST again with a data request
-client A is now on WAITLIST twice
-
-\section{Random Test Code}
-
-I wrote a test program that excercises libpager by making a
+when we get the lock completed message -- precious pages always get returned.
+
+\subsection{Random Test Code}
+
+All kinds of race conditions can be envisioned for \libpager.  Just to
+name a few, multiple updated copies of a page could be received while
+waiting for a single call to \pagerwritepage to complete; a client
+could send an unlock request that crosses paths with a lock request
+generated by \libpager, and the client would then perceive the lock
+request as a response to its unlock request; a client already on the
+WAITLIST could flush and re-request access, going on the WAITLIST
+twice if we're not careful.
+
+I wrote a test program that excercises \libpager by making a
 pseudo-random string of pager requests.  The idea was to run it
@@
 
-Test for BUG 1:
+For example, a possible bug could be triggered by the following
+scenario: A single client has write access and is actively changing
+the page.  {\tt pager_write()} is relatively slow, and {\tt
+  pager_sync()} is called three times in rapid succession.  The client
+returns data three times.  The first data return triggers a {\tt
+  pager_write()}, and the next two data returns are queued waiting for
+the write to complete.  After the first {\tt pager_write()} completes,
+we need to ensure that either the next two writes are either processed
+in order, or the second write is dropped completely.
+
+The test code would test for this scenario by generating the following
+sequence (A is the client; T is the translator):
+
 A2 T4 (the read completes) A5 (the data_supply is processed)
@@
 
+Another possible bug scenario (``bug3''): a single client holds two
+pages with write access.  It returns one (perhaps because \pagersync
+was called) which triggers a call to \pagerwritepage.  While
+\pagerwritepage is running, both pages are returned in a single
+m_o_data_return message.  Since the first page is still writing, we
+sleep to wait for it to finish.  Then the second page is requested
+again.  We {\it should} serve the copy that's queued to write, but if
+we're not careful we'll call \pagerreadpage on the second page.
+
 Test for BUG 3: (one client, two pages A and B)
@@
 
+It was an interesting idea, but didn't quite pan out as well as I had
+hoped.  For one thing, the operations can't be generated with equal
+probability.  Client operation 5 (service a message) has to occur
+fairly frequently, or a queue of pending messages will build up.
+Likewise, translator operation 4 (complete a pending operation), also
+has to occur fairly frequently.  On the other hand, client operation 5
+can't run if there are no pending messages, and translator operation 4
+can't run if there are no pending requests.
+
 \section{POSIX Shared Memory}
@@
 of the filesystem tree.  /run/shm is typically implemented using the
-tmpfs translator, which is linked with libpager.  Thus, a multi-client
-libpager enables the use of inter-node shared memory.  What is
+tmpfs translator, which is linked with \libpager.  Thus, a multi-client
+\libpager enables the use of inter-node shared memory.  What is
 required to make this happen is a shared /run/shm - shared with full
-Mach IPC semantics, not just NFS - so netmsg is required.  unionfs can
+Mach IPC semantics, not just NFS - so \netmsg is required.  unionfs can
 possibly be used to overlay the various /run/shm's on each other.
@@
 
+\section{Blocking Behavior of {\tt mach_msg()}}
+
+An issue with Mach's design caused me no little bit of grief.
+Mach's {\tt mach_msg()} (as I read the documentation),
+allows for non-blocking sends: according to {\it Mach 3
+  Kernel Interfaces}:
+
+\begin{quote}
+MACH_SEND_TIMEOUT
+
+The {\it timeout} argument should specify a maximum time (in
+milliseconds) for the call to block before giving up. If the message
+can’t be queued before the timeout interval elapses, then the call
+returns MACH_SEND_TIMED_OUT. A zero timeout is legitimate.
+\end{quote}
+
+In fact, a call to {\tt mach_msg()} can block indefinitely, even if a
+{\it timeout} value of zero is specified!  The problem occurs when the
+message is an RPC directed at the kernel which in turns triggers other
+RPCs.  The memory manager is a big offender here.  Sending a {\tt
+  vm_map} message to the kernel will cause the kernel to send a
+\moinit message to the memory manager, then block waiting for a reply.
+If both the {\tt vm_map} and the \moinit were handled by a single
+thread in \netmsg, then that thread would block indefinitely,
+requiring considerable additional complexity in \netmsg to ensure that
+additional threads are always available in case calls to {\tt
+  mach_msg()} block.
+
+The consensus on the Hurd mailing list leaned in favor of this being
+legitimate behavior, reading the documentation as guaranteeing that
+the timeout applied to the send, and not to the underlying RPC.  In
+any event, changing this behavior would require quite a bit of work on
+the Mach kernel.
+
 \section{Current Status and Future Work}
 
-Both the netmsg server and the new libpager have been successfully
-tested in a virtual machine running Hurd.
+Both \netmsg and the new \libpager have been successfully
+tested in a virtual machine running Hurd, but additional work is
+required before Hurd can be used as a viable SSI cluster operating
+system.
 
@@
 
-\item improved netmsg
-
-Use a reliable datagram delivery protocol, not TCP.
-
-Direct access to network hardware.  Memory-mapped PCI hardware
-accessed with a library.  This would require a network card
-to be dedicated to netmsg, but with Cisco virtual NICs, that's
-not such a big problem.
+\item \netmsg improvements
+
+\begin{itemize}
+
+\item Use a reliable datagram delivery protocol instead of TCP.
+
+\item Add encryption and authentication.
+
+This might not be necessary if the protocol can be guaranteed to
+operate only over a local network, and might have adverse performance
+impacts, but should be available if desired.
+
+\item Improved port allocation protocol
+
+We need to avoid race conditions when deallocating ports, also also
+detect loops when more than two nodes are interoperating.
+
+\item Direct access to network hardware.
+
+Memory-mapped PCI hardware can be accessed with a library, avoiding a
+context switch to the networking server.  This would require a network
+card to be dedicated to \netmsg, but with networking hardware such as
+Cisco's VIC 1280, that's not a problem, since a VIC 1280 can
+present itself as anywhere from one to sixteen virtual NICs, allowing
+an additional virtual NIC to be allocated solely for use by \netmsg.
+
+\end{itemize}
+
+\item Authentication
+
+As I outlined in
+\href{http://lists.gnu.org/archive/html/bug-hurd/2016-09/msg00012.html}{this
+  proposal}, Hurd's current authentication scheme is inadequate for
+multi-node operation, but a straightforward enhancement is possible.
+
+Currently, if a process wishes to authenticate with a server (say, to
+obtain access to file), it creates a port (used solely for
+authentication) and passes two send rights to that port, one to the
+file server and one to a trusted authentication server.  The file
+server then passes its send right to the authentication server, which
+matches it against the send right passed by the client, verifies that
+the client in question has the desired permissions, and replies to the
+file server.  This allows the client to validate its permissions
+without transferring those permissions to a (potentially malicious)
+file server.
+
+The current authentication scheme relies on Mach's guarantee that
+rights to the same queue will always appear on a single port number,
+no matter how they were received.  This allows the authentication
+server to match the send right that it receives directly from the
+client with the send right that was created by the client, but passed
+via the file server.  The current \netmsg design doesn't honor this
+assumption when more than two nodes are in use.  If the client, file
+server, and authentication server are all on different nodes, then the
+authentication server will receive send rights on two different ports
+and won't be able to match them.  An improved \netmsg design is
+thus essential.
+
+\item C library changes
+
+Mainly allowing cross-node forks.
 

### 2019-05-24 libpager/pagemap.cc
libpager: silence compiler warnings due to int signedness

@@ -320,7 +320,7 @@ void pager::drop_client (mach_port_t control, const char * const reason,
   }
 
   if (some_locks_cleared) {
-    fprintf(stderr, "libpager: warning: dropping client %u (%s) with outstanding locks\n", control, reason);
+    fprintf(stderr, "libpager: warning: dropping client %lu (%s) with outstanding locks\n", control, reason);
   }
 
   // check to see if this client has any outstanding pages
@@ -356,7 +356,7 @@ void pager::drop_client (mach_port_t control, const char * const reason,
   }
 
   if (removed_from_WAITLIST) {
-    fprintf(stderr, "libpager: warning: dropping client %u (%s) with outstanding waits\n", control, reason);
+    fprintf(stderr, "libpager: warning: dropping client %lu (%s) with outstanding waits\n", control, reason);
   }
 
   // pending attribute changes - signal them with "fake" change completed messages
@@ -462,7 +462,7 @@ void pager::lock_object(vm_offset_t OFFSET, vm_size_t LENGTH, int RETURN, bool F
   vm_size_t npages = LENGTH / page_size;
 
   vm_offset_t page = OFFSET / page_size;
-  for (int i = 0; i < npages; i ++, page ++) {
+  for (vm_size_t i = 0; i < npages; i ++, page ++) {
     // std::cerr << pagemap[page];
     for (auto client: pagemap[page]->ACCESSLIST_clients()) {
       if (! only_signal_WRITE_clients || pagemap[page]->get_WRITE_ACCESS_GRANTED()) {
@@ -588,7 +588,7 @@ void pager::internal_lock_completed(memory_object_control_t MEMORY_CONTROL,
   const int UNLOCK = 2;
 
   vm_offset_t page = OFFSET / page_size;
-  for (int i = 0; i < npages; i ++, page ++) {
+  for (vm_size_t i = 0; i < npages; i ++, page ++) {
     operation[i] = 0;
     assert(! pagemap[page]->get_WRITE_ACCESS_GRANTED());
     if (pagemap[page]->is_client_on_ACCESSLIST(MEMORY_CONTROL) && ! pagemap[page]->is_WAITLIST_empty()) {
@@ -615,7 +615,7 @@ void pager::internal_lock_completed(memory_object_control_t MEMORY_CONTROL,
[…]

### 2019-05-24 libpager/pagemap.h

@@ -343,7 +343,7 @@ class pagemap_vector : public std::vector<pagemap_entry>
 {
   public:
 
-  pagemap_entry & operator[](int n)
+  pagemap_entry & operator[](unsigned int n)
   {
     // fprintf(stderr, "[](%d) size=%d\n", n, size());
     if (n >= size()) {
@@ -438,7 +438,7 @@ struct pager {
   // the first data object in struct pager must be struct port_info
   struct port_info port;
 
-  const int page_size = vm_page_size;
+  const vm_size_t page_size = vm_page_size;
 
   bool may_cache;
   memory_object_copy_strategy_t copy_strategy;

### 2019-05-24 libpager/pagemap.cc
minor bug fix in a debugging routine

@@ -54,6 +54,7 @@ std::ostream& operator<< (std::ostream &out, const pagemap_entry::data &entry)
     out << client.client << ",";
   }
   out << "]\n";
+  return out;
 }
 
 /* internal_flush_request() requests a client to flush a page that is needed by

### 2019-05-26 libpager/paper.tex
working on "Building an SSI Cluster" paper

@@
 \usepackage{xspace}
-\usepackage{hyperref}
+\usepackage[hyperfootnotes=false]{hyperref}
 \usepackage{mdframed}
@@
 \def\moinit{{\tt memory_object_init}\xspace}
-\def\modatasupply{{\tt memory_object_data_supply}\xspace}
+\def\moready{{\tt memory_object_ready}\xspace}
+\def\morequest{{\tt memory_object_data_request}\xspace}
+\def\mosupply{{\tt memory_object_data_supply}\xspace}
 \def\modataerror{{\tt memory_object_data_error}\xspace}
 
+%% Disable hypertext links in table of contents
+%% https://tex.stackexchange.com/questions/55190
+
+\makeatletter
+\let\Hy@linktoc\Hy@linktoc@none
+\makeatother
+
 \begin{document}
@@
 \maketitle
+
+\tableofcontents
+
+\vfill\eject
 
@@
     },
+    trans/.style={
+        line width=2pt,
+        transparent!20,
+        ->,
         >=latex
+    },
+    label/.style={
+      midway, align=center,
+      transparent!100
     }
+}
 
 \def\notm{$\overline{\rm m}$}
+
 \begin{tikzpicture}[node distance=3cm]
@@
 
-\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(read)} (wait-r);
-\draw [->] (e) -- node[midway]{{\bf data_request}} node[midway, below]{(write)} (wait-w);
-\draw [->] (wait-r) -- (wait-w);
-\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nr);
-\draw [->] (wait-r) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rr);
-\draw [->] (wait-r) to[out=45, in=135] node[midway, above]{{\bf data_supply}} node[midway, below]{(write access)} (ww);
-
-\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(no access)} (nw);
-\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(read access)} (rw);
-\draw [->] (wait-w) -- node[midway]{{\bf data_supply}} node[midway, below]{(write access)} (ww);
-
-\draw [->] (wait-r) to[out=-135, in=-135] node[midway]{{\bf data_error}} (e);
-\draw [->] (wait-w) to[out=-45, in=-45] node[midway]{{\bf data_error}} (e);
+\draw [trans] (e) -- node[label] {{\bf data_request}\\(read)} (wait-r);
+\draw [trans] (e) -- node[label] {{\bf data_request}\\(write)} (wait-w);
+\draw [trans] (wait-r) -- (wait-w);
+\draw [trans] (wait-r) -- node[label] {{\bf data_supply}\\(no access)} (nr);
+\draw [trans] (wait-r) -- node[label]{{\bf data_supply}\\(read access)} (rr);
+\draw [trans] (wait-r) to[out=45, in=135] node[label]{{\bf data_supply}\\(write access)} (ww);
+
+\draw [trans] (wait-w) -- node[label]{{\bf data_supply}\\(no access)} (nw);
+\draw [trans] (wait-w) -- node[label]{{\bf data_supply}\\(read access)} (rw);
+\draw [trans] (wait-w) -- node[label]{{\bf data_supply}\\(write access)} (ww);
+
+\draw [trans] (wait-r) to[out=-135, in=-135] node[label]{{\bf data_error}} (e);
+\draw [trans] (wait-w) to[out=-45, in=-45] node[label]{{\bf data_error}} (e);
 
@@
 \end{center}
-\caption{Page-in transitions.  Adapted from figure 6 in [1]}
+\caption{Page-in transitions.  Adapted from Figure 6 in \cite{principles}}
 \label{Mach memory}
@@
 Each virtual memory page in the Mach kernel can be in one of fifteen
-different states: empty (e), waiting for either read or write access,
-and twelve different states for pages in memory, depending on
-whether the page has been modified or not, the current access, and the
-desired access.  Figure \ref{Mach memory} shows the transitions
-involved in paging in; [1] uses five similar diagrams to completely
+different states: empty, waiting for either read or write access, and
+twelve different states for pages in memory, depending on whether the
+page has been modified or not, the current access, and the desired
+access.  Figure \ref{Mach memory} shows the transitions involved in
+paging in; \cite{principles} uses five similar diagrams to completely
 illustrate all of the Mach memory manager's possible transitions.
@@
 management.  A page is typically supplied to kernel by attaching it to
-a \modatasupply message.  No copying needs be done, as the
+a \mosupply message.  No copying needs be done, as the
 page is transferred to kernel control simply by adjusting bits in the
@@
 attaching it to a Mach message from the kernel.
+
+For example, consider how Hurd handles an {\tt mmap()} call.  A
+program uses directory lookup calls to obtain a file port, which is a
+send right to a port managed by a file server, and corresponds to a
+standard POSIX file descriptor.  It then uses the {\tt io_map} RPC to
+obtain another send right, also to a port managed by the file server,
+which implements the Mach memory management protocol.  It passes that
+send right to the kernel, using a {\tt vm_map} RPC, indicating that it
+wishes to map that memory object (i.e, the file), into its memory
+space.  The kernel, after a \moinit / \moready handshake with the file
+server, creates a new memory region in the program's address space,
+initially with no physical memory assigned to it, and returns that
+address to the program.
+
+As soon as the program attempts to access this region of memory in any
+way, a page fault is triggered, which causes the kernel to send an
+\morequest message to the file server.  The file server responds by
+reading the required page from disk (if necessary) and passing it to
+the kernel by attaching it to a \mosupply message.  The kernel, now
+managing this region of memory, maps it to the correct page in the
+previously assigned region of memory, and allows the program to
+continue execution.
+
+If the file server had instead replied with \modataerror, the program
+would have received a seg fault instead.  If another program maps the
+same file, the kernel will not request duplicate copies of the pages,
+but will arrange for the physical memory to be shared between the
+processes.  Also, although the kernel now manages the memory, the file
+server can request it back at any time; the model is that the file
+server manages the memory, and the kernel is only caching it.  If the
+kernel runs short on physical memory, it can either discard the page,
+if it is unmodified and if the file server didn't set the ``precious''
+flag, or will return it to the file server.
+
+\tikzstyle{kernel} = [draw, fill=blue!20, rounded corners, minimum width=30em]
+
+\begin{center}
+\begin{tikzpicture}
+
+\task{Filesystem Server}{at (0,0)}{}{
+%\node (memory1) [above left=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
+}
+\task{Program}{at (15 em,0)}{}{
+%\node (memory1) [above right=-0.5, fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
+\node (memory2) [above right=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
+\node [below] at (memory2.north) {\tiny addr};
+}
+
+\node (kernel) [kernel] at (7.5 em, -7 em) {Kernel};
+%%\node (kmemory) [fill=white, minimum size=1.5cm, dashed, thick, draw] at (7.5 em, -10 em) {};
+%%\node [below] at (kernel.north) {Kernel};
+
+\draw [color=red, <-, thick] (2 em,1) -- node [midway, above] {io\_map} (13 em,1);
+\draw [color=red, ->, thick, dotted] (2 em,.9) -- node [midway, below] {\tiny port 153} (13 em,.9);
+
+\draw [color=red, ->, thick] (15 em,-3 em) -- node [midway, left, align=center] {vm\_map\\{\tiny port 153}} (15 em,- 7 em);
+\draw [color=red, <-, thick, dotted] (15.2 em,-3 em) -- node [midway, right] {\tiny addr} (15.2 em,- 7 em);
+
+\draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 7 em);
+\draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 7 em);
+
+%\draw [color=red, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
+%  \draw [color=red, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);
+
+\end{tikzpicture}
+\end{center}
 
@@
 the local filesystem.
-
-EXAMPLE: a \netmsg session
-
-Thus, we have a remote file system, but it's important to realize that
-this remote file system supports full Mach messaging.  Opening a file,
-for example, is done by sending a Mach io_? message that is relayed
-across the network, then received, processed, and answered by the
-remote Hurd server managing that file.  This is in contrast to remote
-file systems like NFS that don't transport Mach messages, and only
-allow a subset of Hurd RPCs which can translated into NFS messages.
-In particular, since Mach virtual memory is implemented using the Mach
-messaging protocol, \netmsg allows a remote file to be {\tt mmap}'ed
-into local memory.
-
 \netmsg works by copying Mach messages almost verbatim across the
@@
 
+\begin{figure}[h]
+\begin{mdframed}
+\begin{verbatim}
+root@qemu-hurd-1:~/github/netmsg# ./netmsg -s ~ &
+[1] 761
+root@qemu-hurd-1:~/github/netmsg# 
+\end{verbatim}
+\end{mdframed}
+
+\begin{mdframed}
+\begin{verbatim}
+root@qemu-hurd-2:~/netmsg/netmsg# settrans -ac node ./netmsg 192.168.55.101
+root@qemu-hurd-2:~/netmsg/netmsg# ls node
+github
+hurd-0.9.git20190331
+hurd_0.9.git20190331-5.debian.tar.bz2
+hurd_0.9.git20190331-5.dsc
+hurd_0.9.git20190331.orig-eth-filter.tar.bz2
+hurd_0.9.git20190331.orig-libddekit.tar.bz2
+hurd_0.9.git20190331.orig-libdde-linux26.tar.bz2
+hurd_0.9.git20190331.orig-libmachdev.tar.bz2
+hurd_0.9.git20190331.orig.tar.bz2
+old
+translator_primer
+root@qemu-hurd-2:~/netmsg/netmsg# 
+\end{verbatim}
+\end{mdframed}
+\caption{A \netmsg session}
+\label{A netmsg session}
+\end{figure}
+
+For example, Figure \ref{A netmsg session} shows a simple \netmsg
+session between two Hurd virtual machines.  On {\tt qemu-hurd-1}, I
+started a \netmsg server ({\tt -s}) sharing {\tt root}'s home
+directory, listening on \netmsg's default TCP port 2345.  Then, on
+{\tt qemu-hurd-2}, I attached an active ({\tt -a}) translator to {\tt
+  node}, creating it ({\tt -c}) if it didn't already exist, running
+\netmsg as a client connecting to {\tt qemu-hurd-1}'s IP address.
+Running {\tt ls node} shows the remote directory.
+
+Thus, we have a remote file system, but it's important to realize that
+this remote file system supports full Mach messaging.  Reading a file,
+for example, is done by sending a Hurd {\tt io_read} message that is
+relayed across the network, then received, processed, and answered by
+the remote Hurd server managing that file.  This is in contrast to
+remote file systems like NFS that don't transport Mach messages, and
+only allow a subset of Hurd RPCs which can translated into NFS
+messages.  In particular, since Mach virtual memory is implemented
+using the Mach messaging protocol, \netmsg allows a remote file to be
+{\tt mmap}'ed into local memory.
+
 While usable, this version of \netmsg has many flaws.  Transporting
@@
 reliable, in-order delivery as guaranteed by Mach, but at the cost of
-serializing all messages between two hosts.  Also, using the system's
-TCP/IP stack requires context switches between \netmsg and the
-TCP/IP server (a user-space program, like so much else in Hurd) for
-every Mach message being relayed.  The current scheme doesn't do a
-good job of handling more than two nodes, because there's no mechanism
-to detect a cycle, which could result in a lot of unnecessary network
-chatter.  Finally, it's become clear that the simplistic means of
-managing ports creates unavoidable race conditions.
-
+serializing all messages, and pausing all communication to wait for a
+TCP re-transmission, when only individual Mach ports need to be
+serialized.  Also, using the system's TCP/IP stack requires context
+switches between \netmsg and the TCP/IP server (a user-space program,
+like so much else in Hurd) for every Mach message being relayed.  The
+current scheme doesn't do a good job of handling more than two nodes,
+because there's no mechanism to detect a cycle, which could result in
+a lot of unnecessary network chatter.  Finally, it's become clear that
+the simplistic means of managing ports creates unavoidable race
+conditions.
+
+% However, once I had a minimally working \netmsg, I moved
+% on the next required component -- \libpager.
+
+If you load a stock\footnote{``Stock'' distributions are available at
+  \href{https://cdimage.debian.org/cdimage/ports/latest/hurd-i386/}{\tt https://cdimage.debian.org/cdimage/ports/latest/hurd-i386/}}
+Hurd distribution and run \netmsg\footnote{\netmsg is available at \href{https://github.com/BrentBaccala/hurd}{\tt https://github.com/BrentBaccala/hurd}} as shown above, you get a remote
+file system that lets you read and write files, but executing remote
+files doesn't work -- the program just hangs.
+That's because Hurd, by default,\footnote{It's not too hard to change
+  this behavior by changing the {\tt exec} server.} executes files
+not by reading them into memory, but by {\tt mmap}'ing them into
+memory.  This requires the {\it remote} file server's memory
+management code to service {\it two} clients -- its local kernel, and
+the remote one that is now trying to {\tt mmap} the file, and a stock
+Hurd file server can't do that.  We need to upgrade {\tt libpager}.
+
+\begin{comment}
 For example, consider (again) what happens when node A drops its last
@@
 B to node A finalizing the port deallocation.  (what about SEQNOS?)
+\end{comment}
 
@@
 Hurd's implementation of the Mach memory management protocol is
-implemented in a library called ``libpager''.  The original \libpager,
+implemented in a library called \libpager.  The original \libpager,
 written in the mid-1990s, only supported a single client (i.e, a
@@
 linked servers, all that is required is to update the system shared
-library to obtain multi-node functionality.
+library to obtain multi-node functionality.  Note, however, that
+Hurd's root file system will almost always be implemented using a
+statically linked server, since it needs to start quite early in
+the boot process.
 
@@
 accepts a send right and identifies which node currently hosts the
-corresponding receive right.  Mach 3 Kernel Interfaces defined the
-{\tt norma_port_location_hint} RPC that might be useful.
+corresponding receive right.  \cite{interfaces} defined the {\tt
+  norma_port_location_hint} RPC that might be useful.
 
@@
 Two bits were allocated to each of two error codes.  {\tt ENOSPC} and
-{\tt EDQUOT} were preserved; all other errors get collapsed into {\tt
-  EIO}.  The remaining bits indicate if the page is in core, if we're
-waiting for it to be paged in or out to the backing store.
+{\tt EDQUOT} were preserved; all other errors got collapsed into {\tt
+  EIO}.  The remaining bits indicated if the page was in core, if we
+were waiting for it to be paged in or out to the backing store, and if
+the last attempt to write the data to the backing store had resulted
+in an error.
 
@@
 
-Handling error on unlock requests is somewhat tricky.  {\it Mach 3
-  Kernel Interfaces} states, in its {\bf NOTES} on {\bf
+Handling error on unlock requests is somewhat tricky.
+\cite{interfaces} states, in its {\bf NOTES} on {\bf
   memory_object_lock_request}:
@@
 
-\subsection{Pseudocode}
+\subsection{Flushing Pages}
+
+Mach kernels are allowed to proactively flush pages unless the
+``precious'' flag is specified, so we can't be certain that a client
+on a page's ACCESSLIST actually possesses that page.  All we know for
+sure is that a client not on ACCESSLIST does not possess the page.
+
+Examining the Mach kernel code for m_o_lock_request, we see that the kernel might block
+while looping over the pages (and unlocks the memory object while doing so), so for a
+multi-page lock request, we need to consider the possibility that part of the page range
+might be processed, then other kernel operations happen (including VM page operations)
+before the remaining part of the page range is processed and the lock request completes.
+
+Consider the following scenario:
+
+\begin{itemize}
+\item five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
+\item pager_flush is called on pages 0-100
+\item page 50 flushes on client A
+\item client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
+\item the flushes complete
+\item only client A now has access to page 50 and should be alone on its ACCESSLIST
+\end{itemize}
+
+This scenario shows that ACCESSLIST, WAITLIST and a FLUSHING flag
+aren't enough; we have to track which clients requested access during
+the flush to correctly compute ACCESSLIST at the end.
+
+PRECIOUS PAGES
+
+Flagging a page precious (so the kernel always returns it to us) isn't that big a deal
+for a local client, because ``returning the page'' just means flipping bits in the VM tables,
+but it's a lot heavier operation for a remote client, since the page will be transmitted
+across the network unnecessarily.
+
+Flagging all of our pages ``precious'' is tempting.  It provides us with a mechanism
+for retrieving a local, in-memory page that we want to serve out over the network
+without having to read it again from disk.  It also simplifies the lock completed
+logic, since we don't have to worry about whether the page got returned or not
+when we get the lock completed message -- precious pages always get returned.
+
+\subsection{Random Test Code}
+
+All kinds of race conditions can be envisioned for \libpager.  Just to
+name a few, multiple updated copies of a page could be received while
+waiting for a single call to \pagerwritepage to complete; a client
+could send an unlock request that crosses paths with a lock request
+generated by \libpager, and the client would then perceive the lock
+request as a response to its unlock request; a client already on the
+WAITLIST could flush and re-request access, going on the WAITLIST
+twice if we're not careful.
+
+I wrote a test program that excercises \libpager by making a
+pseudo-random string of pager requests.  The idea was to run it
+repeatedly using various seed values, then if it detected a bug, it
+could be re-run with the same seed value in order to reproduce the
+bug.  It never actually found any bugs.
+
+There are five possible client operations:
+
+\begin{enumerate}
+\item request READ access
+\item request WRITE access
+\item request unlock READ $\to$ WRITE
+\item flush page
+\item service a message from the memory manager
+\end{enumerate}
+
+If the client has no access to a page, then 1, 2, or 5 are possible.
+If the client has read access, then 3, 4, or 5 are allowed.  With
+write access, 4 or 5 are permissible.
+
+There are four possible translator operations:
+
+\begin{enumerate}
+\item discard request
+\item return request
+\item sync request
+\item complete a pending pager_read, pager_write, or pager_unlock (which one? there could be several)
+\end{enumerate}
+
+1, 2, and 3 can be synchronous or asynchronous.
+
+The test program should (but currently doesn't) check to ensure that
+following conditions are met:
+
+\begin{itemize}
+\item write access to a page is exclusive to a single client
+\item pages get written back in-order (though skips are possible) (partially done)
+\item if a page is modified, nobody ever sees an earlier version of it
+\item all data requests are properly answered with a data supply or a data error (or data unavailable)
+\item all data unlock requests are properly answered with a data lock
+\item if a data lock requests a reply, make sure one is received
+\end{itemize}
+
+For example, a possible bug could be triggered by the following
+scenario: A single client has write access and is actively changing
+the page.  {\tt pager_write()} is relatively slow, and {\tt
+  pager_sync()} is called three times in rapid succession.  The client
+returns data three times.  The first data return triggers a {\tt
+  pager_write()}, and the next two data returns are queued waiting for
+the write to complete.  After the first {\tt pager_write()} completes,
+we need to ensure that either the next two writes are either processed
+in order, or the second write is dropped completely.
+
+The test code would test for this scenario by generating the following
+sequence (A is the client; T is the translator):
+
+A2 T4 (the read completes) A5 (the data_supply is processed)
+    T3 T3 T3 (the syncs) A5 A5 A5 (the syncs are processed) T4 T4 T4 (the writes complete)
+
+client_request_read_access(client, start_page, end_page)
+  ASSERT: page absent in client
+  sends m_o_data_request
+client_request_write_access(client, start_page, end_page)
+  ASSERT: page absent in client
+  sends m_o_data_request
+client_request_unlock(client, start_page, end_page)
+  ASSERT: page present in client with read access
+  sends m_o_data_unlock
+client_flush(client, start_page, end_page)
+  ASSERT: pages present
+  if pages have WRITE access, modify data and data_return
+  if pages have READ access and are precious, data_return
+client_service_message(client)
+  read message
+  data_supply - save pointers, read/write and precious (reply if requested)
+  data_lock - send messages, update pointers, reply if requested
+  data_error - ??
+
+Another possible bug scenario (``bug3''): a single client holds two
+pages with write access.  It returns one (perhaps because \pagersync
+was called) which triggers a call to \pagerwritepage.  While
+\pagerwritepage is running, both pages are returned in a single
+m_o_data_return message.  Since the first page is still writing, we
+sleep to wait for it to finish.  Then the second page is requested
+again.  We {\it should} serve the copy that's queued to write, but if
+we're not careful we'll call \pagerreadpage on the second page.
+
+Test for BUG 3: (one client, two pages A and B)
+  AB2 T4 (the read completes) TA3 A5 (the sync is processed) TAB2 A5 (the sync is processed)
+    B2 (page request) T4 (the read completes)
+    (now we need to close things out and see that data got discarded)
+
+data should always be modified if possible (whenever a client's got WRITE access)
+
+It was an interesting idea, but didn't quite pan out as well as I had
+hoped.  For one thing, the operations can't be generated with equal
+probability.  Client operation 5 (service a message) has to occur
+fairly frequently, or a queue of pending messages will build up.
+Likewise, translator operation 4 (complete a pending operation), also
+has to occur fairly frequently.  On the other hand, client operation 5
+can't run if there are no pending messages, and translator operation 4
+can't run if there are no pending requests.
+
+\section{Distributed Shared Memory}
+
+Two shared memory implementations are available on modern Linux
+systems -- the older System V shared memory API and the newer POSIX
+API.  POSIX shared memory has not yet been implemented for Hurd, but
+System V shared memory is implemented by mapping ``files'',
+corresponding to shared memory keys, in the {\tt /dev/shm} portion of
+the filesystem tree.  Linux uses a similar approach.  Thus, a
+multi-client \libpager enables the use of inter-node shared memory.
+What is required to make this happen is a shared {\tt /dev/shm} --
+shared with full Mach IPC semantics, not just NFS -- so \netmsg is
+required.
+
+% The stock Hurd
+% distribution currently (May 2019) implements {\tt /dev/shm} as a
+% symlink to {\tt tmp}.
+
+Replacing the root filesystem on Hurd requires a reboot, so unless
+you've recompiled the system translators with the new \libpager, it's
+better to mount a new filesystem on {\tt /dev/shm}, using the new
+\libpager.  Ideally, this would done with Hurd's ramdisk translator,
+         {\tt tmpfs}, but {\tt tmpfs} doesn't currently implement
+the \libpager interface, so it's unusable for {\tt /dev/shm}.
+For demonstration purposes, we can create an ext2 filesystem
+and loopback mount it on {\tt /dev/shm}.
+
+\begin{figure}[h]
+\begin{mdframed}
+\begin{verbatim}
+root@qemu-hurd-1:~# dd if=/dev/zero of=shm bs=1M count=10    
+10+0 records in
+10+0 records out
+10485760 bytes (10 MB, 10 MiB) copied, 0.1 s, 105 MB/s
+root@qemu-hurd-1:~# mke2fs shm
+mke2fs 1.44.6 (5-Mar-2019)
+Creating filesystem with 2560 4k blocks and 2560 inodes
+
+Allocating group tables: done                            
+Writing inode tables: done                            
+Writing superblocks and filesystem accounting information: done
+
+root@qemu-hurd-1:~# settrans -ak /dev/shm ~/hurd-0.9.git20190331/build-deb/ext2fs/ext2fs.static shm
+root@qemu-hurd-1:~# ls /dev/shm
+lost+found
+\end{verbatim}
+\end{mdframed}
+\end{figure}
+
+This isn't ideal, because our shared memory is saved to disk, but it
+serves to prove the functionality desired.  We can now start a \netmsg
+server and run a program that allocates a shared memory segment,
+writes a string of data into it, and waits for an acknowledgment
+that it has been read:
+
+\begin{figure}[h]
+\begin{mdframed}
+\begin{verbatim}
+root@qemu-hurd-1:~# ~/github/netmsg/netmsg -s /dev/shm &
+[1] 2533
+root@qemu-hurd-1:~# ~/github/shm/Dave.Marshall/shm_server 
+\end{verbatim}
+\end{mdframed}
+\end{figure}
+
+On another machine, we use \netmsg to connect {\tt /dev/shm} to the
+first machine.  The two machines now share their shared memory space,
+and a client program is able to map the shared memory segment, read
+the string, and acknowledge it.
+
+\begin{figure}[h]
+\begin{mdframed}
+\begin{verbatim}
+root@qemu-hurd-2:~# settrans -ak /dev/shm ./netmsg/netmsg/netmsg 192.168.55.101
+root@qemu-hurd-2:~# ls /dev/shm
+lost+found  sysvshm-162e
+root@qemu-hurd-2:~# ~/shm_client 
+abcdefghijklmnopqrstuvwxyz
+root@qemu-hurd-2:~# 
+\end{verbatim}
+\end{mdframed}
+\end{figure}
+
+, but can also be implemented using the {\tt
+  tmpfs} translator (a ramdisk), which is linked with \libpager.
+
+unionfs can possibly be used to
+overlay the various /run/shm's on each other.
+
+{\tt mmap}, as currently implemented in Hurd's C library, only
+allocates memory on the local node.  It does not seem difficult to
+create a new translator that would service memory allocation requests
+on remote nodes, in a manner similar to Linux's NUMA (Non-Uniform
+Memory Access) support.  Linux's NUMA implementation uses a zonelist
+for memory allocation.  Memory allocation requests are attempted
+locally, and if this fails, then we fall back into the zonelist, which
+gives us a list of other nodes to try for memory allocation.
+
+\section{Distributed Tasks}
+
+In the Hurd architecture, processes are localized to a node, but our
+implementation of distributed shared memory allows cross-node forks
+with full POSIX semantics.  A straightforward modification to Hurd's C
+library would allow fork calls to spawn processes on other nodes, but
+this has not yet been implemented.
+
+Sharing threads between nodes is much more difficult, and seems
+basically impossible to implement using Mach's current design.  The
+main problem is Mach port handling.  Any thread in a task can listen
+for messages on a Mach port, and messages are only delivered once, so
+balancing the need to copy messages to multiple nodes (since any
+thread could be the only one listening to a given port) while
+guaranteeing unique delivery is quite problematic.
+
+This suggests a natural division: processes are local to each node.  A
+multi-threaded program can fork multiple times, and each forked
+process can spawn multiple threads.  The processes are distributed
+between nodes, while each process is local to a node and supports
+multiple threads.  Since we have inter-node, inter-process shared
+memory, almost all of the standard thread synchronization techniques
+(semaphores, mutexes, condition variables) will work unmodified.  Just
+as a typical multi-threaded process should try to avoid thread
+contention on cache lines (typically 64 bytes), a multi-task process
+should try to avoid task contention on memory pages (4 KB on the x86
+architecture).
+
+Since we have a working implementation of distributed shared virtual
+memory, implementing cross-node forks doesn't seem too difficult.
+There is a problem, however.  Mach allows the allocation of memory
+pages unmanaged by any memory manager.  Unmanaged pages are typically
+used for a process's routine data pages, but a cross-node fork would
+require them to become managed in order to police multi-node access.
+
+This suggests the need to attach a memory manager to the unmanaged
+pages to allow them to be distributed across nodes, and this would
+require an enhancement to the Mach kernel itself.  I think that
+a new {\tt memory_object_attach} RPC would be used for this
+purpose.
+
+Once implemented, cross-node forks could be performed with full POSIX
+semantics.  It's interesting that only at this point do we seem
+to need a Mach kernel function that wasn't anticipated by Mach's
+designers in the late 1980s.
+
+The linux scheduler (sched-domains.txt) occasionally balances
+load between scheduling domains.
+
+\begin{comment}
+\section{Blocking Behavior of {\tt mach_msg()}}
+
+An issue with Mach's design caused me no little bit of grief.
+Mach's {\tt mach_msg()} (as I read the documentation),
+allows for non-blocking sends: according to \cite{interfaces}:
+
+\begin{quote}
+MACH_SEND_TIMEOUT
+
+The {\it timeout} argument should specify a maximum time (in
+milliseconds) for the call to block before giving up. If the message
+can’t be queued before the timeout interval elapses, then the call
+returns MACH_SEND_TIMED_OUT. A zero timeout is legitimate.
+\end{quote}
+
+In fact, a call to {\tt mach_msg()} can block indefinitely, even if a
+{\it timeout} value of zero is specified!  The problem occurs when the
+message is an RPC directed at the kernel which in turns triggers other
+RPCs.  The memory manager is a big offender here.  Sending a {\tt
+  vm_map} message to the kernel will cause the kernel to send a
+\moinit message to the memory manager, then block waiting for a reply.
+If both the {\tt vm_map} and the \moinit were handled by a single
+thread in \netmsg, then that thread would block indefinitely,
+requiring considerable additional complexity in \netmsg to ensure that
+additional threads are always available in case calls to {\tt
+  mach_msg()} block.
+
+The consensus on the Hurd mailing list leaned in favor of this being
+legitimate behavior, reading the documentation as guaranteeing that
+the timeout applied to the send, and not to the underlying RPC.  In
+any event, changing this behavior would require quite a bit of work on
+the Mach kernel.
+
+\end{comment}
+
+\section{Current Status and Future Work}
+
+Both \netmsg and the new \libpager have been successfully
+tested in a virtual machine running Hurd, but additional work is
+required before Hurd can be used as a viable SSI cluster operating
+system.
+
+\begin{itemize}
+
+\item 64-bit addressing and SMP
+
+This is the most important work needed on Hurd.  We've got all of the
+major pieces of a working cluster operating system, but nobody cares
+since it can only use a single processor on each node and individual
+processes are limited to a 4 GB address space.
+
+\item \netmsg improvements
+
+\begin{itemize}
+
+\item Use a reliable datagram delivery protocol instead of TCP.
+
+\item Add encryption and authentication.
+
+This might not be necessary if the protocol can be guaranteed to
+operate only over a local network, and might have adverse performance
+impacts, but should be available if desired.
+
+\item Improved port allocation protocol
+
+We need to avoid race conditions when deallocating ports, also also
+detect loops when more than two nodes are interoperating.
+
+\item Direct access to network hardware.
+
+Memory-mapped PCI hardware can be accessed with a library, avoiding a
+context switch to the networking server.  This would require a network
+card to be dedicated to \netmsg, but with networking hardware such as
+Cisco's VIC 1280, that's not a problem, since a VIC 1280 can
+present itself as anywhere from one to sixteen virtual NICs, allowing
+an additional virtual NIC to be allocated solely for use by \netmsg.
+
+\end{itemize}
+
+\item Authentication
+
+As I outlined in
+\href{http://lists.gnu.org/archive/html/bug-hurd/2016-09/msg00012.html}{this
+  proposal}, Hurd's current authentication scheme is inadequate for
+multi-node operation, but a straightforward enhancement is possible.
+
+Currently, if a process wishes to authenticate with a server (say, to
+obtain access to file), it creates a port (used solely for
+authentication) and passes two send rights to that port, one to the
+file server and one to a trusted authentication server.  The file
+server then passes its send right to the authentication server, which
+matches it against the send right passed by the client, verifies that
+the client in question has the desired permissions, and replies to the
+file server.  This allows the client to validate its permissions
+without transferring those permissions to a (potentially malicious)
+file server.
+
+The current authentication scheme relies on Mach's guarantee that
+rights to the same queue will always appear on a single port number,
+no matter how they were received.  This allows the authentication
+server to match the send right that it receives directly from the
+client with the send right that was created by the client, but passed
+via the file server.  The current \netmsg design doesn't honor this
+assumption when more than two nodes are in use.  If the client, file
+server, and authentication server are all on different nodes, then the
+authentication server will receive send rights on two different ports
+and won't be able to match them.  An improved \netmsg design is
+thus essential.
+
+\item C library changes
+
+Mainly allowing cross-node forks.
+
+\item Required Mach kernel enhancement
+
+\begin{itemize}
+
+\item allow memory manager to attach to unmanaged pages
+
+  This is necessary to implement cross-node {\tt fork}'s with full POSIX semantics.
+
+\end{itemize}
+
+\item Optional Mach kernel enhancements
+
+\begin{itemize}
+
+\item m_o_data_return could specify what permissions the kernel still holds on the page
+
+  Then we could easily downgrade from WRITE to READ access without
+  waiting for a lock completed message, or (like the current code),
+  never downgrading from WRITE to READ.
+
+\item m_o_lock_completed message could specify what kind of lock was completed
+
+  This would avoid the need to match reply ports to lock messages
+  unless (like the current code), we just wait for all outstanding
+  locks to complete before continuing with any of them.
+
+\item ability to request a copy of a unmodified, un-precious page from the kernel
+
+  This would allow \libpager to serve out the page to other nodes without reading it again from the disk.
+
+\item ability to notify memory manager when a non-precious page has been evicted
+
+  Over the network, flagging a page ``precious'' is awfully heavy
+  weight just to get notification that a page has been evicted, since
+  a ``precious'' page will be sent back to the memory manager.
+  Knowing that a page has been evicted may eliminate the need for a
+  future lock request.
+
+\item Use a kernel name port common across all memory objects
+
+  This would allow sharing pagemaps between memory objects with the same usage pattern
+  (i.e, kernels A and B both map files C and D)
+
+\item Add the ability for a kernel to respond to a lock request by
+  sending the page directly to another kernel, rather than back to the
+  memory manager.
+
+  This would cut our network utilization roughly in half for pages
+  dominated by contention between nodes.
+
+\end{itemize}
+
+\item distributed filesystem
+
+\end{itemize}
+
+\begin{thebibliography}{9}
+
+\bibitem{interfaces} {\it Mach 3 Kernel Interfaces}.
+
+\bibitem{principles} {\it Mach 3 Kernel Principles}.
+
+\end{thebibliography}
+
+\appendix
+
+\section{\libpager Pseudocode}
 
@@
 
-\subsection{Flushing Pages}
-
-Mach kernels are allowed to proactively flush pages unless the
-``precious'' flag is specified, so we can't be certain that a client
-on a page's ACCESSLIST actually possesses that page.  All we know for
-sure is that a client not on ACCESSLIST does not possess the page.
-
-Examining the Mach kernel code for m_o_lock_request, we see that the kernel might block
-while looping over the pages (and unlocks the memory object while doing so), so for a
-multi-page lock request, we need to consider the possibility that part of the page range
-might be processed, then other kernel operations happen (including VM page operations)
-before the remaining part of the page range is processed and the lock request completes.
-
-Consider the following scenario:
-
-\begin{itemize}
-\item five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
-\item pager_flush is called on pages 0-100
-\item page 50 flushes on client A
-\item client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
-\item the flushes complete
-\item only client A now has access to page 50 and should be alone on its ACCESSLIST
-\end{itemize}
-
-This scenario shows that ACCESSLIST, WAITLIST and a FLUSHING flag
-aren't enough; we have to track which clients requested access during
-the flush to correctly compute ACCESSLIST at the end.
-
-PRECIOUS PAGES
-
-Flagging a page precious (so the kernel always returns it to us) isn't that big a deal
-for a local client, because ``returning the page'' just means flipping bits in the VM tables,
-but it's a lot heavier operation for a remote client, since the page will be transmitted
-across the network unnecessarily.
-
-Flagging all of our pages ``precious'' is tempting.  It provides us with a mechanism
-for retrieving a local, in-memory page that we want to serve out over the network
-without having to read it again from disk.  It also simplifies the lock completed
-logic, since we don't have to worry about whether the page got returned or not
-when we get the lock completed message -- precious pages always get returned.
-
-\subsection{Random Test Code}
-
-All kinds of race conditions can be envisioned for \libpager.  Just to
-name a few, multiple updated copies of a page could be received while
-waiting for a single call to \pagerwritepage to complete; a client
-could send an unlock request that crosses paths with a lock request
-generated by \libpager, and the client would then perceive the lock
-request as a response to its unlock request; a client already on the
-WAITLIST could flush and re-request access, going on the WAITLIST
-twice if we're not careful.
-
-I wrote a test program that excercises \libpager by making a
-pseudo-random string of pager requests.  The idea was to run it
-repeatedly using various seed values, then if it detected a bug, it
-could be re-run with the same seed value in order to reproduce the
-bug.  It never actually found any bugs.
-
-There are five possible client operations:
-
-\begin{enumerate}
-\item request READ access
-\item request WRITE access
-\item request unlock READ $\to$ WRITE
-\item flush page
-\item service a message from the memory manager
-\end{enumerate}
-
-If the client has no access to a page, then 1, 2, or 5 are possible.
-If the client has read access, then 3, 4, or 5 are allowed.  With
-write access, 4 or 5 are permissible.
-
-There are four possible translator operations:
-
-\begin{enumerate}
-\item discard request
-\item return request
-\item sync request
-\item complete a pending pager_read, pager_write, or pager_unlock (which one? there could be several)
-\end{enumerate}
-
-1, 2, and 3 can be synchronous or asynchronous.
-
-The test program should (but currently doesn't) check to ensure that
-following conditions are met:
-
-\begin{itemize}
-\item write access to a page is exclusive to a single client
-\item pages get written back in-order (though skips are possible) (partially done)
-\item if a page is modified, nobody ever sees an earlier version of it
-\item all data requests are properly answered with a data supply or a data error (or data unavailable)
-\item all data unlock requests are properly answered with a data lock
-\item if a data lock requests a reply, make sure one is received
-\end{itemize}
-
-For example, a possible bug could be triggered by the following
-scenario: A single client has write access and is actively changing
-the page.  {\tt pager_write()} is relatively slow, and {\tt
-  pager_sync()} is called three times in rapid succession.  The client
-returns data three times.  The first data return triggers a {\tt
-  pager_write()}, and the next two data returns are queued waiting for
-the write to complete.  After the first {\tt pager_write()} completes,
-we need to ensure that either the next two writes are either processed
-in order, or the second write is dropped completely.
-
-The test code would test for this scenario by generating the following
-sequence (A is the client; T is the translator):
-
-A2 T4 (the read completes) A5 (the data_supply is processed)
-    T3 T3 T3 (the syncs) A5 A5 A5 (the syncs are processed) T4 T4 T4 (the writes complete)
-
-client_request_read_access(client, start_page, end_page)
-  ASSERT: page absent in client
-  sends m_o_data_request
-client_request_write_access(client, start_page, end_page)
-  ASSERT: page absent in client
-  sends m_o_data_request
-client_request_unlock(client, start_page, end_page)
-  ASSERT: page present in client with read access
-  sends m_o_data_unlock
-client_flush(client, start_page, end_page)
-  ASSERT: pages present
-  if pages have WRITE access, modify data and data_return
-  if pages have READ access and are precious, data_return
-client_service_message(client)
-  read message
-  data_supply - save pointers, read/write and precious (reply if requested)
-  data_lock - send messages, update pointers, reply if requested
-  data_error - ??
-
-Another possible bug scenario (``bug3''): a single client holds two
-pages with write access.  It returns one (perhaps because \pagersync
-was called) which triggers a call to \pagerwritepage.  While
-\pagerwritepage is running, both pages are returned in a single
-m_o_data_return message.  Since the first page is still writing, we
-sleep to wait for it to finish.  Then the second page is requested
-again.  We {\it should} serve the copy that's queued to write, but if
-we're not careful we'll call \pagerreadpage on the second page.
-
-Test for BUG 3: (one client, two pages A and B)
-  AB2 T4 (the read completes) TA3 A5 (the sync is processed) TAB2 A5 (the sync is processed)
-    B2 (page request) T4 (the read completes)
-    (now we need to close things out and see that data got discarded)
-
-data should always be modified if possible (whenever a client's got WRITE access)
-
-It was an interesting idea, but didn't quite pan out as well as I had
-hoped.  For one thing, the operations can't be generated with equal
-probability.  Client operation 5 (service a message) has to occur
-fairly frequently, or a queue of pending messages will build up.
-Likewise, translator operation 4 (complete a pending operation), also
-has to occur fairly frequently.  On the other hand, client operation 5
-can't run if there are no pending messages, and translator operation 4
-can't run if there are no pending requests.
-
-\section{POSIX Shared Memory}
-
-Hurd's System V shared memory implementation, in glibc's
-sysdeps/mach/hurd, operates by mapping "files" in the /run/shm portion
-of the filesystem tree.  /run/shm is typically implemented using the
-tmpfs translator, which is linked with \libpager.  Thus, a multi-client
-\libpager enables the use of inter-node shared memory.  What is
-required to make this happen is a shared /run/shm - shared with full
-Mach IPC semantics, not just NFS - so \netmsg is required.  unionfs can
-possibly be used to overlay the various /run/shm's on each other.
-
-\section{Distributed Tasks and Process Migration}
-
-In the Hurd architecture, processes are localized to a node, but our
-implementation of distributed shared memory allows cross-node forks
-with full POSIX semantics.  A straightforward modification to
-Hurd's C library would allow fork calls to spawn processes
-on other nodes.
-
-How can a process share threads between nodes?  Two main problems:
-sharing Mach ports, and sharing memory.  Mach allows the allocation
-of un-backed memory pages.  We need to attach a memory manager
-to the un-backed pages to allow them to be distributed across nodes.
-This requires relatively minor changes to Mach.
-
-This would allow cross-node forks, but not cross-node threads, which
-would be difficult to implement using Mach's current design.  Any
-thread in a task can listen for messages on a Mach port, and messages
-are only delivered once, so balancing the need to copy messages to
-multiple nodes (since any thread could be the only one listening to a
-given port) while guaranteeing unique delivery is quite problematic.
-
-This suggests a natural division: processes are local to each node.  A
-multi-threaded program can fork multiple times, and each forked
-process can spawn multiple threads.  The processes are distributed
-between nodes, while each process is local to a node and supports
-multiple threads.
-
-Linux's NUMA implementation uses a zonelist for memory allocation.
-Memory allocation requests are attempted locally, and if this fails,
-then we fall back into the zonelist, which gives us a list of other
-nodes to try for memory allocation.
-
-The linux scheduler (sched-domains.txt) occasionally balances
-load between scheduling domains.
-
-\section{Blocking Behavior of {\tt mach_msg()}}
-
-An issue with Mach's design caused me no little bit of grief.
-Mach's {\tt mach_msg()} (as I read the documentation),
-allows for non-blocking sends: according to {\it Mach 3
-  Kernel Interfaces}:
-
-\begin{quote}
-MACH_SEND_TIMEOUT
-
-The {\it timeout} argument should specify a maximum time (in
-milliseconds) for the call to block before giving up. If the message
-can’t be queued before the timeout interval elapses, then the call
-returns MACH_SEND_TIMED_OUT. A zero timeout is legitimate.
-\end{quote}
-
-In fact, a call to {\tt mach_msg()} can block indefinitely, even if a
-{\it timeout} value of zero is specified!  The problem occurs when the
-message is an RPC directed at the kernel which in turns triggers other
-RPCs.  The memory manager is a big offender here.  Sending a {\tt
-  vm_map} message to the kernel will cause the kernel to send a
-\moinit message to the memory manager, then block waiting for a reply.
-If both the {\tt vm_map} and the \moinit were handled by a single
-thread in \netmsg, then that thread would block indefinitely,
-requiring considerable additional complexity in \netmsg to ensure that
-additional threads are always available in case calls to {\tt
-  mach_msg()} block.
-
-The consensus on the Hurd mailing list leaned in favor of this being
-legitimate behavior, reading the documentation as guaranteeing that
-the timeout applied to the send, and not to the underlying RPC.  In
-any event, changing this behavior would require quite a bit of work on
-the Mach kernel.
-
-\section{Current Status and Future Work}
-
-Both \netmsg and the new \libpager have been successfully
-tested in a virtual machine running Hurd, but additional work is
-required before Hurd can be used as a viable SSI cluster operating
-system.
-
-\begin{itemize}
-
-\item 64-bit addressing and SMP
-
-This is the most important work needed on Hurd.  We've got all of the
-major pieces of a working cluster operating system, but nobody cares
-since it can only use a single processor on each node and individual
-processes are limited to a 4 GB address space.
-
-\item \netmsg improvements
-
-\begin{itemize}
-
-\item Use a reliable datagram delivery protocol instead of TCP.
-
-\item Add encryption and authentication.
-
-This might not be necessary if the protocol can be guaranteed to
-operate only over a local network, and might have adverse performance
-impacts, but should be available if desired.
-
-\item Improved port allocation protocol
-
-We need to avoid race conditions when deallocating ports, also also
-detect loops when more than two nodes are interoperating.
-
-\item Direct access to network hardware.
-
-Memory-mapped PCI hardware can be accessed with a library, avoiding a
-context switch to the networking server.  This would require a network
-card to be dedicated to \netmsg, but with networking hardware such as
-Cisco's VIC 1280, that's not a problem, since a VIC 1280 can
-present itself as anywhere from one to sixteen virtual NICs, allowing
-an additional virtual NIC to be allocated solely for use by \netmsg.
-
-\end{itemize}
-
-\item Authentication
-
-As I outlined in
-\href{http://lists.gnu.org/archive/html/bug-hurd/2016-09/msg00012.html}{this
-  proposal}, Hurd's current authentication scheme is inadequate for
-multi-node operation, but a straightforward enhancement is possible.
-
-Currently, if a process wishes to authenticate with a server (say, to
-obtain access to file), it creates a port (used solely for
-authentication) and passes two send rights to that port, one to the
-file server and one to a trusted authentication server.  The file
-server then passes its send right to the authentication server, which
-matches it against the send right passed by the client, verifies that
-the client in question has the desired permissions, and replies to the
-file server.  This allows the client to validate its permissions
-without transferring those permissions to a (potentially malicious)
-file server.
-
-The current authentication scheme relies on Mach's guarantee that
-rights to the same queue will always appear on a single port number,
-no matter how they were received.  This allows the authentication
-server to match the send right that it receives directly from the
-client with the send right that was created by the client, but passed
-via the file server.  The current \netmsg design doesn't honor this
-assumption when more than two nodes are in use.  If the client, file
-server, and authentication server are all on different nodes, then the
-authentication server will receive send rights on two different ports
-and won't be able to match them.  An improved \netmsg design is
-thus essential.
-
-\item C library changes
-
-Mainly allowing cross-node forks.
-
-\item Useful Mach kernel changes
-
-\begin{itemize}
-
-\item allow memory manager to attach to unmanaged pages
-
-  This is necessary to implement cross-node {\tt fork}'s with full POSIX semantics.
-
-\item m_o_data_return could specify what permissions the kernel still holds on the page
-
-  Then we could easily downgrade from WRITE to READ access without
-  waiting for a lock completed message, or (like the current code),
-  never downgrading from WRITE to READ.
-
-\item m_o_lock_completed message could specify what kind of lock was completed
-
-  This would avoid the need to match reply ports to lock messages
-  unless (like the current code), we just wait for all outstanding
-  locks to complete before continuing with any of them.
-
-\item ability to request a copy of a unmodified, un-precious page from the kernel
-
-  This would allow \libpager to serve out the page to other nodes without reading it again from the disk.
-
-\item ability to notify memory manager when a non-precious page has been evicted
-
-  Over the network, flagging a page ``precious'' is awfully heavy
-  weight just to get notification that a page has been evicted, since
-  a ``precious'' page will be sent back to the memory manager.
-  Knowing that a page has been evicted may eliminate the need for a
-  future lock request.
-
-\item Use a kernel name port common across all memory objects
-
-  This would allow sharing pagemaps between memory objects with the same usage pattern
-  (i.e, kernels A and B both map files C and D)
-
-\item Add the ability for a kernel to respond to a lock request by
-  sending the page directly to another kernel, rather than back to the
-  memory manager.
-
-  This would cut our network utilization roughly in half for pages
-  dominated by contention between nodes.
-
-\end{itemize}
-
-\item distributed filesystem
-
-\end{itemize}
-
 \end{document}

### 2019-05-27 libpager/paper.tex
update pseudo code for dealing with lock replies before data returns

@@
   for all pages covered by message:
-    [ if this client had WRITE access, print error (should have used m_o_data_return instead of m_o_lock_completed) ]
+    if this client had WRITE access, do nothing (we're waiting for an m_o_data_return)
     if this client isn't on ACCESSLIST, do nothing (m_o_data_return already processed the data)
@@
 m_o_data_return: (kernel returning pages)
-  (this can not be used in liu of a lock completed message for a WRITE,
-     because it doesn't indicate if the kernel still maintains WRITE access)
+  (this can not be used in liu of a lock completed message for revoking WRITE access,
+     because it could be spontaneously generated by the kernel, and
+     doesn't indicate if the kernel still maintains WRITE access)
   npages = length / __vm_page_size;

### 2019-05-31 libpager/paper.tex
close to finishing 'Building an SSI Cluster' paper

@@
 \usepackage{xcolor}
+\usepackage{enumitem}
+\usepackage{amssymb}
 
@@
 \def\pagersync{{\tt pager_sync()}\xspace}
+\def\pagerflush{{\tt pager_flush()}\xspace}
 \def\pagerwritepage{{\tt pager_write_page()}\xspace}
+\def\pagerunlockpage{{\tt pager_unlock_page()}\xspace}
 \def\pagerreadpage{{\tt pager_read_page()}\xspace}
@@
 \def\morequest{{\tt memory_object_data_request}\xspace}
+\def\moreturn{{\tt memory_object_data_return}\xspace}
+\def\molockrequest{{\tt memory_object_lock_request}\xspace}
+\def\molockcompleted{{\tt memory_object_lock_completed}\xspace}
 \def\mosupply{{\tt memory_object_data_supply}\xspace}
@@
 suspended until the page has been supplied to the kernel.
+The current kernel implementation requests pages one at
+a time, as the process attempts to access them, with no attempt
+to anticipate future page accesses.
 
@@
 
+\libpager acts as a translator between the Mach kernel and a
+server program that parses the on-disk structure of a file system.
 On one side, \libpager implements the Mach memory management protocol
@@
 presents an API to the server program, which is responsible for
-producing the pages, as follows:
+producing the pages.  The APIs primary functions are as follows:
 
@@
 
+This routine is called when a page that was previously retrieved
+for read-only access must now be promoted to read-write access.
+
 \item {\tt pager_sync()} and {\tt pager_sync_some()}
@@
 
+\item {\tt pager_return()} and {\tt pager_return_some()}
+
+This routine is similar to {\tt pager_sync()}, but it requests that
+the kernel(s) discard their copies of the pages after returning them.
+Nothing prevents the kernels from immediately re-requesting copies
+of the pages, but those new requests will trigger calls to
+\pagerreadpage.
+
 \item {\tt pager_flush()} and {\tt pager_flush_some()}
 
-discard dirty pages!
-
-\item {\tt pager_return()} and {\tt pager_return_some()}
-
- flush and return dirty (and precious) pages.
-
-\item {\tt pager_get_error()}
+This routine instructs the kernel to discard dirty pages without
+bothering to save them!  Thus, the ``sync'', ``return'', and ``flush''
+routines cover all possible combinations of two binary options:
+``discard pages'' and ``return pages'':
+
+\begin{tabular}{lcc}
+& discard & return \\
+{\tt pager_sync()} & & \checkmark \\
+{\tt pager_return()} & \checkmark & \checkmark \\
+{\tt pager_flush()} & \checkmark &
+\end{tabular}
 
 \end{itemize}
-
-The current kernel implementation seems to request pages one at
-a time, as the process attempts to access them.
-
-If we're servicing a page read-only to multiple clients, where do we
-get the data from?  If our local kernel has a copy, we almost
-certainly want to get it there, since it's just a manner of flipping
-bits in a hardware page table to obtain a new "copy" of the data.
-Otherwise, we probably want to get the data from the local disk, but
-if the disk is very busy and we've got fast network connections, we
-might want to get that data from another node in the cluster.
-
-Mach's current memory object API makes this a lot more difficult.  You
-can't just request that the kernel hand you back an unmodified page,
-except by flagging it "precious".  That's not too bad if the kernel
-and the memory manager are both local on one node, but if you do this
-over the network, you're sending unnecessary copies of the memory
-pages around.
-
-How do we figure out which client is the local kernel?  One possible
-solution is for \netmsg to listen on /servers/netmsg for an RPC that
-accepts a send right and identifies which node currently hosts the
-corresponding receive right.  \cite{interfaces} defined the {\tt
-  norma_port_location_hint} RPC that might be useful.
-
-[[ the kernel uses different control ports for different memory objects ]]
-
-  If the kernel uses the same control port for all of its pager
-  operations, then we wouldn't have to make this query for every memory
-  object; a single ext2fs translator, for example, could globally
-  identify the location of all of the control ports, one for each node,
-  and this information could be shared amongst all of that translator's
-  pager objects.
-
-DISTRIBUTED LIBPAGER.  We could use a shim process on remote nodes, to
-allow requests for an existing page go to a node that holds a copy of
-that page, instead of to the node with the disk.  Two ways I can
-imagine this.  Either ext2fs forks off multiple processes on different
-nodes, or a mechanism is developed to detect when a port is remote,
-figure out which node it (currently) resides on, and then \libpager can
-fork a process and push it to the other node, invisible to ext2fs
-proper.
 
@@
   local kernel has a copy, it's probably faster to ask the kernel for
-  a copy.  We don't have a good way of detecting which client is the
-  local kernel, and the Mach memory management API doesn't provide a
-  way to request a copy of an unmodified, non-``precious'' page
-  from the kernel.
+  a copy, since it's just a manner of flipping bits in a hardware page
+  table to obtain a new ``copy'' of the data.  Otherwise, we probably
+  want to get the data from the local disk, but if the disk is very
+  busy and we've got fast network connections, we might want to get
+  that data from another node in the cluster.
+
+  We don't have a good way of detecting which client is the local
+  kernel, and the Mach memory management API doesn't provide a way to
+  request a copy of an unmodified, non-``precious'' page from the
+  kernel.  You can't just request that the kernel hand you back an
+  unmodified page, except by flagging it ``precious''.  That's not too
+  bad if the kernel and the memory manager are both local on one node,
+  but if you do this over the network, you're sending unnecessary
+  copies of the memory pages around.
+
+  How do we figure out which client is the local kernel?  One possible
+  solution is for \netmsg to listen on {\tt /servers/netmsg} for an
+  RPC that accepts a send right and identifies which node currently
+  hosts the corresponding receive right.  \cite{interfaces} defined
+  the {\tt norma_port_location_hint} RPC that might be useful.
+
+  If the kernel used the same control port for all of its pager
+  operations (it currently doesn't), then we wouldn't have to make
+  this query for every memory object; a single ext2fs translator, for
+  example, could globally identify the location of all of the control
+  ports, one for each node, and this information could be shared
+  amongst all of that translator's pager objects.
 
@@
 The primary data structures for mediating contention between clients
-are the ACCESSLIST and the WAITLIST in the pagemap.
+are the ACCESSLIST and the WAITLIST in the pagemap.  The ACCESSLIST
+can have multiple clients on it, if they all have read-only access.
+Write access can only be issued to one client at a time.
+
+A client requesting access goes on a previously empty WAITLIST only when:
+\begin{enumerate}[noitemsep, itemsep=2pt, topsep=0pt]
+\item The client has requested any access and we have to start a page in.  This happens when
+  \begin{itemize}[label={}]
+  \item READ access requested - only READ clients on ACCESSLIST
+  \item WRITE access requested - ACCESSLIST empty
+  \end{itemize}
+\item The client has requested any access and we're waiting for the page to page out
+\item The client has requested any access and another client (on ACCESSLIST) has WRITE access
+\item The client has requested WRITE access and other clients (on ACCESSLIST) have any access
+\end{enumerate}
 
@@
 clients on the ACCESSLIST, with the exception of the requesting client
-itself, if it's already got read access and is requesting write
-access.  Once the ACCESSLIST drops to zero (or one, in the exceptional
-case just described), then the original request can be satisfied.
+itself, if it's already got read access and is requesting write access
+(an unlock request).  Once the number of clients on the ACCESSLIST
+drops to zero (or one, in the unlock case), then the original request
+can be satisfied.
 
@@
 not empty, then we're already trying to revoke access to the clients
-on ACCESSLIST.  Remember, there's currently no guarantee that a client
-will have access for any period of time at all.
+on ACCESSLIST.  Thus, the wheels have already been set in motion for
+the WAITLIST to be serviced, and we are free to add additional clients
+to the end of the WAITLIST.  Remember, there's currently no guarantee
+that a client will have access for any period of time at all, so
+there's no time period when any clients will be on WAITLIST without an
+attempt being made to revoke existing access.
 
@@
 get the pages.  If not, we flag that a write is in progress and start
-it.  When it completes, we check the WAITLIST again to see if anything
-is now waiting.  If so, we service it.  If not, we discard the data
-and notify the user that the page has been evicted.
-
-If multiple clients with READ access attempt to unlock near
-simultaneously, will the data_unlock logic will trigger multiple lock
-requests (a set for each client attempting to unlock)?
-
-initial client goes on WAITLIST only when:
-1. client has requested any access and page is PAGINGOUT, or
-2. client has requested any access and another client has WRITE access
-3. client has requested WRITE access (data request or unlock) and another client has any access
-4. client has requested any access and is waiting for the page to page in
-     READ access requested - only READ clients on ACCESSLIST
-     WRITE access requested - ACCESSLIST empty
+it by calling \pagerwritepage.  When it completes, we check the
+WAITLIST again to see if anything is now waiting.  If so, we service
+the request.  If not, we discard the data and notify the server (if
+it requested notifications) that the page has been evicted.
 
@@
 on a page's ACCESSLIST actually possesses that page.  All we know for
-sure is that a client not on ACCESSLIST does not possess the page.
-
-Examining the Mach kernel code for m_o_lock_request, we see that the kernel might block
+sure is that a client not on ACCESSLIST does not possess the page.  This
+makes it somewhat difficult to compute ACCESSLIST after a \pagerflush.
+
+Examining the Mach kernel code for \molockrequest, we see that the kernel might block
 while looping over the pages (and unlocks the memory object while doing so), so for a
@@
 \item five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
-\item pager_flush is called on pages 0-100
+\item \pagerflush is called on pages 0-100
 \item page 50 flushes on client A
@@
 
-PRECIOUS PAGES
-
-Flagging a page precious (so the kernel always returns it to us) isn't that big a deal
-for a local client, because ``returning the page'' just means flipping bits in the VM tables,
-but it's a lot heavier operation for a remote client, since the page will be transmitted
-across the network unnecessarily.
-
-Flagging all of our pages ``precious'' is tempting.  It provides us with a mechanism
-for retrieving a local, in-memory page that we want to serve out over the network
-without having to read it again from disk.  It also simplifies the lock completed
-logic, since we don't have to worry about whether the page got returned or not
-when we get the lock completed message -- precious pages always get returned.
+Flagging all of our pages ``precious'' (so the kernel always returns
+it to us) is tempting.  It provides us with a mechanism for retrieving
+a local, in-memory page that we want to serve out over the network
+without having to read it again from disk.  It also simplifies the
+flushing logic above, since we don't have to worry about whether the
+page got returned or not when we get the lock completed message --
+precious pages always get returned.  Flagging a page ``precious''
+isn't that big a deal for a local client, because ``returning the
+page'' just means flipping bits in the VM tables, but it's a lot
+heavier operation for a remote client, since the page will be
+transmitted across the network unnecessarily, so we don't do this.
 
@@
 
-client_request_read_access(client, start_page, end_page)
-  ASSERT: page absent in client
-  sends m_o_data_request
-client_request_write_access(client, start_page, end_page)
-  ASSERT: page absent in client
-  sends m_o_data_request
-client_request_unlock(client, start_page, end_page)
-  ASSERT: page present in client with read access
-  sends m_o_data_unlock
-client_flush(client, start_page, end_page)
-  ASSERT: pages present
-  if pages have WRITE access, modify data and data_return
-  if pages have READ access and are precious, data_return
-client_service_message(client)
-  read message
-  data_supply - save pointers, read/write and precious (reply if requested)
-  data_lock - send messages, update pointers, reply if requested
-  data_error - ??
-
 Another possible bug scenario (``bug3''): a single client holds two
@@
 \pagerwritepage is running, both pages are returned in a single
-m_o_data_return message.  Since the first page is still writing, we
+\moreturn message.  Since the first page is still writing, we
 sleep to wait for it to finish.  Then the second page is requested
@@
 
-Test for BUG 3: (one client, two pages A and B)
+The test code would test for this bug as follows: (one client, two pages A and B)
+
 AB2 T4 (the read completes) TA3 A5 (the sync is processed) TAB2 A5 (the sync is processed)
 B2 (page request) T4 (the read completes)
-    (now we need to close things out and see that data got discarded)
-
-data should always be modified if possible (whenever a client's got WRITE access)
 
@@
 can't run if there are no pending messages, and translator operation 4
-can't run if there are no pending requests.
+can't run if there are no pending requests, so there is some
+synchronization that has to occur.  Finally, I'm not sure that
+\libpager can be fully tested using anything other than actual Mach
+kernels.
 
@@
 
-, but can also be implemented using the {\tt
-  tmpfs} translator (a ramdisk), which is linked with \libpager.
-
-unionfs can possibly be used to
-overlay the various /run/shm's on each other.
+This implementation services all shared memory requests by directing
+them to a single node.  Obviously, this is not ideal.  The {\tt
+  unionfs}\footnote{{\tt unionfs} is not part of the stock Hurd
+  distribution.  Its source is available from
+  \href{https://www.gnu.org/software/hurd/hurd/translator/unionfs.html}{\tt
+    https://www.gnu.org/software/hurd/hurd/translator/unionfs.html}}
+translator allows directories from different nodes to be overlaid on
+top of each other, and could possibly be used to overlay the various
+{\tt /dev/shm}'s on each other.  Then, a new shared memory segment
+would be created on the original process's node, but would immediately
+become visible to all other nodes.  A more sophisticated translator
+could be developed that would dynamically balance shared memory
+management among the nodes.
 
@@
 
-The linux scheduler (sched-domains.txt) occasionally balances
-load between scheduling domains.
-
-\begin{comment}
-\section{Blocking Behavior of {\tt mach_msg()}}
-
-An issue with Mach's design caused me no little bit of grief.
-Mach's {\tt mach_msg()} (as I read the documentation),
-allows for non-blocking sends: according to \cite{interfaces}:
-
-\begin{quote}
-MACH_SEND_TIMEOUT
-
-The {\it timeout} argument should specify a maximum time (in
-milliseconds) for the call to block before giving up. If the message
-can’t be queued before the timeout interval elapses, then the call
-returns MACH_SEND_TIMED_OUT. A zero timeout is legitimate.
-\end{quote}
-
-In fact, a call to {\tt mach_msg()} can block indefinitely, even if a
-{\it timeout} value of zero is specified!  The problem occurs when the
-message is an RPC directed at the kernel which in turns triggers other
-RPCs.  The memory manager is a big offender here.  Sending a {\tt
-  vm_map} message to the kernel will cause the kernel to send a
-\moinit message to the memory manager, then block waiting for a reply.
-If both the {\tt vm_map} and the \moinit were handled by a single
-thread in \netmsg, then that thread would block indefinitely,
-requiring considerable additional complexity in \netmsg to ensure that
-additional threads are always available in case calls to {\tt
-  mach_msg()} block.
-
-The consensus on the Hurd mailing list leaned in favor of this being
-legitimate behavior, reading the documentation as guaranteeing that
-the timeout applied to the send, and not to the underlying RPC.  In
-any event, changing this behavior would require quite a bit of work on
-the Mach kernel.
-
-\end{comment}
+The same mechanisms used for cross-node forks could also be used to
+permit process migration.  After using cross-node memory maps to
+create a duplicate process on another node (without any threads
+running), the threads of the original process would be stopped, the
+Mach port rights transferred using \netmsg, duplicate threads would be
+created and started.
+
+Some investigation would be required to see if Mach can cleanly
+transfer thread state like this.  For example, I'm not sure how Mach
+would save state for a thread currently blocked in a {\tt mach_msg()}
+system call.  I think that it saves the user-space state with the
+instruction pointer set to the system call instruction.  Restarting
+from this state would rerun the system call, which would create
+duplicate messages if, for example, the system call was the {\tt
+  mach_msg_send()} variant.
 
@@
 Both \netmsg and the new \libpager have been successfully
-tested in a virtual machine running Hurd, but additional work is
+tested in a virtual machine running Hurd, but their operation
+is unreliable and additional work is
 required before Hurd can be used as a viable SSI cluster operating
@@
 an additional virtual NIC to be allocated solely for use by \netmsg.
+
+\end{itemize}
+
+\item Other translator improvements
+
+\begin{itemize}
+
+\item \libpager integration with {\tt tmpfs}
+
+{\tt tmpfs}, the Hurd ramdisk, currently throws {\tt SIGABRT} if any
+\libpager functions are called.
+Since {\tt tmpfs} is the most obvious choice to mount on
+{\tt /dev/shm}, fixing this is a high-priority, low-difficulty task.
+
+\item distributed filesystem
+
+Once we have the ability to run processes using distributed
+shared memory and distributed scheduling, we should enhance our file
+system translators to support replicated copies of files.  This
+suggests the need for a...
+
+\item distributed \libpager
+
+We could use a shim process on remote nodes, to allow requests for an
+existing page go to a node that holds a copy of that page, instead of
+to the node with the disk.  Two ways I can imagine this.  Either
+ext2fs forks off multiple processes on different nodes, or a mechanism
+is developed to detect when a port is remote, figure out which node it
+(currently) resides on, and then \libpager can fork a process and push
+it to the other node, invisible to ext2fs proper.
 
@@
 
-\item m_o_data_return could specify what permissions the kernel still holds on the page
+\item \moreturn could specify what permissions the kernel still holds on the page
 
@@
 
-\item m_o_lock_completed message could specify what kind of lock was completed
+\item \molockcompleted message could specify what kind of lock was completed
 
@@
 
-\item distributed filesystem
-
 \end{itemize}
@@
 \section{\libpager Pseudocode}
+
+\small
 
@@
 
-SHUTDOWN CODE
-
-Old \libpager would flush everything (no special flags set), then
-destroy the receive right and ignore any final messages.  This created
-a minor race condition where clients could request a page while the
-flush was processing, which would then get dropped.
-
 \end{document}

### 2019-05-31 libpager/paper.tex
finishing up "Building an SSI cluster on Hurd"

@@
 At the time of this writing (2019), the major barriers to building a
-Hurd cluster are the 32-bit address space and the lack of SMP support,
-enhancements that are beyond the scope of the work described in this
-paper.
-
-The CAP theorem states that a distributed system can't have all of
-three things: consistency, availability, partitioning.  Right now, we
-sacrifice availability -- if the network partitions, \libpager can
-deadlock.  If we introduce a timeout on lock requests, then we're
-sacrificing consistency.
+Hurd cluster are the 32-bit address space and the lack of SMP support.
 
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \begin{center}
@@
 advertise its services in Hurd is for it to listen for messages on a
-port that it attachs to a name in the file system.
+port that it attaches to a name in the file system.
 
@@
 
-For example, consider how Hurd handles an {\tt mmap()} call.  A
-program uses directory lookup calls to obtain a file port, which is a
-send right to a port managed by a file server, and corresponds to a
-standard POSIX file descriptor.  It then uses the {\tt io_map} RPC to
-obtain another send right, also to a port managed by the file server,
-which implements the Mach memory management protocol.  It passes that
-send right to the kernel, using a {\tt vm_map} RPC, indicating that it
+For example, consider how Hurd handles an {\tt mmap()} call (Figure
+\ref{Initial mapping of a Mach memory object}).  A program uses
+directory lookup calls to obtain a file port, which is a send right to
+a port managed by a file server, and corresponds to a standard POSIX
+file descriptor.  It then uses the {\tt io_map} RPC to obtain another
+send right, also to a port managed by the file server, which
+implements the Mach memory management protocol.  It passes that send
+right to the kernel, using a {\tt vm_map} RPC, indicating that it
 wishes to map that memory object (i.e, the file), into its memory
@@
 As soon as the program attempts to access this region of memory in any
-way, a page fault is triggered, which causes the kernel to send an
-\morequest message to the file server.  The file server responds by
-reading the required page from disk (if necessary) and passing it to
-the kernel by attaching it to a \mosupply message.  The kernel, now
-managing this region of memory, maps it to the correct page in the
-previously assigned region of memory, and allows the program to
-continue execution.
+way (Figure \ref{Initial access of a Mach memory object}), a page
+fault is triggered, which causes the kernel to send an \morequest
+message to the file server.  The file server responds by reading the
+required page from disk (if necessary) and passing it to the kernel by
+attaching it to a \mosupply message.  The kernel, now managing this
+region of memory, maps it to the correct page in the previously
+assigned region of memory, and allows the program to continue
+execution.
 
@@
 
+\tikzstyle{message} = []
+
+\begin{figure}
 \begin{center}
@@
 
-\draw [color=red, <-, thick] (2 em,1) -- node [midway, above] {io\_map} (13 em,1);
-\draw [color=red, ->, thick, dotted] (2 em,.9) -- node [midway, below] {\tiny port 153} (13 em,.9);
-
-\draw [color=red, ->, thick] (15 em,-3 em) -- node [midway, left, align=center] {vm\_map\\{\tiny port 153}} (15 em,- 7 em);
-\draw [color=red, <-, thick, dotted] (15.2 em,-3 em) -- node [midway, right] {\tiny addr} (15.2 em,- 7 em);
-
-\draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 7 em);
-\draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 7 em);
-
-%\draw [color=red, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
-%  \draw [color=red, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);
+\draw [message, <-, thick] (2 em,1) -- node [midway, above] {io\_map} (13 em,1);
+\draw [message, ->, thick] (2 em,.9) -- node [midway, below] {\tiny port 153} (13 em,.9);
+
+\draw [message, ->, thick] (15 em,-3 em) -- (15 em,- 7 em);
+\draw [message, <-, thick] (15.2 em,-3 em) -- node [midway, right] {\tiny addr} (15.2 em,- 6 em);
+
+\node [message, left, align=center] at (15 em, - 4 em) {vm\_map\\{\tiny port 153}};
+
+\draw [message, <-, thick] (-2 em,-3 em) -- (-2 em,- 7 em);
+\draw [message, ->, thick] (-1.5 em,-3 em) -- (-1.5 em,- 7 em);
+
+\node [message, left] at (-2 em, -5.6 em) {m\_o\_init};
+\node [message, right] at (-1.5 em, -5.6 em) {m_o_ready};
+
+\node (port) at (-2 em, -1.5 em) [minimum size=1.5cm, draw, thick] {Port 153};
+
+%\draw [message, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
+%  \draw [message, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);
 
@@
 \end{center}
+\caption{Initial mapping of a Mach memory object}
+\label{Initial mapping of a Mach memory object}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}
+
+\task{Filesystem Server}{at (0,0)}{}{
+%\node (memory1) [above left=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
+}
+\task{Program}{at (15 em,0)}{}{
+%\node (memory1) [above right=-0.5, fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
+\node (memory2) [above right=-0.5, fill=white, minimum size=1.5cm, dashed, thick, draw] {};
+\node [below] at (memory2.north) {\tiny addr};
+}
+
+\node (kernel) [kernel] at (7.5 em, -7 em) {Kernel};
+%%\node (kmemory) [fill=white, minimum size=1.5cm, dashed, thick, draw] at (7.5 em, -10 em) {};
+%%\node [below] at (kernel.north) {Kernel};
+
+\draw [message, <-, thick] (-2 em,-3 em) -- (-2 em,- 7 em);
+\draw [message, ->, thick] (-1.5 em,-3 em) -- (-1.5 em,- 7 em);
+
+\node [message, left] at (-2 em, -5.6 em) {m\_o\_data_request};
+\node [message, right] at (-1.5 em, -5.6 em) {m_o_data_supply};
+
+%%\draw [message, ->, thick, dotted] (memory2.south) -- node [midway, right] {fault} (memory2.south |- kernel);
+
+\node (port) at (-2 em, -1.5 em) [minimum size=1.5cm, draw, thick] {Port 153};
+
+\node (instr) [right, align=center, font=\tiny] at (Program -| Program.west) {INSTR\\ READ\\ INSTR};
+\node [draw] (location) at (Program) {};
+\draw [->] (instr.east) -- (location.west);
+
+\draw [color=red, ->, thick, dotted] (location.south) -- node [midway, right] {fault} (location.south |- kernel);
+
+%\draw [message, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
+%  \draw [message, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);
+
+\end{tikzpicture}
+\end{center}
+\caption{Initial access of a Mach memory object}
+\label{Initial access of a Mach memory object}
+\end{figure}
 
@@
 
-Several \libpager implementation choices exist, and perhaps should be configuration options.
+Several \libpager implementation choices exist, and perhaps should be configuration options,
+though they are currently hard-wired into the code.
 
@@
   is only an issue if system load is high enough to prevent that.
+
+\item How to handle network partitions / failed nodes / malfunctioning kernels
+
+  The CAP theorem states that a distributed system can't have all of
+  three things: consistency, availability, partitioning.  Right now,
+  we sacrifice availability -- if the network partitions, \libpager
+  can deadlock.  If we introduce a timeout on lock requests, then
+  we're sacrificing consistency.
+
+  Currently we deadlock if a client doesn't respond; perhaps we should
+  timeout while waiting for a client?  How to pick the timeout value?
 
@@
   this routine if it appears to be overloaded.
-
-\item How to handle network partitions / failed nodes / malfunctioning kernels
-
-  Currently we deadlock if a client doesn't respond; perhaps we should
-  timeout while waiting for a client?  How to pick the timeout value?
 
@@
 
-\begin{verbatim}
-Global structure {
-   /* NEXTERROR list
-    * A linked list of outstanding (error code, page, client) tuples.
-    * They'll be transmitted to the client that requested them.
-    */
-
-   /* WRITEWAIT list
-    * A linked list of pages waiting to be written to backing store
-    */
-
-    std::map<std::pair<vm_offset_t, vm_size_t>, outstanding_lock> outstanding_locks;
-
-    std::set<outstanding_change_request> outstanding_change_requests;
-}
-\end{verbatim}
+The pagemap is organized as an array with one entry for each virtual
+memory page in the memory object.  In addition to the pagemap, there
+are several additional data structures maintained globally (i.e, for
+the entire memory object):
+
+\begin{itemize}[noitemsep]
+\item The NEXTERROR list, a linked list of outstanding (error code, page, client) tuples
+\item A list of pages waiting to be written to backing store
+\item A list of outstanding lock requests
+\item A list of outstanding change requests
+\end{itemize}
 
@@
 
-Another possible bug scenario (``bug3''): a single client holds two
+Another possible bug scenario runs like this: a single client holds two
 pages with write access.  It returns one (perhaps because \pagersync

### 2019-06-03 doc/cua.tex
reorganize all docs into a 'doc' directory

--- 2018-01-21 slides/cua.tex
@@
 \begin{tikzpicture}
-  [grow via three points={one child at (0.5,-0.7) and
+  [node distance = 1em, grow via three points={one child at (0.5,-0.7) and
   two children at (0.5,-0.7) and (0.5,-1.4)},
@@
 \task{ext2 File Server}{at (20 em,-5 em)}{l}{}
-\task{httpfs}{at (20 em, -12 em)}{ls}{}
-\task{pfinet}{at (20 em, -17 em)}{ls}{}
+\task [below = of ext2 File Server] {httpfs}{}{ls}{}
+\task [below = of httpfs] {pfinet}{}{ls}{}
 
@@
 
-\draw [link] (http) -- +(2,0) |- (httpfs Port 0);
+\draw [link] (http) -- (httpfs Port 1);
 \draw [link] (inet) -- (pfinet Port -1);

### 2019-06-03 doc/misc.tex
reorganize all docs into a 'doc' directory

\documentclass{article}

\usepackage{nopageno}
\usepackage{vmargin}
\usepackage{comment}
\usepackage{xspace}
\usepackage[hyperfootnotes=false]{hyperref}
\usepackage{mdframed}
\usepackage{xcolor}

\definecolor{lightblue}{rgb}{0.8,0.8,1.0}

\usepackage{../slides/hurd}
%\usepackage{tikz}
\usepackage{tkz-graph}
\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes, arrows.meta}

% I like smaller margins than the default
\setmargnohfrb{0.5in}{0.5in}{0.5in}{0.5in}

% I prefer unindented paragraphs with a blank line between them
\parindent 0pt
\parskip \baselineskip

% Redefine underscore to print as a normal character
\catcode`\_=\active
\def_{\_}

\title{Miscellaneous Notes on GNU/Hurd}
\author{Brent Baccala}
\date{May 2019}

\pdfinfo{
  /Title	(Miscellaneous Notes on GNU/Hurd)
  /Author	(Brent Baccala)
}

% Shortcuts just for this document
\def\libpager{{\tt libpager}\xspace}
\def\rpctrace{{\tt rpctrace}\xspace}
\def\netmsg{{\tt netmsg}\xspace}
\def\pagersync{{\tt pager_sync()}\xspace}
\def\pagerwritepage{{\tt pager_write_page()}\xspace}
\def\pagerreadpage{{\tt pager_read_page()}\xspace}
\def\moinit{{\tt memory_object_init}\xspace}
\def\moready{{\tt memory_object_ready}\xspace}
\def\morequest{{\tt memory_object_data_request}\xspace}
\def\mosupply{{\tt memory_object_data_supply}\xspace}
\def\modataerror{{\tt memory_object_data_error}\xspace}

%% Disable hypertext links in table of contents
%% https://tex.stackexchange.com/questions/55190

\makeatletter
\let\Hy@linktoc\Hy@linktoc@none
\makeatother

\begin{document}

\maketitle

\tableofcontents

\vfill\eject

\section{Blocking Behavior of {\tt mach_msg()}}

An issue with Mach's design caused me no little bit of grief.
Mach's {\tt mach_msg()} (as I read the documentation),
allows for non-blocking sends: according to \cite{interfaces}:

\begin{quote}
MACH_SEND_TIMEOUT

The {\it timeout} argument should specify a maximum time (in
milliseconds) for the call to block before giving up. If the message
can’t be queued before the timeout interval elapses, then the call
returns MACH_SEND_TIMED_OUT. A zero timeout is legitimate.
\end{quote}

In fact, a call to {\tt mach_msg()} can block indefinitely, even if a
{\it timeout} value of zero is specified!  The problem occurs when the
message is an RPC directed at the kernel which in turns triggers other
RPCs.  The memory manager is a big offender here.  Sending a {\tt
  vm_map} message to the kernel will cause the kernel to send a
\moinit message to the memory manager, then block waiting for a reply.
If both the {\tt vm_map} and the \moinit were handled by a single
thread in \netmsg, then that thread would block indefinitely,
requiring considerable additional complexity in \netmsg to ensure that
additional threads are always available in case calls to {\tt
  mach_msg()} block.

The consensus on the Hurd mailing list leaned in favor of this being
legitimate behavior, reading the documentation as guaranteeing that
the timeout applied to the send, and not to the underlying RPC.  In
any event, changing this behavior would require quite a bit of work on
the Mach kernel.

\section{\rpctrace}

\rpctrace is Hurd's version of {\tt strace}.  It works by wrapping
around all of a process's ports, intercepting all Mach messages send
to and from that process, and printing them out.

It's got some problems, starting with the difficulty of attaching
and detaching running processes.

Attaching -- if the process has a receive right, we want to move the
receive right to \rpctrace (all of the send rights remain attached to
the same queue) and insert a new receive right in the process.  If one
of the process's threads is blocked trying to receive a message on
that receive right at the moment of \rpctrace attachment, the {\tt
  mach_msg} call will return with an {\tt MACH_RCV_PORT_DIED} error.
Portsets don't cause this kind of problem, since we can move receive
rights in and out of them without triggering errors returns from
{\tt mach_msg}.

Detaching -- there's a race condition here.  Could use a new Mach
primitive to combine two queues together.

\section{New Mach System Calls}

\begin{itemize}

\item {\tt mach_port_merge_ports}

Given a send right and a receive right to two different ports, collapse
them together so that the receive right and everything that was in its
queue goes to the send right's port, as if all messages on the receive
right's port were atomically removed and sent to the send right.

\item {\tt mach_port_split_port}

Given a receive right, split it, returning a new receive right such
that all of the old port's send rights now point to the new receive
right, but any {\tt mach_msg()} calls that were in progress on the
receive right continue un-interrupted.  Messages in the old port's
queue go to the new receive right.  In short, we just move the old
port/queue to a new name, but continue any {\tt mach_msg()} calls
uninterrupted.

\item {\tt thread_get_state} and {\tt thread_set_state}

How to handle thread state when the process is in a system call?

We want to be able to save state well enough to re-create the
thread, maybe on a different node, and restart from the point
it was suspended.

Just re-running the system call would result in messages being
transmitted twice.

\end{itemize}

\begin{thebibliography}{9}

\bibitem{interfaces} {\it Mach 3 Kernel Interfaces}.

\bibitem{principles} {\it Mach 3 Kernel Principles}.

\end{thebibliography}

\end{document}

### 2019-06-03 doc/rpctrace.tex
some old changes to 'rpctrace.tex' that I don't remember much about

--- 2018-01-29 slides/rpctrace.tex
@@
 
-\task {rpctrace}{}{lr}{}
+\useasboundingbox (-6,-5) rectangle (6.5,5.5);
+\scope[transform canvas={scale=0.8}]
 
-\task [right = of rpctrace] {Traced Task 2}{}{sl}{}
-\task [above = of Traced Task 2] {Traced Task 1}{}{sl}{}
+\task {rpctrace}{}{blr}{}
+
+\task [above right = of rpctrace.north east] {Traced Task 1}{}{sl}{}
+\task [below = of Traced Task 1] {Traced Task 2}{}{sl}{}
 \task [below = of Traced Task 2] {Traced Task 3}{}{sl}{}
+\task [below = of Traced Task 3] {Traced Task 4}{}{sl}{}
 
-\task [above = of rpctrace] {Unknown Task}{}{sl}{}
+\task [left = of rpctrace] {Unknown Task}{}{r}{
+}
 
-\draw [port link] (rpctrace Left Port 0) -- +(-1,0) |- (Unknown Task Port 0);
-\draw [port link] (Traced Task 2 Port 0) -- (rpctrace Right Port 0);
-\draw [port link] (Traced Task 1 Port 0) -| ($(Traced Task 1 Port 0)!.5!(rpctrace Right Port 5)$) |- (rpctrace Right Port 5);
+\node [message] at (Unknown Task) {};
+
+\draw [port link] (rpctrace Left Port 0) -- (Unknown Task Port 7);
+\draw [port link] (Traced Task 1 Port 0) -| ($(Traced Task 1 Port 0)!.6!(rpctrace Right Port 5)$) |- (rpctrace Right Port 5);
+\draw [port link] (Traced Task 2 Port 0) -| ($(Traced Task 2 Port 0)!.4!(rpctrace Right Port 0)$) |- (rpctrace Right Port 0);
 \draw [port link] (Traced Task 3 Port 0) -| ($(Traced Task 3 Port 0)!.5!(rpctrace Right Port -5)$) |- (rpctrace Right Port -5);
@@
 
-\node (Unknown Task recvright) at (Unknown Task Port 0) [recv, anchor=west] {R};
+\node (Task 4 recvright) at (Traced Task 4 Port 0) [recv, anchor=west] {R};
+
+\node (Unknown Task recvright) at (Unknown Task Port 7) [recv, anchor=east] {R};
+\node (Unknown Task sendright) at (Unknown Task Port -6) [send, anchor=east] {S};
 
@@
 
+\node (S4) at (rpctrace Right Port -16) [send, anchor=east] {S};
+\node (R4) at (rpctrace Left Port -16) [recv, anchor=west] {R};
+
+\draw [port link] (Unknown Task sendright.east) -| ($(Unknown Task sendright.east)!.5!(R4.west)$) |- (R4.west);
+\draw [internal link] (R4) -- (S4);
+
+\draw [port link] (S4) -- (Task 4 recvright);
+
 \draw [internal link] (S1) -- (R2);
@@
 
+\node (rpctrace message) [message] at ($(R4)!.5!(S4)$) {};
+
+\node (rpctrace message recvright) [recv] at (rpctrace message) {R};
+
+\endscope
 \end{tikzpicture}
 \end{frame}
+
+\begin{comment}
 
@@
 
+\end{comment}
+
 \end{document}

### 2019-06-03 doc/ex1.c
add simple programs use to illustrate superscalar instructions in cua slidedeck

@@ -1,0 +1,9 @@
+int a[512];
+int b[512];
+int c[512];
+
+int main () {
+  for (int i=0; i<512; i++){
+    a[i] = b[i] + c[i];
+  }
+}

### 2019-06-03 doc/ex2.c

@@ -1,0 +1,9 @@
+int a[512] __attribute__((aligned(0x20)));
+int b[512] __attribute__((aligned(0x20)));
+int c[512] __attribute__((aligned(0x20)));
+
+int main () {
+  for (int i=0; i<512; i++){
+    a[i] = b[i] + c[i];
+  }
+}

### 2019-06-03 doc/ex3.c

@@ -1,0 +1,9 @@
+int a[512] __attribute__((aligned(64)));
+int b[512] __attribute__((aligned(64)));
+int c[512] __attribute__((aligned(64)));
+
+int main () {
+  for (int i=0; i<512; i++){
+    a[i] = b[i] + c[i];
+  }
+}

### 2019-06-03 doc/ex4.cc

@@ -1,0 +1,35 @@
+#include <thread>
+
+int a[512];
+int b[512];
+int c[512];
+
+int operation(int * ap, int * bp, int * cp, int count)
+{
+  for (int i = 0; i < count; i++) {
+    cp[i] = ap[i] + bp[i];
+  }
+}
+
+const int num_threads = 3;
+
+int main () {
+  int start_index[num_threads+1];
+  std::thread thread[num_threads];
+
+  start_index[0] = 0;
+  for (int i=1; i<num_threads; i++) {
+    start_index[i] = i*(512/num_threads) & ~15;
+  }
+  start_index[num_threads] = 512;
+
+  for (int i=0; i<num_threads; i++) {
+    thread[i] = std::thread(operation,
+			    &a[start_index[i]], &b[start_index[i]], &c[start_index[i]],
+			    start_index[i+1] - start_index[i]);
+  }
+
+  for (int i=0; i<num_threads; i++) {
+    thread[i].join();
+  }
+}
